---
title: "Topic Modeling of Perturb-Responsive Cells: Gene-Set Mediation Analysis for MCH and Asthma"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
date: "2026-01-03"
knitr:
  opts_chunk:
    echo: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

This notebook implements a three-layer gene-set mediation framework that links GWAS peripheral genes to complex traits through de novo co-regulated transcriptional programs discovered by topic modeling of Perturb-seq data. The analysis is applied to two traits:

1. **Mean Corpuscular Hemoglobin (MCH)** — a quantitative hematological trait (GWAS: GCST90002390), using the **Replogle** K562 Perturb-seq dataset. 689 GWAS loci were mapped to effector genes via a 200 kb window and FLAMES (4,475 genes from 200 kb; 283 from FLAMES), yielding 1,297 perturbations. 441 out of 689 (64%) GWAS loci are explained by our model. 

2. **Asthma** — a complex immune-mediated disease (GWAS: GCST010043), using the **Nadig** Perturb-seq dataset. GWAS loci were mapped to effector genes via a 200 kb window and FLAMES (1251 genes from 200 kb; 127 from FLAMES). This yielded 75 asthma-relevant perturbations.

### Analysis framework

The mediation framework proceeds in three stages:

**Stage 1 — Topic modeling (split-sample, first half of cells):**
For each trait, we restrict to perturbation-receiving cells whose guide targets are linked to GWAS loci. Half of these cells are used to fit a topic model with `fastTopics`, and grade-of-membership differential expression (DE) analysis (LFSR ≤ 0.05, standardized effect size ≥ 2.5, log fold change >0.5) identifies genes that define each latent topic (i.e., co-regulated transcriptional programs). Topics with ≥5 genes are retained as candidate gene sets.

**Stage 2 — OPERA: gene → gene-set associations (held-out second half of cells):**
The remaining half of perturbation-receiving cells are passed to OPERA, which estimates perturbation–gene-set association p-values. This split-sample design ensures that the same cells are not used both to define gene sets and to test perturbation–mediator associations, guarding against overfitting.

**Stage 3 — Gene-set → trait associations + AdaFilter mediation test:**
Gene-level GWAS burden-test p-values for the trait are aggregated within each gene set using Fisher's method, capturing the cumulative trait-relevance of each program. The two p-value layers (gene → gene-set from OPERA; gene-set → trait from Fisher aggregation) are then jointly tested for mediation using **AdaFilter**, which adaptively combines evidence while accounting for dependence across the two layers.

### Key design choices

- **Cis-gene exclusion:** For each perturbation, genes that are cis-regulated by the perturbed locus (from `discovery_pairs_cis`) are removed from the corresponding topic modules before computing geneset–trait p-values. This prevents circular evidence where the perturbed gene's own cis-effects inflate the geneset–trait signal.
- **Negative control:** 514 non-targeting perturbations from the Replogle dataset serve as negative controls. Under a 5% FDR threshold, only 0.07% of negative-control pairs are called significant, demonstrating excellent calibration.
- **Comparison with Ota et al.:** Of 16 Ota (peripheral gene - downstream regulated pathway) pairs present in our test set, 11 are identified as significant by OPERA. confirming biological concordance.

## Utility functions and libraries

The setup chunk below loads all required libraries and defines shared helper functions used throughout the analysis:

- `gg_qqplot` / `gg_qqplot_list` / `gg_qqplot_df`: QQ plot utilities for visualizing p-value calibration under the null.
- `combine_p_by_target`: Combines gRNA-level p-values to the gene-target level using Bonferroni correction (k × min p-value).
- `plot_gene_module_network2`: Bipartite igraph visualization connecting peripheral genes (circles) to latent topic modules (squares), with edge width scaled by gene→module association strength.

```{r message=FALSE, warning=FALSE}
library(igraph)
library(ggplot2)
library(dplyr)
library(readr)
library(tibble)
library(tidyr)
library(sceptre)
library(readr)
library(dplyr)
library(Matrix)
library(parallel)
library(BH)
library(Rcpp)
library(peakRAM)
library(heavytailcombtest)
library(gridExtra)
library(eulerr)
library(ggplot2)
library(dplyr)
library(readr)
library(tibble)
library(ACAT)
library(MASS)


gg_qqplot_list <- function(pval_list, ci = 0.95, cap = 6, title = "QQ Plot") {
  # pval_list: named list of numeric p-value vectors

  # 1. 构造数据框：method, p
  pts_df <- map_dfr(names(pval_list), function(name) {
    tibble(
      Method = name,
      p = pval_list[[name]]
    )
  })

  # 2. 清洗和排序
  pts_df <- pts_df %>%
    filter(!is.na(p), p > 0) %>%
    mutate(
      observed = pmin(-log10(p), cap)
    ) %>%
    group_by(Method) %>%
    arrange(p, .by_group = TRUE) %>%
    mutate(
      rank_idx = row_number(),
      n = n(),
      expected = -log10(ppoints(n))[rank_idx]
    ) %>%
    ungroup()

  # 3. 构造 envelope（以最大 n 为准）
  max_n <- max(pts_df$n)
  exp_env <- -log10(ppoints(max_n))
  lower_env <- pmin(-log10(qbeta((1 - ci)/2, seq_len(max_n), rev(seq_len(max_n)))), cap)
  upper_env <- pmin(-log10(qbeta((1 + ci)/2, seq_len(max_n), rev(seq_len(max_n)))), cap)

  env_df <- tibble(
    expected = c(exp_env, rev(exp_env)),
    ymin = c(lower_env, rev(upper_env)),
    ymax = c(upper_env, rev(lower_env))
  )

  # 4. 作图
  ggplot() +
    geom_ribbon(data = env_df, aes(x = expected, ymin = ymin, ymax = ymax),
                fill = "grey80", alpha = 0.3) +
    geom_line(data = pts_df,
              aes(x = expected, y = observed, color = Method, group = Method),
              size = 1) +
    geom_point(data = pts_df,
               aes(x = expected, y = observed, color = Method),
               size = 1.5, alpha = 0.6) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5) +
    labs(
      title = title,
      x = expression(Expected~-log[10](italic(P))),
      y = expression(Observed~-log[10](italic(P))),
      color = "Method"
    ) +
    ylim(0, cap) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", hjust = 0.5))
}


gg_qqplot <- function(ps, ci = 0.95, title = "QQ Plot") {
  n <- length(ps)
  df <- data.frame(
    observed = pmin(-log10(sort(ps)), 6), # Cap observed values at 4
    expected = -log10(ppoints(n)),
    clower = pmin(-log10(qbeta(p = (1 - ci) / 2, shape1 = 1:n, shape2 = n:1)), 6), # Cap clower values at 4
    cupper = pmin(-log10(qbeta(p = (1 + ci) / 2, shape1 = 1:n, shape2 = n:1)), 6)  # Cap cupper values at 4
  )

  log10Pe <- expression(paste("Expected -log"[10], plain(P)))
  log10Po <- expression(paste("Observed -log"[10], plain(P)))

  ggplot(df) +
    geom_ribbon(mapping = aes(x = expected, ymin = clower, ymax = cupper), alpha = 0.1) +
    geom_point(aes(expected, observed), size = 1.5, color = "red") +
    geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
    xlab(log10Pe) + 
    ylab(log10Po) +
    ylim(0, 6) + # Limit y-axis to 0-5
    ggtitle(title) # Add title
}



library(ggplot2)
library(dplyr)
library(tidyr)

gg_qqplot_df <- function(pmat,
                         ci    = 0.95,
                         title = "QQ Plot",
                         cap   = 6) {
  # pmat: numeric data.frame or matrix; rows = methods, cols = p-values
  # ci:   confidence level for the envelope
  # title: plot title
  # cap:  maximum -log10(p) to display
  
  # 1. reshape & filter out NAs, compute expected & observed per method
  pmat[pmat<=0]<-1e-250
    pmat[pmat>=1]<-1-1e-250
  pts_df <- pmat %>% 
    as.data.frame() %>% 
    rownames_to_column("Method") %>% 
    pivot_longer(-Method, names_to="dummy", values_to="p") %>% 
    filter(!is.na(p)) %>%          # drop NAs
    group_by(Method) %>% 
    arrange(p) %>% 
    mutate(
      rank_idx = row_number(),
      n_i      = n(),                                 # how many p’s
      expected = -log10(ppoints(n_i))[rank_idx],      # expected quantile
      observed = pmin(-log10(p), cap)                 # cap at ‘cap’
    ) %>% 
    ungroup()
  
  # 2. build a single envelope for the *largest* group size
  max_n <- pts_df %>% pull(n_i) %>% max()
  exp_env <- -log10(ppoints(max_n))
  lower_env <- pmin(-log10(qbeta((1-ci)/2,  seq_len(max_n), rev(seq_len(max_n)))), cap)
  upper_env <- pmin(-log10(qbeta((1+ci)/2,  seq_len(max_n), rev(seq_len(max_n)))), cap)
  env_df <- tibble(expected = exp_env, ymin = lower_env, ymax = upper_env)
  
  # 3. plot
  ggplot() +
    geom_ribbon(data = env_df,
                aes(x = expected, ymin = ymin, ymax = ymax),
                fill  = "grey80", alpha = 0.3) +
    geom_line(data = pts_df,
              aes(x = expected, y = observed, color = Method, group = Method),
              size = 1) +
    geom_point(data = pts_df,
               aes(x = expected, y = observed, color = Method),
               size = 1.5) +
    geom_abline(intercept = 0, slope = 1,
                linetype = "dashed", alpha = 0.5) +
    labs(title = title,
         x     = expression(Expected~ -log[10](P)),
         y     = expression(Observed~ -log[10](P)),
         color = "Method") +
    ylim(0, cap) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", hjust = 0.5))
}


combine_p_by_target<- function(P, map, non_targeting_label = "non-targeting") {
  stopifnot(!is.null(rownames(P)))
  # 去重并按 P 行对齐
  if (any(duplicated(map$grna_id))) {
    map <- map[!duplicated(map$grna_id), , drop = FALSE]
  }
  miss <- setdiff(rownames(P), map$grna_id)
  if (length(miss)) stop("map 缺少这些 grna_id: ", paste(miss, collapse = ", "))
  map <- map[match(rownames(P), map$grna_id), , drop = FALSE]

  is_nt <- map$grna_target == non_targeting_label

  ## 1) 非靶向：原样保留（行名继续用 grna_id）
  P_nt   <- P[is_nt, , drop = FALSE]               # 不做任何修改

  ## 2) 其它 target：按 target 合并为一行（Bonferroni：k * min）
  P_g    <- P[!is_nt, , drop = FALSE]
  map_g  <- map[!is_nt, , drop = FALSE]
  groups <- split(seq_len(nrow(P_g)), map_g$grna_target)

  res_list <- lapply(names(groups), function(tg) {
    idx <- groups[[tg]]
    sub <- P_g[idx, , drop = FALSE]
    k   <- length(idx)

    any_non_na <- colSums(!is.na(sub)) > 0
    min_p <- rep(NA_real_, ncol(P_g))
    if (any(any_non_na)) {
      min_p[any_non_na] <- apply(sub[, any_non_na, drop = FALSE], 2, min, na.rm = TRUE)
    }
    pmin(1, k * min_p)
  })
  P_comb <- do.call(rbind, res_list)
  rownames(P_comb) <- names(groups)
  colnames(P_comb) <- colnames(P)

  ## 3) 绑定结果：先合并后的 targets，再原样的 non-targeting
  out <- rbind(P_comb, P_nt)
  out
}

## 用法
# P <- as.matrix(max_stat)  # 你的 p 值矩阵
# res <- combine_p_by_target_keep_nt(P, map, non_targeting_label = "non-targeting")
# 结果 res：行包含 每个 target 的一行 + 每条 non-targeting guide 的原样行

layout_with_fr <- function (graph, coords = NULL, dim = 2, niter = 500, start.temp = sqrt(vcount(graph)), 
    grid = c("auto", "grid", "nogrid"), weights = NULL, minx = NULL, 
    maxx = NULL, miny = NULL, maxy = NULL, minz = NULL, maxz = NULL, 
    coolexp = deprecated(), maxdelta = deprecated(), area = deprecated(), 
    repulserad = deprecated(), maxiter = deprecated()) 
{
    ensure_igraph(graph)
    coords[] <- as.numeric(coords)
    dim <- as.numeric(dim)
    if (dim != 2L && dim != 3L) {
        stop("Dimension must be two or three")
    }
    if (!missing(niter) && !missing(maxiter)) {
        stop("Both `niter' and `maxiter' are given, give only one of them")
    }
    if (!missing(maxiter)) 
        niter <- maxiter
    niter <- as.numeric(niter)
    start.temp <- as.numeric(start.temp)
    grid <- igraph.match.arg(grid)
    grid <- switch(grid, grid = 0L, nogrid = 1L, auto = 2L)
    if (is.null(weights) && "weight" %in% edge_attr_names(graph)) {
        weights <- E(graph)$weight
    }
    if (!is.null(weights) && any(!is.na(weights))) {
        weights <- as.numeric(weights)
    }
    else {
        weights <- NULL
    }
    if (!is.null(minx)) 
        minx <- as.numeric(minx)
    if (!is.null(maxx)) 
        maxx <- as.numeric(maxx)
    if (!is.null(miny)) 
        miny <- as.numeric(miny)
    if (!is.null(maxy)) 
        maxy <- as.numeric(maxy)
    if (!is.null(minz)) 
        minz <- as.numeric(minz)
    if (!is.null(maxz)) 
        maxz <- as.numeric(maxz)
    if (lifecycle::is_present(coolexp)) {
        lifecycle::deprecate_warn("0.8.0", "layout_with_fr(coolexp = )")
    }
    if (lifecycle::is_present(maxdelta)) {
        lifecycle::deprecate_warn("0.8.0", "layout_with_fr(maxdelta = )")
    }
    if (lifecycle::is_present(area)) {
        lifecycle::deprecate_warn("0.8.0", "layout_with_fr(area = )")
    }
    if (lifecycle::is_present(repulserad)) {
        lifecycle::deprecate_warn("0.8.0", "layout_with_fr(repulserad = )")
    }
    on.exit(.Call(R_igraph_finalizer))
    if (dim == 2) {
        res <- .Call(R_igraph_layout_fruchterman_reingold, graph, 
            coords, niter, start.temp, weights, minx, maxx, miny, 
            maxy, grid)
    }
    else {
        res <- .Call(R_igraph_layout_fruchterman_reingold_3d, 
            graph, coords, niter, start.temp, weights, minx, 
            maxx, miny, maxy, minz, maxz)
    }
    res
}
plot_gene_module_network2 <- function(df12,
                                      gene_modules_id = NULL,
                                      top_n_per_module = NULL,
                                      top_n_gene = 80,
                                      pcap = 1e-300,
                                      edge_by = c("none","snp_geneset"),
                                      layout_fun = NULL,  
                                      highlight_mod = NULL,
                                      out_pdf = NULL,
                                      width = 12, height = 9,
                                      label_genes = TRUE,
                                      label_modules = TRUE,
                                      size_scale = 0.7,
                                      mod_size_range = c(14, 40),
                                      gene_size_range = c(6, 18),
                                      highlight_genes = NULL
                                      ) {

  edge_by <- match.arg(edge_by)
  if (is.null(layout_fun)) layout_fun <- igraph::layout_with_fr

  d <- df12 %>%
    dplyr::transmute(
      module = as.character(module),
      gene   = as.character(gene_name),
      p_snp  = pmax(as.numeric(snp_geneset), pcap),
      p_mod  = pmax(as.numeric(geneset_trait), pcap)
    ) %>%
    dplyr::filter(!is.na(module), !is.na(gene)) %>%
    dplyr::group_by(module, gene) %>%
    dplyr::summarise(
      p_snp = min(p_snp, na.rm = TRUE),
      p_mod = min(p_mod, na.rm = TRUE),
      .groups = "drop"
    )

  if (!is.null(top_n_per_module)) {
    d <- d %>%
      dplyr::group_by(module) %>%
      dplyr::arrange(p_snp, .by_group = TRUE) %>%
      dplyr::slice_head(n = top_n_per_module) %>%
      dplyr::ungroup()
  }

  if (!is.null(top_n_gene)) {
    top_genes <- d %>%
      dplyr::group_by(gene) %>%
      dplyr::summarise(best_p = min(p_snp, na.rm = TRUE), .groups = "drop") %>%
      dplyr::arrange(best_p) %>%
      dplyr::slice_head(n = top_n_gene) %>%
      dplyr::pull(gene)
    d <- d %>% dplyr::filter(gene %in% top_genes)
  }

  mod_tbl <- d %>%
    dplyr::group_by(module) %>%
    dplyr::summarise(p_mod = min(p_mod, na.rm = TRUE), .groups = "drop") %>%
    dplyr::mutate(
      w_mod = if (!is.null(gene_modules_id)) {
        module_sizes <- sapply(gene_modules_id, length)
        module_sizes[module]
      } else {
        -log10(p_mod)
      }
    )

  # 用连接的 module 数量作为基因节点大小权重
  gene_mod_count_tbl <- d %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(
      p_snp   = min(p_snp, na.rm = TRUE),
      n_mods  = dplyr::n_distinct(module),
      .groups = "drop"
    )

  vertices <- dplyr::bind_rows(
    mod_tbl %>% dplyr::transmute(name = module, type = "module", weight = w_mod),
    gene_mod_count_tbl %>% dplyr::transmute(name = gene, type = "gene", weight = n_mods)
  )

  edges <- d %>% dplyr::transmute(from = module, to = gene, p_snp = p_snp)

  g <- igraph::graph_from_data_frame(edges, directed = FALSE, vertices = vertices)

  igraph::V(g)$shape <- ifelse(igraph::V(g)$type == "module", "square", "circle")
  igraph::V(g)$color <- ifelse(igraph::V(g)$type == "module", "grey30", "lightskyblue")
  igraph::V(g)$label.color <- "black"
  igraph::V(g)$frame.color <- NA

  mod_ids  <- which(igraph::V(g)$type == "module")
  gene_ids <- which(igraph::V(g)$type == "gene")

  # 只 highlight 传入的 highlight_genes，不再自动标红多模块基因
  if (!is.null(highlight_mod)) {
    highlight_mod_ids <- mod_ids[igraph::V(g)$name[mod_ids] %in% highlight_mod]
    igraph::V(g)$color[highlight_mod_ids] <- "green3"
  }

  if (!is.null(highlight_genes)) {
    highlight_ids <- gene_ids[igraph::V(g)$name[gene_ids] %in% highlight_genes]
    igraph::V(g)$color[highlight_ids] <- "red"
  }

  igraph::V(g)$size <- NA_real_
  if (length(mod_ids) > 0) {
    igraph::V(g)$size[mod_ids] <- scales::rescale(igraph::V(g)$weight[mod_ids],
                                                   to = mod_size_range * size_scale)
  }
  if (length(gene_ids) > 0) {
    w <- igraph::V(g)$weight[gene_ids]
    if (length(unique(w)) == 1) {
      # 所有基因连接相同数量的 module，给统一大小
      igraph::V(g)$size[gene_ids] <- mean(gene_size_range) * size_scale
    } else {
      igraph::V(g)$size[gene_ids] <- scales::rescale(w, to = gene_size_range * size_scale)
    }
  }

  if (edge_by == "snp_geneset") {
    igraph::E(g)$width <- scales::rescale(-log10(igraph::E(g)$p_snp), to = c(0.5, 4))
  } else {
    igraph::E(g)$width <- 1
  }
  igraph::E(g)$color <- "grey70"

  igraph::V(g)$label <- NA_character_
  if (label_modules) igraph::V(g)$label[mod_ids] <- igraph::V(g)$name[mod_ids]
  if (label_genes)   igraph::V(g)$label[gene_ids] <- igraph::V(g)$name[gene_ids]
  igraph::V(g)$label.cex <- ifelse(igraph::V(g)$type == "module", 0.9, 0.7)

  if (!is.null(out_pdf)) grDevices::pdf(out_pdf, width = width, height = height)
  plot(g, layout = layout_fun(g), vertex.label.dist = 0, edge.curved = 0)
  if (!is.null(out_pdf)) grDevices::dev.off()

  invisible(g)
}

```

# Construct topic gene sets

Topic modeling was applied with `fastTopics` to the perturb-responsive cells (first half of cells receiving MCH-associated perturbations from the Replogle dataset). The `de` object below contains the result of grade-of-membership differential expression analysis on K=50 topics fitted to cells whose guide targets overlap MCH GWAS effector genes (mapped via FLAMES + 200 kb window).

`get_de_genes()` extracts the "top" genes for each topic using three filters simultaneously:
- **LFSR ≤ 0.05**: high posterior confidence (local false sign rate, analogous to a Bayesian FDR)
- **|z| ≥ 2.5**: large standardized effect size
- **|β| ≥ 0.5**: minimum fold-change threshold (log-scale posterior mean)

Topics with fewer than 5 genes after filtering are discarded (`filtered_list_2`), as they are too small to support reliable downstream aggregation tests.

```{r}
get_de_genes <- function(de, k,
                         lfsr_thr = 0.05,
                         z_thr    = 2.5,
                         fc_thr   = 0.5,
                         gene_names) {

  z_k     <- de$z[, k]
  lfsr_k  <- de$lfsr[, k]
  beta_k  <- de$postmean[, k]   

  # 先去掉 NA
  ok <- !is.na(z_k) & !is.na(lfsr_k) & !is.na(beta_k)

  idx <- which(
    ok &
      lfsr_k <= lfsr_thr &      # 置信度高
      abs(z_k)   >= z_thr &     # z 比较大
      abs(beta_k) >= fc_thr     # fold change 比较大
  )

  gene_names[idx]
}





de <-  readRDS("/project/xuanyao/jiaming/Getting_started/data/CDT4/topic_model/flames+200kb/de_40.rds")



de_genes_list <- lapply(seq_len(ncol(de$z)),
                        function(k) get_de_genes(de, k,gene_names=rownames(de$F)))
names(de_genes_list) <- paste0("k", seq_len(ncol(de$z)))

filtered_list_2 <- Filter(function(x) length(x) > 4, de_genes_list)

length(filtered_list_2)

```


# MCH analysis (Replogle dataset, K = 50 topics)

The primary MCH analysis uses K = 50 topics fitted to Replogle K562 Perturb-seq cells receiving perturbations targeting MCH-associated GWAS genes (1,297 perturbations; loci mapped via FLAMES + 200 kb window). Gene sets were derived with thresholds LFSR ≤ 0.01, |z| ≥ 2, |β| ≥ 0.5. Of the 50 topics, 47 contain more than 5 genes and are retained.

The recovered programs can be grouped into five broad biological categories consistent with erythroid biology:
1. **Hemoglobin synthesis** — iron handling, heme metabolism, red blood cell maturation (e.g., ABCB10, HBA1, HBD, KLF1, SLC25A37, TFRC)
2. **Cell cycle and DNA replication** — G0/G1, S-phase, G2/M checkpoint control (e.g., CCNA2, CCNF, CLSPN, BRIP1, NUP214, POLE, E2F2)
3. **Autophagy and stress adaptation** — quality control and remodeling under high biosynthetic demand (e.g., AP2A1, ATR, EIF2B4, EIF3E/F/M, HSPA9, WDFY3, TRAPPC10)
4. **Translation and proteostasis** — cytoplasmic translation, ribosome biogenesis, ubiquitin-mediated protein catabolism
5. **Core metabolism and redox biology** — amino acid biosynthesis, one-carbon metabolism, purine metabolism, heme pathways, antioxidant/detoxification

## GO enrichment of topic gene sets

Gene sets from each topic are submitted to g:Profiler (`gost`) as a multi-query, testing against GO:BP, GO:MF, and KEGG with FDR correction. The result is reshaped into a long table (`go_long`) and the top terms per topic are extracted for annotation of significant mediating modules downstream.

```{r}
library(gprofiler2)
 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/repologle/topic_geneset/flames+200kb/de_50_0.01_2_0.5.rds")

 de_10_0.01_2_0.5 <-gene_modules_id

gene_sets <- lapply(de_10_0.01_2_0.5, function(x) unique(as.character(x)))

# gene_sets <- lapply(de_10_0.01_2_0.5 , function(x) unique(as.character(x)))

# 建议给每个 set 一个名字（没有就补上）
if (is.null(names(gene_sets))) names(gene_sets) <- paste0("set_", seq_along(gene_sets))

res <- gost(
  query = gene_sets,                 # list of 100 gene vectors
  organism = "hsapiens",
  multi_query = TRUE,
  correction_method = "fdr",
  user_threshold = 0.05,
  sources = c("GO:BP","GO:MF","KEGG")
)



# geneset 名字：如果你有 gene_sets 的 names 就用它；否则自动生成
gs_names <- names(gene_sets)
if (is.null(gs_names)) gs_names <- paste0("set_", seq_along(res$result$significant[[1]]))

# GO term 名字：用 term_id 或 term_name（建议 term_id 更稳定）
go_names <- paste0(res$result$source, ":", res$result$term_id)
# 如果你更想用可读名字：
# go_names <- paste0(res$result$source, ":", res$result$term_name)

# 1) list -> 矩阵：每列一个 term，每行一个 geneset
sig_term_by_set <- do.call(cbind, res$result$significant)
storage.mode(sig_term_by_set) <- "logical"

# 2) 转成“geneset(行) × term(列)”的 index matrix
idx_mat <- sig_term_by_set  # 现在行=geneset, 列=term（通常就是这个方向）
rownames(idx_mat) <- gs_names
colnames(idx_mat) <- go_names


# 如果你发现方向反了（极少数情况），就用这一句修正：
# idx_mat <- t(idx_mat)

library(dplyr)
library(tidyr)

go_tbl <- as_tibble(res$result)

# geneset 名称（如果你有 names(gene_sets) 就用它；否则自动生成）
G <- length(go_tbl$p_values[[1]])
gs_names <- names(gene_sets)
if (is.null(gs_names) || length(gs_names) != G) gs_names <- paste0("set_", seq_len(G))

# 展开：term × geneset 的长表
go_long <- go_tbl %>%
  mutate(
    query   = list(gs_names),
    p_value = p_values,
    sig     = significant       # significant 列也是 list，记录每个 geneset 是否显著
  ) %>%
  dplyr::select(source, term_id, term_name, query, p_value, sig) %>%
  unnest(c(query, p_value, sig)) %>%
  filter(sig == TRUE)  

# 每个 geneset 取 top10（p_value 最小）
top10_tbl <- go_long %>%
  group_by(query) %>%
  arrange(p_value, .by_group = TRUE) %>%
  ungroup()

# 你想要的：每个 geneset 的 top10 term_name（以及可选的 p_value）
top10_names <- top10_tbl %>%
  group_by(query) %>%
  summarise(top10_term_name = list(term_name), .groups = "drop")



# 注意！！！！！！！！！！！！！！！是按 1， 10 ，11 等排序
```

```r
library(openxlsx)
wb <- createWorkbook()
for (q in sort(unique(top10_tbl$query))) {
  df_sheet <- top10_tbl %>%
    filter(query == q) %>%
    dplyr::select(term_name, p_value) %>%
    as.data.frame()
  addWorksheet(wb, q)
  writeData(wb, q, df_sheet)
}
saveWorkbook(wb, "/project/xuanyao/jiaming/paper/GO_terms_pvalue_MCH_k50.xlsx", overwrite = TRUE)
```





## Geneset–trait associations and AdaFilter mediation test (MCH, K50)

This section implements the third stage of the mediation framework: computing gene-set → trait association p-values and running AdaFilter for joint mediation testing.

**Gene-set → trait p-values (Fisher aggregation):**
Gene-level burden-test p-values for MCH (from `MCH_summary.tsv`, model M1.01) are aggregated within each topic gene set using Fisher's method. This captures the cumulative evidence that genes in a given transcriptional program are enriched for MCH associations. Before aggregation, each topic's gene set is filtered to remove genes that are cis-regulated by the target perturbation (from `discovery_pairs_cis`), implementing the cis-gene exclusion described in the Methods. This is the split-sample held-out half of cells — OPERA p-values from the first half were pre-computed and stored in `geneset_p_value.rds`.

**AdaFilter:**
The two p-value columns — `p_snp_geneset` (gene → gene set, from OPERA) and `p_fisher` (gene set → trait, from Fisher aggregation) — are passed jointly to `adaFilter(r=2, alpha=0.05)`. AdaFilter tests the composite null that at least one of the two layers is null, adaptively combining evidence while controlling for dependence. Pairs with `decision = 1` constitute the significant mediation signals.

```{r}
library(data.table)
library(dplyr)
library(stringr)
library(purrr)
library(heavytailcombtest)

# -----------------------------
# 0) Read inputs
# -----------------------------
df <- fread("/project/xuanyao/jiaming/Getting_started/data/geneset_to_trait/MCH/MCH_summary.tsv")
df2 <- df[grepl("M1.01", df$Name), ]

Nadig_associated_perturbation <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/MCH/flames+200kb_mapping/Repologle_associated_perturbations.rds")



discovery_pairs_cis <- readRDS("/project/xuanyao/jiaming/Getting_started/data/repologle/discovery_pairs_cis.rds")


 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/repologle/topic_geneset/flames+200kb/de_50_0.01_2_0.5.rds")



# -----------------------------
# 1) Build pv_map: gene-level MCH burden-test p-values
#    Source: MCH_summary.tsv (GCST90002390), model M1.01
#    Gene ID is extracted from the "Name" column (format: "...(ENSG...)")
#    If multiple rows correspond to the same gene, take the minimum p-value
# -----------------------------
gene_id <- str_extract(df2$Name, "(?<=\\()[^()]+(?=\\))")

lookup <- df2 %>%
  mutate(ID = gene_id) %>%
  filter(!is.na(ID), !is.na(p_value)) %>%
  group_by(ID) %>%
  summarise(p_value = min(p_value), .groups = "drop")

pv_map <- setNames(lookup$p_value, lookup$ID)


# -----------------------------
# 2) Helper functions for p-value combination
#    clean_p:         sanitize p-values (set non-positive/NA to NA)
#    safe_comb:       Cauchy combination test (ACAT) — robust to correlation
#    safe_comb_fisher: Fisher's method — chi-squared with 2k df; used as the
#                     primary geneset-trait p-value because it weights all genes
#                     equally and is interpretable for downstream aggregation
# -----------------------------
clean_p <- function(p) {
  p <- as.numeric(p)
  p[is.na(p) | p <= 0] <- NA
  p
}

safe_comb <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  combination.test(
    p,
    weights = NA,
    method = "Cauchy",
    tail.idx = 1
  )
}

safe_comb_fisher <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  stat <- -2 * sum(log(p))
  stats::pchisq(stat, df = 2 * length(p), lower.tail = FALSE)
}


# -----------------------------
# 3) Cis-gene exclusion (split-sample design)
#    For each perturbation, identify cis-regulated response genes from
#    discovery_pairs_cis and remove them from all topic modules.
#    This ensures that the geneset-trait signal is not inflated by the
#    perturbed gene's own cis-effects (the same reason we use the held-out
#    half of cells for OPERA).
#    n_removed records how many cis genes were excluded per perturbation.
# -----------------------------
make_filtered_modules <- function(gene_symbol, gene_modules_id, pairs_df) {
  target_name <- gsub("^gRNA:", "", gene_symbol)


  ids_to_remove <- unique(pairs_df$response_id[pairs_df$grna_target == target_name])

  filtered_modules <- lapply(gene_modules_id, function(mod_ids) {
    setdiff(mod_ids, ids_to_remove)
  })

  list(
    modules   = filtered_modules,
    n_removed = length(ids_to_remove)
  )
}

filtered_by_perturbation <- setNames(
  lapply(
    Nadig_associated_perturbation,
    make_filtered_modules,
    gene_modules_id = gene_modules_id,
    pairs_df = discovery_pairs_cis
  ),
  Nadig_associated_perturbation
)


# -----------------------------
# 4) Compute module-level combined p-values per perturbation
#    For each perturbation × topic module pair:
#      - Collect gene-level MCH p-values for genes in the (filtered) module
#      - Combine with ACAT (p_acat) and Fisher (p_fisher)
#    p_fisher is used as the geneset-trait layer in AdaFilter
# -----------------------------
compute_for_one_pert <- function(pert) {
  obj <- filtered_by_perturbation[[pert]]
  gene_modules_id_pert <- obj$modules
  n_removed <- obj$n_removed

  # Collect gene-level MCH p-values for each module (after cis exclusion)
  module_pvals <- lapply(gene_modules_id_pert, function(genes) {
    p <- pv_map[genes]
    names(p) <- genes
    p[!is.na(p)]
  })

  # Aggregate within each module
  p_acat <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb(p))
  })

  p_fisher <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb_fisher(p))
  })

  tibble(
    snp       = pert,
    n_removed = n_removed,  # number of cis-regulated genes removed from modules
    module    = names(module_pvals),
    n_genes   = lengths(gene_modules_id_pert),
    n_matched = sapply(module_pvals, function(v) sum(!is.na(v))),
    p_acat    = p_acat,
    p_fisher  = p_fisher
  )
}

module_results <- map_dfr(Nadig_associated_perturbation, compute_for_one_pert)

# Inspect top module-level geneset-trait associations (sorted by Fisher p-value)


```

**Below are the significant peripheral gene-module pairs**

```{r}
library(tidyr)
library(dplyr)
library(tibble)

# Load pre-computed OPERA p-values (gene → gene-set layer, held-out second half of cells)
# Rows = perturbations (gRNA IDs); columns = topic modules (k1..k50)
# Separate targeting perturbations from non-targeting controls
snp_geneset_p_value<-  readRDS("/project/xuanyao/jiaming/paper/output/snake/repologle_MCH/flames+200kb/K50/3000_t_ACAT_resample/p_values/geneset_p_value.rds")
snp_geneset_p_value_test <- snp_geneset_p_value[which(!grepl("non-",rownames(snp_geneset_p_value))),]
snp_geneset_p_value_control <- snp_geneset_p_value[which(grepl("non-",rownames(snp_geneset_p_value))),]

# Layer 1 (gene → gene-set): pivot OPERA p-value matrix to long format
# Each row becomes one (perturbation, module) pair
df1_long <- snp_geneset_p_value_test %>%
  as.data.frame() %>%
  rownames_to_column("snp") %>%
  pivot_longer(
    -snp,
    names_to  = "module",
    values_to = "p_snp_geneset"
  )

# Layer 2 (gene-set → trait): Fisher aggregated p-values computed above
df2 <- module_results %>%
  dplyr::select(module, p_fisher,snp)

# Join both layers on (snp, module); drop pairs where either layer is NA
merged <- df1_long %>%
  left_join(df2, by = c("snp", "module")) %>%
  relocate(snp, module, p_snp_geneset, p_fisher)%>% tidyr::drop_na()

# AdaFilter joint mediation test (r=2: both layers must contribute signal)
# alpha = 0.05 nominal threshold; decision=1 indicates a significant mediating pair
X <- merged[,c(3,4)]

library(adaFilter)

result <- adaFilter(X, r = 2,alpha=0.05)

merged$decision <- result$decision
colnames(merged)[c(3,4)] <- c("snp_geneset","geneset_trait")

head(merged[merged$decision==1,])


dfl2 <- merged[merged$decision==1,]
dfl2$gene_name <- dfl2$snp
```

The chunk below maps significant peripheral gene IDs (ENSG format from OPERA row names) back to gene symbols, and clusters them by genomic locus (200 kb window) so that co-located genes at the same GWAS locus are grouped together in downstream visualizations.

```{r}
# 0. ENSG -> symbol mapping from GTF
gtf <- rtracklayer::import("/project/xuanyao/jiaming/Getting_started/data/protein_coding_genes/Homo_sapiens.GRCh38.112.gtf.gz",
                            feature.type = "gene")
gtf_df <- as.data.frame(gtf)[, c("gene_id", "gene_name")]
colnames(gtf_df) <- c("ensg", "symbol")
ensg_to_symbol <- setNames(gtf_df$symbol, gtf_df$ensg)

# 1. Build rsid -> cluster_id: merge GWAS loci within 200kb on the same chromosome
gwas_result <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/MCH/flames+200kb_mapping/result.rds")

rsid_locations <- gwas_result %>%
  dplyr::select(rsid, locations) %>%
  distinct() %>%
  filter(!is.na(locations)) %>%
  mutate(
    chr = sub(":.*", "", locations),
    # suppressWarnings: non-autosomal (X/Y/MT) or malformed entries produce NA,
    # which are removed by the chr filter below — this is expected
    pos = suppressWarnings(as.numeric(sub(".*:", "", locations)))
  ) %>%
  filter(chr %in% as.character(1:22)) %>%
  arrange(chr, pos) %>%
  group_by(chr) %>%
  mutate(cluster_num = cumsum(c(1, diff(pos) > 200000))) %>%
  ungroup() %>%
  mutate(cluster_id = paste0(chr, "_", cluster_num)) %>%
  dplyr::select(rsid, chr, pos, cluster_id)

# 2. symbol -> rsid mapping from gwas_result (bridge: ENSG -> symbol -> rsid)
symbol_to_rsid <- gwas_result %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol, rsid) %>%
  distinct()

# 3. Parse dfl2$snp ("gRNA:ENSG...") -> ENSG -> symbol -> rsid -> cluster_id
tbl <- dfl2 %>%
  mutate(ensg  = sub("^gRNA:", "", snp),
         symbol = ensg_to_symbol[ensg]) %>%
  # many-to-many is expected: one gene can be near multiple GWAS loci (multiple rsids),
  # and the same gene can appear in multiple modules. Rows are collapsed in step 4
  # by group_by(module, cluster_id).
  left_join(symbol_to_rsid, by = "symbol", relationship = "many-to-many") %>%
  left_join(rsid_locations %>% dplyr::select(rsid, cluster_id), by = "rsid") %>%
  mutate(cluster_id = ifelse(is.na(cluster_id), paste0("solo_", ensg), cluster_id))

# 4. Within each module, find connected components of a gene–cluster bipartite graph.
#    Two genes (or two clusters) end up in the same component if they are bridged by a
#    shared gene or a shared cluster anywhere in the chain — enabling transitive merging.
#    e.g. gene1↔cluster_A, gene1↔cluster_B  →  gene1 + all genes in cluster_B form one row.
library(igraph)

merged_df <- tbl %>%
  group_by(module) %>%
  group_modify(function(df, key) {
    # Build bipartite edges: gene node  <->  cluster node
    # Use ENSG as fallback node label when symbol is missing
    edges <- df %>%
      transmute(
        gene_node    = paste0("g:", ifelse(!is.na(symbol), symbol, ensg)),
        cluster_node = paste0("c:", cluster_id)
      ) %>%
      distinct()

    g    <- igraph::graph_from_data_frame(edges, directed = FALSE)
    comp <- igraph::components(g)$membership          # named vector: node -> comp_id

    # Map each row back to its component via the cluster node
    node_comp <- tibble(node = names(comp), comp_id = as.integer(comp))

    df %>%
      left_join(
        node_comp %>%
          filter(startsWith(node, "c:")) %>%
          transmute(cluster_id = sub("^c:", "", node), comp_id),
        by = "cluster_id"
      ) %>%
      group_by(comp_id) %>%
      summarise(
        snp_geneset      = min(snp_geneset,   na.rm = TRUE),
        geneset_trait    = min(geneset_trait, na.rm = TRUE),
        merged_gene_name = paste(sort(unique(symbol[!is.na(symbol)])), collapse = "-"),
        merged_loci      = paste(sort(unique(rsid[!is.na(rsid)])),     collapse = "-"),
        .groups = "drop"
      ) %>%
      dplyr::select(-comp_id)
  }) %>%
  ungroup() %>%
  mutate(
    summary = paste0(merged_gene_name, "(", merged_loci, ")")
  ) %>%
  dplyr::select(module, merged_gene_name, merged_loci, summary)

head(merged_df)
```

### Go term of 21 significant genesets




The 21 significant mediating gene sets identified by AdaFilter (at nominal α = 0.05) are listed below. Functional annotation via g:Profiler (`sig_GO`) reveals five broad categories:

1. **Hemoglobin and erythroid programs** (k2, k4): iron handling, heme metabolism, red blood cell maturation — most directly relevant to MCH (average hemoglobin content per red blood cell).
2. **Cell-cycle and DNA-replication programs** (k5, k8, k16, k19, k33, k39): G0/G1, S-phase, and G2/M modules — reflecting the tightly regulated cell divisions required before terminal erythroid maturation.
3. **Autophagy and stress-adaptation programs** (k7, k44): quality control, remodeling, and stress-response processes supporting high biosynthetic demand.
4. **Translation and proteostasis** (multiple modules): cytoplasmic translation, ribosome biogenesis, ubiquitin-dependent protein catabolism.
5. **Core metabolism and redox biology** (multiple modules): amino acid biosynthesis, one-carbon metabolism, purine metabolism, oxygen transport, antioxidant activity.

```{r}
significant_modules <- c("k6","k8",  "k2" , "k26" ,"k3" , "k30" ,"k32" ,"k33" ,"k38" ,"k39", "k4",  "k45", "k46", "k49" ,"k7",
"k5",  "k16" ,"k44" ,"k19","k25","k28")

sig_GO <- readRDS("/project/xuanyao/jiaming/paper/output/mediation_analysis/MCH/flames+200kb/k50/sig_GO.rds")

# sig_GO<-top10_names %>%
#   filter(query %in% significant_modules) %>%
#   mutate(top10_term_name = sapply(top10_term_name, paste, collapse = "; "))
# 
# sig_GO$summary <- sig_GO$query
# sig_GO$summary[1] <- "cell cycle"
# sig_GO$summary[2] <- "cell cycle"
# sig_GO$summary[3] <- "hemoglobin"
# sig_GO$summary[4] <- "amino acid metabolism / amino acid transport / tRNA charging / integrated stress response"
# sig_GO$summary[5] <- "ribosome biogenesis / rRNA processing / translation / biosynthetic activity autophagy"
# 
# sig_GO$summary[6] <- "translation / protein trafficking / RNA localization / organelle organization / ER stress adaptation"
# 
# sig_GO$summary[7] <- "protein binding"
# 
# sig_GO$summary[8] <- "cell cycle"
# sig_GO$summary[9] <- "ribosome biogenesis / rRNA processing / RNA metabolism / nucleotide biosynthesis"
# sig_GO$summary[10] <- "cell cycle"
# sig_GO$summary[11] <- "hemoglobin"
# sig_GO$summary[12] <- "autophagy"
# sig_GO$summary[13] <- "mitochondrial respiration / oxidative phosphorylation / electron transport chain program"
# sig_GO$summary[14] <- "pyrimidine biosynthesis / nucleotide metabolism + RNA processing/decay"
# sig_GO$summary[15] <- "T-cell / immune activation + translation/ribosome activity + guanine nucleotide biosynthesis"
# sig_GO$summary[16] <- "cell cycle"
# sig_GO$summary[17] <- "OXPHOS / mitochondrial respiration + proteostasis / ubiquitin-proteasome + translation/ribosome biogenesis"
# sig_GO$summary[18] <- "autophagy"
# sig_GO$summary[19] <- "cell cycle"


```

```{r}
sig_GO[,c(1,3,4)]
```




## Comparison with Ota et al.

To assess concordance with prior work, we compare our significant gene-set members to the three major biological axes reported by Ota et al. for MCH: hemoglobin synthesis, cell cycle, and autophagy.

The chunks below define the gene lists for each Ota et al. axis and check overlap with:
Peripheral genes in our significant mediating modules — do the genes have the same regulatory effect as Ota?

Of 16 Ota (peripheral gene - downstream regulated pathway) pair present in our test set, 11 (68.75%) are identified as significant by OPERA. At the gene-set level, cell-cycle module genes include G2E3, CCNF, CCNA2, H4C3, CLSPN, ATAD2, and BRIP1; hemoglobin module genes include HBA1, HBD, KLF1, SLC25A37, and EPOR.


**Below are the peripheral genes from Ota, catogorized by pathways they regulate**

```{r,echo=T}
haemoglobin_genes <-c(
  "KLF1", "NPC1", "SUPT5H", "CCNF", "CAD", "CALR", "CBFA2T3",
  "CCNA2", "CLSPN", "DNAJC24", "NDUFB7", "NUP214", "POLE",
  "ABCB10", "(DNAJC13)", "MED13L", "MED23", "(MTOR)", "(NCOR1)",
  "(RPTOR)", "(SALL4)"
)
cell_cycle_genes <- c(
  "HSPA9", "MED14", "MED17", "SUPT5H", "SYMPK"
)

autophagy_genes <- c(
  "AP2A1", "ATR", "EIF2B4", "EIF3E", "EIF3F", "EIF3M", "HSPA9",
  "MED13L", "MED14", "MED17", "MED26", "MTOR", "NIPBL", "POLR2G",
  "RNF20", "RPTOR", "RTTN", "SPRED2", "SUPT5H", "SYMPK", "TP73",
  "TRAPPC10", "WDFY3", "CLSPN", "(EEF2)", "POLE", "BPGM", "UBE2D3",
  "(USP10)", "(XPOT)"
)

```

```{r}
gene_names_clean <- sub("\\s*\\(.*\\)$", "", merged_df$merged_gene_name)

gene_vector <- unique(unlist(
  strsplit(gene_names_clean, "-(?=[A-Za-z])", perl = TRUE)
))


Repologle_associated_perturbations <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/MCH/flames+200kb_mapping/Repologle_associated_perturbations.rds")

# ensg_to_symbol already defined above from GTF

# 2. 去掉 "gRNA:" 前缀，提取 ENSG ID
ensg_ids <- sub("^gRNA:", "", Repologle_associated_perturbations)

# 3. 映射成 symbol
symbol_vector <- ensg_to_symbol[ensg_ids]
```

**We next check how many of them can be tested (i.e. in our peripheral gene list), and how many of them pass the significance level**

```{r}
for (cat_name in c("Haemoglobin", "Cell_cycle", "Autophagy")) {
  genes <- switch(cat_name,
    Haemoglobin = haemoglobin_genes,
    Cell_cycle  = cell_cycle_genes,
    Autophagy   = autophagy_genes
  )
  
  in_test <- intersect(genes, symbol_vector)
  in_sig  <- intersect(genes, gene_vector)
  
  cat("═══════════════════════════════════════\n")
  cat("Category:", cat_name, "\n")
  cat("Total genes in Ota:", length(genes), "\n\n")
  
  cat("  In our test set —", length(in_test), "genes:\n")
  cat(" ", paste(sort(in_test), collapse = ", "), "\n\n")
  
  cat("  Pass our test —", length(in_sig), "genes:\n")
  cat(" ", paste(sort(in_sig), collapse = ", "), "\n\n")
}
```



```{r}
# Pool genes from topic modules assigned to each biological category
# (significant modules per category determined from sig_GO annotations above)
cell_cycle_module_genes<- c(gene_modules_id$k16,gene_modules_id$k19,gene_modules_id$k33,gene_modules_id$k39,gene_modules_id$k5,gene_modules_id$k8)

autophagy_module_genes <- c(gene_modules_id$k44,gene_modules_id$k7)

haemoglobin_module_genes <- c(gene_modules_id$k2,gene_modules_id$k4)
```

```{r,include=F}
to_symbol <- function(ensg_vec) {
  ensg_ids <- sub("^gRNA:", "", ensg_vec)
  symbols <- ensg_to_symbol[ensg_ids]
  # Keep ENSG ID for any genes without a symbol mapping
  ifelse(is.na(symbols) | symbols == "NA", ensg_ids, symbols)
}

# Convert ENSG IDs (with "gRNA:" prefix) in module gene lists to gene symbols
cell_cycle_module_symbols <- to_symbol(cell_cycle_module_genes)
autophagy_module_symbols <- to_symbol(autophagy_module_genes)
haemoglobin_module_symbols <- to_symbol(haemoglobin_module_genes)

hemoglobin_module_genes_ota<- c(
  "ABCB10", "CSTB", "HBA1", "HBD", "KLF1", "RTTN", "SLC25A37",
  "TFRC", "(EPOR)", "(GYPA)"
)

cell_cycle_module_genes_ota<- c(
  "CCNA2", "CCNF", "(CEP350)", "(G2E3)", "(H4C3)",
  "CCNA2", "CCNF", "(DDX39A)", "(G2E3)",
  "CCNA2", "CLSPN", "EIF2", "(ATAD2)", "(H4C3)",
  "BRIP1", "CCNA2", "CLSPN", "EIF2", "NUP214", "POLE", "(ATAD2)"
)



hemoglobin_module_genes_ota<- c(
  "ABCB10", "CSTB", "HBA1", "HBD", "KLF1", "RTTN", "SLC25A37",
  "TFRC", "EPOR", "GYPA"
)


cell_cycle_module_genes_ota<- c(
  "CCNA2", "CCNF", "CEP350", "G2E3", "H4C3",
  "CCNA2", "CCNF", "DDX39A", "G2E3",
  "CCNA2", "CLSPN", "EIF2", "ATAD2", "H4C3",
  "BRIP1", "CCNA2", "CLSPN", "EIF2", "NUP214", "POLE", "ATAD2"
)


intersect(cell_cycle_module_symbols,cell_cycle_module_genes_ota)
intersect(hemoglobin_module_genes_ota,haemoglobin_module_symbols)
```

**Then we consider whether the significant peripheral genes have the same regulating effect as indicated by Ota. Below comparing the functional regulation of shared peripheral genes between OPERA and Ota**

For example, CAD, in our model, are significantly associated with Autophagy, Cell Cycle, and Haemoglobin pathway, while in Ota are only associated with Haemoglobin. Then we treat CAD-Haemoglobin as a successfully replicated pair. We have replicated 11 out of 16 pairs from Ota.

```{r}
# ── Step 1: 获取有分类的模块列表 ──────────────────────────────────────────────
classified_modules <- sig_GO %>%
  filter(grepl("hemoglobin|haemoglobin|cell cycle|autophagy", summary, ignore.case = TRUE)) %>%
  distinct(module) %>%
  pull(module)

# ── Step 2: 处理 df12，拆分连字符基因名 ───────────────────────────────────────
gene_module_df <- merged_df %>%
  transmute(
    module   = as.character(module),
    gene_raw = as.character(merged_gene_name)
  ) %>%
  filter(!is.na(module), !is.na(gene_raw)) %>%
  filter(module %in% classified_modules) %>%
  distinct(module, gene_raw) %>%
  mutate(gene = strsplit(gene_raw, "-")) %>%
  tidyr::unnest(gene) %>%
  mutate(gene = trimws(sub("\\s*\\(.*", "", gene))) %>%
  filter(gene != "") %>%
  filter(gene %in% c(haemoglobin_genes, cell_cycle_genes, autophagy_genes)) %>%
  distinct(gene, module)

# ── Step 3: 获取每个模块的功能分类 ───────────────────────────────────────────
get_module_function <- function(mod_name, sig_GO) {
  rows <- sig_GO[sig_GO$module == mod_name, , drop = FALSE]
  if (nrow(rows) == 0) return(NA_character_)
  summ <- paste(rows$summary, collapse = " ")
  cats <- character(0)
  if (grepl("hemoglobin|haemoglobin", summ, ignore.case = TRUE)) cats <- c(cats, "Haemoglobin")
  if (grepl("cell cycle",             summ, ignore.case = TRUE)) cats <- c(cats, "Cell cycle")
  if (grepl("autophagy",              summ, ignore.case = TRUE)) cats <- c(cats, "Autophagy")
  paste(cats, collapse = " + ")
}

# ── Step 4: 获取每个基因本身的功能分类 ───────────────────────────────────────
get_gene_function <- function(gene_name) {
  if (length(gene_name) == 0 || is.na(gene_name)) return("Unclassified")
  cats <- character(0)
  if (gene_name %in% haemoglobin_genes) cats <- c(cats, "Haemoglobin")
  if (gene_name %in% cell_cycle_genes)  cats <- c(cats, "Cell cycle")
  if (gene_name %in% autophagy_genes)   cats <- c(cats, "Autophagy")
  if (length(cats) == 0) return("Unclassified")
  paste(cats, collapse = " | ")
}

# ── Step 5: 给每个 gene-module pair 加上模块功能标签 ─────────────────────────
gene_module_func <- gene_module_df %>%
  rowwise() %>%
  mutate(func = get_module_function(module, sig_GO)) %>%
  ungroup()

# ── Step 6: 汇总 ─────────────────────────────────────────────────────────────
result_with_func <- gene_module_func %>%
  group_by(gene) %>%
  summarise(
    n            = n(),
    modules      = paste(sort(module),       collapse = ", "),
    functions    = paste(func[order(module)], collapse = ", "),
    func_summary = paste(sort(unique(func)), collapse = " | ")
  ) %>%
  arrange(desc(n), gene) %>%
  mutate(gene_category = sapply(gene, get_gene_function))

# ── Step 7: 输出 ─────────────────────────────────────────────────────────────


colnames(result_with_func)[5:6]<-c("OPERA_downstream_function","Ota_downstream_function")
print(result_with_func, n = Inf)
```




## Gene–module network visualization (MCH)



The MCH-specific network function `plot_gene_module_network_MCH` extends the generic `plot_gene_module_network2` with three-category biological color coding that matches the Ota et al. axes:
- **Hemoglobin** (red #E74C3C): genes/modules involved in heme metabolism and red blood cell function
- **Cell cycle** (yellow #F1C40F): genes/modules involved in G1/S/G2/M progression
- **Autophagy** (blue #3498DB): genes/modules involved in stress adaptation and quality control
- **Mixed**: nodes belonging to multiple categories receive a blended color
- Edges connecting a gene and module that share a biological category are rendered in the matching color and bolded; non-concordant edges are grey

```{r}
plot_gene_module_network_MCH <- function(df12,
                                      gene_modules_id = NULL,
                                      top_n_per_module = NULL,
                                      top_n_gene = 600,
                                      pcap = 1e-300,
                                      edge_by = c("none","snp_geneset"),
                                      layout_fun = NULL,
                                      highlight_mod = NULL,
                                      out_pdf = NULL,
                                      width = 12, height = 9,
                                      label_genes = TRUE,
                                      label_modules = TRUE,
                                      size_scale = 0.7,
                                      mod_size_range  = c(15, 40),
                                      gene_size_range = c(4,  20),
                                      highlight_genes = NULL,
                                      sig_GO = NULL,
                                      haemoglobin_genes = NULL,
                                      cell_cycle_genes  = NULL,
                                      autophagy_genes   = NULL,
                                      unclassified_alpha = 0.3,
                                      label_unclassified_genes = FALSE,
                                      gene_label_cex = 0.7
                                      ) {

  edge_by <- match.arg(edge_by)
  if (is.null(layout_fun)) layout_fun <- igraph::layout_with_fr

  # ── 颜色辅助 ─────────────────────────────────────────────────────────────────
  mix_hex_colors <- function(hex_vec) {
    if (length(hex_vec) == 1) return(hex_vec)
    rgb_mat <- grDevices::col2rgb(hex_vec)
    avg <- rowMeans(rgb_mat)
    grDevices::rgb(avg[1], avg[2], avg[3], maxColorValue = 255)
  }

  add_alpha <- function(col, alpha) adjustcolor(col, alpha.f = alpha)

  col_haem  <- "#E74C3C"
  col_cycle <- "#F1C40F"
  col_auto  <- "#3498DB"

  default_gene_col   <- "grey60"
  default_module_col <- "grey10"

  # ── 基因颜色 ──────────────────────────────────────────────────────────────────
  get_gene_color <- function(raw_name) {
    no_snp     <- trimws(sub("\\s*\\(.*", "", raw_name))
    pure_names <- unlist(strsplit(no_snp, "-"))
    cats <- character(0)
    if (!is.null(haemoglobin_genes) && any(pure_names %in% haemoglobin_genes)) cats <- c(cats, col_haem)
    if (!is.null(cell_cycle_genes)  && any(pure_names %in% cell_cycle_genes))  cats <- c(cats, col_cycle)
    if (!is.null(autophagy_genes)   && any(pure_names %in% autophagy_genes))   cats <- c(cats, col_auto)
    if (length(cats) == 0) return(add_alpha(default_gene_col, unclassified_alpha))
    add_alpha(mix_hex_colors(cats), 1.0)
  }

  # ── 模块颜色 ──────────────────────────────────────────────────────────────────
  get_module_color <- function(mod_name) {
    if (is.null(sig_GO)) return(add_alpha(default_module_col, unclassified_alpha))
    rows <- sig_GO[sig_GO$module == mod_name, , drop = FALSE]
    if (nrow(rows) == 0) return(add_alpha(default_module_col, unclassified_alpha))
    summ <- paste(rows$summary, collapse = " ")
    cats <- character(0)
    if (grepl("hemoglobin|haemoglobin", summ, ignore.case = TRUE)) cats <- c(cats, col_haem)
    if (grepl("cell cycle",             summ, ignore.case = TRUE)) cats <- c(cats, col_cycle)
    if (grepl("autophagy",              summ, ignore.case = TRUE)) cats <- c(cats, col_auto)
    if (length(cats) == 0) return(add_alpha(default_module_col, unclassified_alpha))
    mix_hex_colors(cats)
  }

  # ── 类别辅助（用于边着色）────────────────────────────────────────────────────
  get_gene_categories <- function(raw_name) {
    no_snp     <- trimws(sub("\\s*\\(.*", "", raw_name))
    pure_names <- unlist(strsplit(no_snp, "-"))
    cats <- character(0)
    if (!is.null(haemoglobin_genes) && any(pure_names %in% haemoglobin_genes)) cats <- c(cats, "haem")
    if (!is.null(cell_cycle_genes)  && any(pure_names %in% cell_cycle_genes))  cats <- c(cats, "cycle")
    if (!is.null(autophagy_genes)   && any(pure_names %in% autophagy_genes))   cats <- c(cats, "auto")
    cats
  }

  get_module_categories <- function(mod_name) {
    if (is.null(sig_GO)) return(character(0))
    rows <- sig_GO[sig_GO$module == mod_name, , drop = FALSE]
    if (nrow(rows) == 0) return(character(0))
    summ <- paste(rows$summary, collapse = " ")
    cats <- character(0)
    if (grepl("hemoglobin|haemoglobin", summ, ignore.case = TRUE)) cats <- c(cats, "haem")
    if (grepl("cell cycle",             summ, ignore.case = TRUE)) cats <- c(cats, "cycle")
    if (grepl("autophagy",              summ, ignore.case = TRUE)) cats <- c(cats, "auto")
    cats
  }

  cat_to_col <- c(haem = col_haem, cycle = col_cycle, auto = col_auto)

  # ── 数据处理 ─────────────────────────────────────────────────────────────────
  d <- df12 %>%
    dplyr::transmute(
      module = as.character(module),
      gene   = as.character(merged_gene_name),
      p_snp  = pcap,
      p_mod  = pcap
    ) %>%
    dplyr::filter(!is.na(module), !is.na(gene)) %>%
    dplyr::group_by(module, gene) %>%
    dplyr::summarise(
      p_snp = min(p_snp, na.rm = TRUE),
      p_mod = min(p_mod, na.rm = TRUE),
      .groups = "drop"
    )

  if (!is.null(top_n_per_module)) {
    d <- d %>%
      dplyr::group_by(module) %>%
      dplyr::arrange(p_snp, .by_group = TRUE) %>%
      dplyr::slice_head(n = top_n_per_module) %>%
      dplyr::ungroup()
  }

  if (!is.null(top_n_gene)) {
    top_genes <- d %>%
      dplyr::group_by(gene) %>%
      dplyr::summarise(best_p = min(p_snp, na.rm = TRUE), .groups = "drop") %>%
      dplyr::arrange(best_p) %>%
      dplyr::slice_head(n = top_n_gene) %>%
      dplyr::pull(gene)
    d <- d %>% dplyr::filter(gene %in% top_genes)
  }

  # ── 权重 ─────────────────────────────────────────────────────────────────────
  mod_tbl <- d %>%
    dplyr::group_by(module) %>%
    dplyr::summarise(
      p_mod   = min(p_mod, na.rm = TRUE),
      n_assoc = dplyr::n_distinct(gene),
      .groups = "drop"
    )

  gene_tbl <- d %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(
      p_snp   = min(p_snp, na.rm = TRUE),
      n_assoc = dplyr::n_distinct(module),
      .groups = "drop"
    )

  vertices <- dplyr::bind_rows(
    mod_tbl  %>% dplyr::transmute(name = module, type = "module", weight = n_assoc),
    gene_tbl %>% dplyr::transmute(name = gene,   type = "gene",   weight = n_assoc)
  )

  edges <- d %>% dplyr::transmute(from = module, to = gene, p_snp = p_snp)

  g <- igraph::graph_from_data_frame(edges, directed = FALSE, vertices = vertices)

  # ── 全量赋值基础属性 ──────────────────────────────────────────────────────────
  igraph::V(g)$shape       <- ifelse(igraph::V(g)$type == "module", "square", "circle")
  igraph::V(g)$label.color <- "white"
  igraph::V(g)$frame.color <- NA
  igraph::V(g)$color       <- ifelse(igraph::V(g)$type == "module",
                                     add_alpha(default_module_col, unclassified_alpha),
                                     add_alpha(default_gene_col,   unclassified_alpha))

  mod_ids  <- which(igraph::V(g)$type == "module")
  gene_ids <- which(igraph::V(g)$type == "gene")

  # ── 基因着色 ──────────────────────────────────────────────────────────────────
  for (idx in gene_ids) {
    igraph::V(g)$color[idx] <- get_gene_color(igraph::V(g)$name[idx])
  }

  # ── 模块着色 ──────────────────────────────────────────────────────────────────
  for (idx in mod_ids) {
    igraph::V(g)$color[idx] <- get_module_color(igraph::V(g)$name[idx])
  }

  # ── highlight 覆盖 ───────────────────────────────────────────────────────────
  if (!is.null(highlight_mod)) {
    for (idx in mod_ids[igraph::V(g)$name[mod_ids] %in% highlight_mod])
      igraph::V(g)$color[idx] <- "green3"
  }
  if (!is.null(highlight_genes)) {
    for (idx in gene_ids[igraph::V(g)$name[gene_ids] %in% highlight_genes])
      igraph::V(g)$color[idx] <- "red"
  }

  # ── 节点大小 ─────────────────────────────────────────────────────────────────
  igraph::V(g)$size <- rep(NA_real_, igraph::vcount(g))

  mod_w <- igraph::V(g)$weight[mod_ids]
  igraph::V(g)$size[mod_ids] <- if (length(unique(mod_w)) == 1) {
    rep(mean(mod_size_range) * size_scale, length(mod_ids))
  } else {
    scales::rescale(mod_w, to = mod_size_range * size_scale)
  }

  gene_w <- igraph::V(g)$weight[gene_ids]
  igraph::V(g)$size[gene_ids] <- if (length(unique(gene_w)) == 1) {
    rep(mean(gene_size_range) * size_scale, length(gene_ids))
  } else {
    scales::rescale(gene_w, to = gene_size_range * size_scale)
  }

  # ── 边 ───────────────────────────────────────────────────────────────────────
  el        <- igraph::get.edgelist(g)
  vtype_map <- setNames(igraph::V(g)$type, igraph::V(g)$name)
  edge_col  <- rep("grey70", nrow(el))
  edge_bold <- rep(FALSE,    nrow(el))

  for (i in seq_len(nrow(el))) {
    n1 <- el[i, 1]; n2 <- el[i, 2]
    if (vtype_map[n1] == "gene") { gn <- n1; mn <- n2 } else { gn <- n2; mn <- n1 }
    shared <- intersect(get_gene_categories(gn), get_module_categories(mn))
    if (length(shared) > 0) {
      edge_col[i]  <- mix_hex_colors(cat_to_col[shared])
      edge_bold[i] <- TRUE
    }
  }

  igraph::E(g)$color <- edge_col
  base_w <- if (edge_by == "snp_geneset") {
    scales::rescale(-log10(igraph::E(g)$p_snp), to = c(0.5, 4))
  } else {
    rep(1, igraph::ecount(g))
  }
  igraph::E(g)$width <- ifelse(edge_bold, pmax(base_w * 2, 3), base_w)

  # ── 标签 ─────────────────────────────────────────────────────────────────────
  igraph::V(g)$label      <- NA_character_
  igraph::V(g)$label.cex  <- ifelse(igraph::V(g)$type == "module", 0.9, gene_label_cex)
  igraph::V(g)$label.color <- "black"
  if (label_modules) {
    for (idx in mod_ids)  igraph::V(g)$label[idx] <- igraph::V(g)$name[idx]
  }
  if (label_genes) {
    for (idx in gene_ids) {
      cats <- get_gene_categories(igraph::V(g)$name[idx])
      if (length(cats) > 0 || label_unclassified_genes)
        igraph::V(g)$label[idx] <- igraph::V(g)$name[idx]
    }
  }

  # ── 绘图 ─────────────────────────────────────────────────────────────────────
  lay <- layout_fun(g)
  if (!is.null(out_pdf)) grDevices::pdf(out_pdf, width = width, height = height)

  plot(g, layout = lay, vertex.label.dist = 0, edge.curved = 0)

  # ── 获取绘图坐标范围，用于定位图例 ──────────────────────────────────────────
  usr     <- graphics::par("usr")          # c(x1, x2, y1, y2)
  x_left  <- usr[1] + 0.02 * (usr[2] - usr[1])
  x_right <- usr[2] - 0.02 * (usr[2] - usr[1])
  y_bot   <- usr[3] + 0.02 * (usr[4] - usr[3])

  # ── 节点类别图例 ─────────────────────────────────────────────────────────────
  leg_labels <- c(
    "Haemoglobin", "Cell cycle", "Autophagy",
    "Cycle + Haem", "Cycle + Auto", "Haem + Auto", "All three",
    "Unclassified (gene)", "Unclassified (module)"
  )
  leg_cols <- c(
    col_haem, col_cycle, col_auto,
    mix_hex_colors(c(col_cycle, col_haem)),
    mix_hex_colors(c(col_cycle, col_auto)),
    mix_hex_colors(c(col_haem,  col_auto)),
    mix_hex_colors(c(col_haem, col_cycle, col_auto)),
    add_alpha(default_gene_col,   unclassified_alpha),
    add_alpha(default_module_col, unclassified_alpha)
  )

  graphics::legend(
    x         = x_left,
    y         = y_bot,
    xjust     = 0, yjust = 0,
    legend    = leg_labels,
    pt.bg     = leg_cols,
    col       = leg_cols,
    pch       = c(rep(19, 8), 15),
    pt.cex    = 2.2,
    cex       = 1.1,
    bty       = "o",
    box.col   = "grey30",
    box.lwd   = 1.2,
    bg        = "white",
    title     = expression(bold("Category")),
    title.col = "black",
    title.cex = 1.2
  )

  # ── 边类别图例 ───────────────────────────────────────────────────────────────
  graphics::legend(
    x         = x_right,
    y         = y_bot,
    xjust     = 1, yjust = 0,
    legend    = c("Haemoglobin (gene–module)", "Cell cycle (gene–module)",
                  "Autophagy (gene–module)",   "No shared category"),
    col       = c(col_haem, col_cycle, col_auto, "grey70"),
    lwd       = c(4, 4, 4, 1.5),
    lty       = 1,
    cex       = 1.1,
    bty       = "o",
    box.col   = "grey30",
    box.lwd   = 1.2,
    bg        = "white",
    title     = expression(bold("Edge (shared category)")),
    title.col = "black",
    title.cex = 1.2
  )

  if (!is.null(out_pdf)) grDevices::dev.off()
  invisible(g)
}
```


The bipartite network connects peripheral gene nodes (circles) to topic module nodes (squares). Color coding:
- **Red** (hemoglobin), **Yellow** (cell cycle), **Blue** (autophagy) — matching the three biological axes described by Ota et al.
- Mixed colors indicate modules/genes belonging to multiple categories.
- Unclassified nodes are shown in grey with reduced opacity.
- Edge color reflects the shared biological category between gene and module; bold edges highlight concordant gene–module pairs.
- Node size scales with the number of modules (for genes) or associated peripheral genes (for modules).



```{r,fig.width=30,fig.height=30}

merged_df <- readRDS("/project/xuanyao/jiaming/paper/output/mediation_analysis/MCH/flames+200kb/k50/merged_df.rds")
plot_gene_module_network_MCH(
  df12              = merged_df,
  sig_GO            = sig_GO,          # 含 module + summary 列
  haemoglobin_genes = haemoglobin_genes,
  cell_cycle_genes  = cell_cycle_genes,
  autophagy_genes   = autophagy_genes,
  top_n_gene        = 800,
 size_scale = 0.3,
  edge_by           = "none",
 unclassified_alpha =0.3
)
```

```{r erythroid-spec-network, fig.width=30, fig.height=30}
target_genes  <- c("KLF1", "TAL1", "GFI1B", "ZFPM1", "MYB", "LDB1", "NFE2")
erythroid_df  <- merged_df %>%
  dplyr::filter(
    sapply(strsplit(merged_gene_name, "-"),
           function(parts) any(parts %in% target_genes))
  )

plot_gene_module_network_MCH(
  df12                     = erythroid_df,
  sig_GO                   = sig_GO,
  haemoglobin_genes        = haemoglobin_genes,
  cell_cycle_genes         = cell_cycle_genes,
  autophagy_genes          = autophagy_genes,
  top_n_gene               = 800,
  size_scale               = 0.3,
  edge_by                  = "none",
  unclassified_alpha       = 0.3,
  label_unclassified_genes = TRUE,
  gene_label_cex           = 2
)
title("Canonical erythroid regulators peripheral genes", cex.main = 2)

setwd("/project/xuanyao/jiaming/paper/output/mediation_analysis/MCH/flames+200kb/k50")
pdf("erythroid_spec_network.pdf", width = 20, height = 20)
plot_gene_module_network_MCH(
  df12                     = erythroid_df,
  sig_GO                   = sig_GO,
  haemoglobin_genes        = haemoglobin_genes,
  cell_cycle_genes         = cell_cycle_genes,
  autophagy_genes          = autophagy_genes,
  top_n_gene               = 800,
  size_scale               = 0.6,
  edge_by                  = "none",
  unclassified_alpha       = 0.3,
  label_unclassified_genes = TRUE,
  gene_label_cex           = 2
)
title("Canonical erythroid regulators peripheral genes", cex.main = 2)
dev.off()
```

```{r, fig.width=30, fig.height=30}
target_genes  <- c("ALAS2", "ABCB10", "TFRC")
erythroid_df  <- merged_df %>%
  dplyr::filter(
    sapply(strsplit(merged_gene_name, "-"),
           function(parts) any(parts %in% target_genes))
  )

plot_gene_module_network_MCH(
  df12                     = erythroid_df,
  sig_GO                   = sig_GO,
  haemoglobin_genes        = haemoglobin_genes,
  cell_cycle_genes         = cell_cycle_genes,
  autophagy_genes          = autophagy_genes,
  top_n_gene               = 800,
  size_scale               = 0.3,
  edge_by                  = "none",
  unclassified_alpha       = 0.3,
  label_unclassified_genes = TRUE,
  gene_label_cex           = 2
)
title("erythroid specification peripheral genes", cex.main = 2)

setwd("/project/xuanyao/jiaming/paper/output/mediation_analysis/MCH/flames+200kb/k50")
pdf("heme_and_iron_biology.pdf", width = 20, height = 20)
plot_gene_module_network_MCH(
  df12                     = erythroid_df,
  sig_GO                   = sig_GO,
  haemoglobin_genes        = haemoglobin_genes,
  cell_cycle_genes         = cell_cycle_genes,
  autophagy_genes          = autophagy_genes,
  top_n_gene               = 800,
  size_scale               = 0.6,
  edge_by                  = "none",
  unclassified_alpha       = 0.3,
  label_unclassified_genes = TRUE,
  gene_label_cex           = 2
)
title("heme and iron biology peripheral genes", cex.main = 2)
dev.off()
```


```{r,include=F}
g <- plot_gene_module_network_MCH(
  df12              = merged_df,
  sig_GO            = sig_GO,
  haemoglobin_genes = haemoglobin_genes,
  cell_cycle_genes  = cell_cycle_genes,
  autophagy_genes   = autophagy_genes,
  top_n_gene        = 800,
  size_scale        = 0.3,
  edge_by           = "none",
  unclassified_alpha = 0.3,
  out_pdf           = NULL   # 不输出PDF，直接画到PNG设备
)

# ↓ 只需要这两行，不要再单独 plot(g)
png("/project/xuanyao/jiaming/paper/output/mediation_analysis/MCH/flames+200kb/k50/gene_module_network.png",
    width = 12000, height = 12000, res = 500)
# 重新调用函数并传入 out_pdf=NULL，让函数内部画到已打开的PNG设备
plot_gene_module_network_MCH(
  df12              = merged_df,
  sig_GO            = sig_GO,
  haemoglobin_genes = haemoglobin_genes,
  cell_cycle_genes  = cell_cycle_genes,
  autophagy_genes   = autophagy_genes,
  top_n_gene        = 800,
  size_scale        = 0.3,
  edge_by           = "none",
  unclassified_alpha = 0.3,
  out_pdf           = NULL
)
dev.off()
```


### Per-module ego networks

For each module, the focal module (square) is placed at the centre; its peripheral gene nodes from `merged_df$summary` (circles) form an inner ring; and any other module that shares at least one peripheral gene is placed in an outer ring (without expanding its own peripheral genes). Colors and edge coloring follow the same logic as `plot_gene_module_network_MCH`.

```{r, fig.width=12, fig.height=12}
plot_module_ego_MCH <- function(focal_module,
                                 df12,
                                 sig_GO,
                                 geneset_categories = NULL,
                                 unclassified_alpha = 0.3,
                                 gene_size      = 6,
                                 focal_size     = 30,
                                 other_mod_size = 22,
                                 label_genes    = TRUE,
                                 label_cex      = 0.75) {

  # ── colour helpers ───────────────────────────────────────────────────────────
  mix_hex_colors <- function(hex_vec) {
    if (length(hex_vec) == 1) return(hex_vec)
    rgb_mat <- grDevices::col2rgb(hex_vec)
    avg <- rowMeans(rgb_mat)
    grDevices::rgb(avg[1], avg[2], avg[3], maxColorValue = 255)
  }
  add_alpha <- function(col, alpha) adjustcolor(col, alpha.f = alpha)

  col_haem  <- "#E74C3C"
  col_cycle <- "#F1C40F"
  col_auto  <- "#3498DB"
  default_module_col <- "grey10"

  get_module_color <- function(mod_name) {
    if (is.null(sig_GO)) return(add_alpha(default_module_col, unclassified_alpha))
    rows <- sig_GO[sig_GO$module == mod_name, , drop = FALSE]
    if (nrow(rows) == 0) return(add_alpha(default_module_col, unclassified_alpha))
    summ <- paste(rows$summary, collapse = " ")
    cats <- character(0)
    if (grepl("hemoglobin|haemoglobin", summ, ignore.case = TRUE)) cats <- c(cats, col_haem)
    if (grepl("cell cycle",             summ, ignore.case = TRUE)) cats <- c(cats, col_cycle)
    if (grepl("autophagy",              summ, ignore.case = TRUE)) cats <- c(cats, col_auto)
    if (length(cats) == 0) return(add_alpha(default_module_col, unclassified_alpha))
    mix_hex_colors(cats)
  }

  # ── data ─────────────────────────────────────────────────────────────────────
  focal_rows <- df12 %>% dplyr::filter(module == focal_module)
  if (nrow(focal_rows) == 0) {
    message("No genes for module: ", focal_module)
    return(invisible(NULL))
  }

  other_mod_rows <- df12 %>%
    dplyr::filter(merged_gene_name %in% focal_rows$merged_gene_name,
                  module != focal_module) %>%
    dplyr::select(module, merged_gene_name) %>%
    dplyr::distinct()

  # ── gene connectivity: # other modules each gene connects to ─────────────────
  gene_conn_tbl <- other_mod_rows %>%
    dplyr::group_by(merged_gene_name) %>%
    dplyr::summarise(n_modules = dplyr::n_distinct(module), .groups = "drop")
  gene_conn_map <- setNames(gene_conn_tbl$n_modules, gene_conn_tbl$merged_gene_name)

  max_conn <- max(c(unlist(gene_conn_map), 1L))  # at least 1 to avoid degenerate palette
  gene_pal <- colorRampPalette(c("#cfe2f3", "#08306b"))(max_conn + 1)

  get_gene_color <- function(raw_name) {
    if (is.null(raw_name) || is.na(raw_name)) return("grey60")
    n <- if (raw_name %in% names(gene_conn_map)) gene_conn_map[[raw_name]] else 0L
    gene_pal[n + 1]
  }

  # ── vertices ─────────────────────────────────────────────────────────────────
  focal_v <- data.frame(name      = focal_module,
                         vtype     = "focal_module",
                         gene_raw  = NA_character_,
                         gene_loci = NA_character_,
                         stringsAsFactors = FALSE)

  gene_v <- focal_rows %>%
    dplyr::transmute(name = summary, vtype = "gene",
                     gene_raw = merged_gene_name, gene_loci = merged_loci)

  other_v <- other_mod_rows %>%
    dplyr::distinct(module) %>%
    dplyr::transmute(name = module, vtype = "other_module",
                     gene_raw = NA_character_, gene_loci = NA_character_)

  verts <- dplyr::bind_rows(focal_v, gene_v, other_v)

  # ── edges ─────────────────────────────────────────────────────────────────────
  edges_focal_gene <- focal_rows %>%
    dplyr::transmute(from = focal_module, to = summary)

  edges_gene_other <- other_mod_rows %>%
    dplyr::left_join(
      focal_rows %>% dplyr::select(merged_gene_name, summary),
      by = "merged_gene_name",
      relationship = "many-to-many"
    ) %>%
    dplyr::transmute(from = summary, to = module) %>%
    dplyr::distinct()

  all_edges <- dplyr::bind_rows(edges_focal_gene, edges_gene_other)

  g <- igraph::graph_from_data_frame(all_edges, directed = FALSE, vertices = verts)

  # ── node attributes ───────────────────────────────────────────────────────────
  is_module_node <- igraph::V(g)$vtype %in% c("focal_module", "other_module")
  igraph::V(g)$shape       <- ifelse(is_module_node, "square", "circle")
  igraph::V(g)$frame.color <- NA
  igraph::V(g)$label.color <- "black"

  igraph::V(g)$color <- vapply(seq_len(igraph::vcount(g)), function(i) {
    if (igraph::V(g)$vtype[i] == "gene")
      get_gene_color(igraph::V(g)$gene_raw[i])
    else
      get_module_color(igraph::V(g)$name[i])
  }, character(1))

  igraph::V(g)$size <- ifelse(igraph::V(g)$vtype == "focal_module", focal_size,
                               ifelse(igraph::V(g)$vtype == "other_module", other_mod_size,
                                      gene_size))

  # module label helper: "kX\nfunction summary"
  get_module_label <- function(mod_name) {
    if (is.null(sig_GO)) return(mod_name)
    rows <- sig_GO[sig_GO$module == mod_name, , drop = FALSE]
    if (nrow(rows) == 0 || all(is.na(rows$summary))) return(mod_name)
    fn <- rows$summary[1]
    fn_wrapped <- paste(strwrap(fn, width = 30), collapse = "\n")
    paste0(mod_name, "\n", fn_wrapped)
  }

igraph::V(g)$label <- vapply(seq_len(igraph::vcount(g)), function(i) {
    vt  <- igraph::V(g)$vtype[i]
    nm  <- igraph::V(g)$name[i]
    raw <- igraph::V(g)$gene_raw[i]
    if (vt == "focal_module") {
      get_module_label(nm)
    } else if (vt == "other_module") {
      nm                          # ← 只显示模块名
    } else {
      if (!label_genes) NA_character_ else raw
    }
  }, character(1))

  igraph::V(g)$label.cex <- ifelse(igraph::V(g)$vtype == "focal_module", 0.85,
                                    ifelse(igraph::V(g)$vtype == "other_module", 0.7,
                                           label_cex))

  # ── edge colours ─────────────────────────────────────────────────────────────
  igraph::E(g)$color <- "grey70"
  igraph::E(g)$width <- 1

  # ── concentric layout ─────────────────────────────────────────────────────────
  focal_idx <- which(igraph::V(g)$vtype == "focal_module")
  gene_idx  <- which(igraph::V(g)$vtype == "gene")
  other_idx <- which(igraph::V(g)$vtype == "other_module")
  n_genes   <- length(gene_idx)
  n_others  <- length(other_idx)

  lay <- matrix(0, nrow = igraph::vcount(g), ncol = 2)
  lay[focal_idx, ] <- c(0, 0)

  if (n_genes > 0) {
    th <- seq(0, 2 * pi, length.out = n_genes + 1)[seq_len(n_genes)]

    conn_layout <- vapply(igraph::V(g)$gene_raw[gene_idx], function(r) {
      if (!is.null(r) && !is.na(r) && r %in% names(gene_conn_map)) gene_conn_map[[r]] else 0L
    }, integer(1))
    connected_pos   <- which(conn_layout > 0)
    unconnected_pos <- which(conn_layout == 0)
    n_connected     <- length(connected_pos)

    if (n_connected > 0 && n_connected < n_genes) {
      connected_slots <- round(seq(1, n_genes, length.out = n_connected))
      new_order                          <- integer(n_genes)
      new_order[connected_slots]           <- connected_pos
      new_order[setdiff(seq_len(n_genes), connected_slots)] <- unconnected_pos
      gene_idx_ord <- gene_idx[new_order]
    } else {
      gene_idx_ord <- gene_idx
    }

    lay[gene_idx_ord, 1] <- cos(th)
    lay[gene_idx_ord, 2] <- sin(th)
  }

  if (n_others > 0) {
    gene_angles  <- setNames(atan2(lay[gene_idx, 2], lay[gene_idx, 1]),
                              igraph::V(g)$name[gene_idx])
    other_names  <- igraph::V(g)$name[other_idx]
    for (j in seq_along(other_idx)) {
      connected_summaries <- other_mod_rows %>%
        dplyr::filter(module == other_names[j]) %>%
        dplyr::left_join(focal_rows %>% dplyr::select(merged_gene_name, summary),
                         by = "merged_gene_name") %>%
        dplyr::pull(summary)
      valid_angles <- gene_angles[intersect(connected_summaries, names(gene_angles))]
      mean_angle <- if (length(valid_angles) > 0)
        atan2(mean(sin(valid_angles)), mean(cos(valid_angles)))
      else
        2 * pi * (j - 1) / n_others
      lay[other_idx[j], 1] <- 2 * cos(mean_angle)
      lay[other_idx[j], 2] <- 2 * sin(mean_angle)
    }
  }

  # ── title ─────────────────────────────────────────────────────────────────────
  detail_label <- if (!is.null(geneset_categories)) {
    cat_row <- geneset_categories[geneset_categories$module == focal_module, , drop = FALSE]
    if (nrow(cat_row) > 0 && !is.na(cat_row$detail[1]) && nzchar(cat_row$detail[1]))
      cat_row$detail[1]
    else
      NA_character_
  } else NA_character_

  main_title <- if (!is.na(detail_label))
    paste0("Module ", focal_module, "\n", detail_label)
  else
    paste0("Module ", focal_module)

  # ── per-node label placement ──────────────────────────────────────────────────
  n_v          <- igraph::vcount(g)
  lbl_dist     <- rep(0.5, n_v)
  lbl_degree   <- rep(-pi / 2, n_v)

  for (i in seq_len(n_v)) {
    if (igraph::V(g)$vtype[i] == "gene") {
      angle            <- atan2(lay[i, 2], lay[i, 1])
      lbl_degree[i]    <- angle
      lbl_dist[i]      <- 1.5
    } else if (igraph::V(g)$vtype[i] == "other_module") {
      # other module 标签也向外辐射，距离稍大以避免与节点重叠
      angle            <- atan2(lay[i, 2], lay[i, 1])
      lbl_degree[i]    <- angle
      lbl_dist[i]      <- 1.2
    }
  }

  igraph::V(g)$label.dist   <- lbl_dist
  igraph::V(g)$label.degree <- lbl_degree

  plot(g, layout = lay,
       main        = main_title,
       edge.curved = 0.2)

  # ── legend: gene connectivity gradient ───────────────────────────────────────
  legend_n      <- min(max_conn + 1, 6)
  legend_counts <- round(seq(0, max_conn, length.out = legend_n))
  legend_cols   <- gene_pal[legend_counts + 1]
  legend_labels <- ifelse(legend_counts == 0, "0 other modules",
                          paste0(legend_counts, " other module",
                                 ifelse(legend_counts == 1, "", "s")))
  legend(
    "bottomright",
    legend = legend_labels,
    pt.bg  = legend_cols,
    col    = "black",
    pch    = 21,
    pt.cex = 1.6,
    bty    = "n",
    cex    = 0.8,
    title  = expression(bold("# connected modules"))
  )

  invisible(g)
}
```


```{r plot-module-ego-MCH, fig.width=12, fig.height=12}
library(readxl)
geneset_categories <- read_excel("/project/xuanyao/jiaming/paper/output/mediation_analysis/MCH/flames+200kb/k50/sig_GO_MCH_k50.xlsx")
setwd("/project/xuanyao/jiaming/paper/output/mediation_analysis/MCH/flames+200kb/k50")

for (mod in sort(unique(merged_df$module))) {
  plot_module_ego_MCH(
    focal_module       = mod,
    df12               = merged_df,
    sig_GO             = sig_GO,
    geneset_categories = geneset_categories,
    unclassified_alpha = 0.3
  )
  pdf(paste0("MCH_ego_", mod, ".pdf"), width = 12, height = 12)
  plot_module_ego_MCH(
    focal_module       = mod,
    df12               = merged_df,
    sig_GO             = sig_GO,
    geneset_categories = geneset_categories,
    unclassified_alpha = 0.3
  )
  dev.off()
}
```

###————————April 12 Update——————————————

#### Summary: how many modules does each peripheral gene regulate?

```{r fig.width=7, fig.height=4}
gene_module_counts <- merged_df %>%
  filter(merged_gene_name != "") %>%
  separate_rows(merged_gene_name, sep = "-") %>%
  filter(merged_gene_name != "") %>%
  distinct(merged_gene_name, module) %>%
  count(merged_gene_name, name = "n_modules")

ggplot(gene_module_counts, aes(x = n_modules)) +
  geom_histogram(binwidth = 1, fill = "#4C72B0", color = "white") +
  scale_x_continuous(breaks = seq(1, max(gene_module_counts$n_modules), by = 1)) +
  labs(
    title = "Distribution of modules regulated per peripheral gene",
    x = "Number of modules regulated",
    y = "Number of genes"
  ) +
  theme_bw(base_size = 13)
```

```{r}
which(table(merged_df$merged_gene_name)>17)

gene_symbols <- c("EIF2B5", "MED12", "PTPN1", "TRRAP", "XRN2")

# 用 gtf_df 做映射
result <- data.frame(
  symbol = gene_symbols
) %>%
  left_join(gtf_df, by = "symbol")

result
```




#### Summary: how many peripheral genes regulate each module?

```{r fig.width=9, fig.height=5}
module_gene_counts <- merged_df %>%
  filter(merged_gene_name != "") %>%
  separate_rows(merged_gene_name, sep = "-") %>%
  filter(merged_gene_name != "") %>%
  distinct(module, merged_gene_name) %>%
  count(module, name = "n_genes") %>%
  arrange(desc(n_genes)) %>%
  mutate(module = factor(module, levels = module))

ggplot(module_gene_counts, aes(x = module, y = n_genes)) +
  geom_bar(stat = "identity", fill = "#DD8452") +
  labs(
    title = "Number of peripheral genes regulating each module (sorted)",
    x = "Module",
    y = "Number of regulating genes"
  ) +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Gene-level evidence within significant mediating modules

To verify that the significant mediation signals reflect genuine gene-set–level effects (rather than being driven by a small number of outlier genes), we examine the distribution of gene-level MCH p-values within each significant (perturbation, module) pair.

For each pair, we estimate **π₁** (the proportion of true non-null associations) using Storey's method: π₁ = 1 − π₀, where π₀ is estimated from the fraction of p-values > 0.5. If mediation were driven only by one or two extreme genes, the majority of p-values within the module would be uniform and π₁ would be low. Instead, we observe elevated π₁ across genes in significant modules, indicating that multiple genes within the same co-regulated program contribute to the MCH signal.

This analysis also provides empirical support for the **omnigenic model**: each significant gene set is associated with many distinct peripheral gene perturbations, rather than being regulated by only a few specific genes.

Key comparison with Ota et al.: of 13 Ota et al. MCH-pathway genes in our 1,297-gene test set (CLSPN, SPRED2, CAD, POLE, CBFA2T3, CCNA2, KLF1, EIF3F, CALR, EIF2B4, ABCB10, ATR, SYMPK), **11 are identified as significant peripheral genes by OPERA**, and many appear as members of the significant gene-set modules.

```{r}

filtered_list_50 <- readRDS("/project/xuanyao/jiaming/Getting_started/data/repologle/topic_geneset/flames+200kb/de_50_0.01_2_0.5.rds")
genewise_p_value <- readRDS("/project/xuanyao/jiaming/paper/output/snake/repologle_MCH/flames+200kb/K50/3000_t_ACAT_resample/p_values/genewise_p_value.rds")
genewise_p_value[genewise_p_value==0] <- 1e-100


# ============================================================
# Step 1: 找 key SNPs（与 >5 个 module 有联系的 SNP）
# ============================================================

snp_module_count <- dfl2 %>%
  group_by(snp) %>%
  summarise(n_modules = n_distinct(module)) %>%
  arrange(desc(n_modules))

key_snps <- snp_module_count %>%
  filter(n_modules > 0) %>%
  pull(snp)

# 报告比例
total_snps <- n_distinct(dfl2$snp)
n_key_snps <- length(key_snps)

cat("Key SNPs数量:", n_key_snps, "\n")
cat("总SNP数量:", total_snps, "\n")
cat("Key SNP比例:", round(n_key_snps / total_snps * 100, 2), "%\n")

# ============================================================
# Step 2: 对每个 key SNP，找联系的 module，再找 module 中的基因
# 然后做 Bonferroni 校正，统计 significant genes
# ============================================================

# genewise_p_value 格式：行是 SNP，列是 gene
# 先转成长格式方便操作
p_long <- genewise_p_value %>%
  as.data.frame() %>%
  rownames_to_column("snp") %>%
  pivot_longer(-snp, names_to = "gene_id", values_to = "p_value")

# Bonferroni 校正阈值（按每个 snp-module 组合的基因数校正）
library(qvalue)

results_list <- list()
pval_list    <- list()   # 新增：储存每个 snp-module 的 p value 向量

# 设为 NULL 则计算所有 guide；设为具体 guide ID 则只计算该 guide
target_guide <- c("gRNA:ENSG00000145191","gRNA:ENSG00000184634","gRNA:ENSG00000196396","gRNA:ENSG00000196367","gRNA:ENSG00000088930")
if (!is.null(target_guide)) key_snps <- intersect(key_snps, target_guide)

for (i in seq_along(key_snps)) {
  s <- key_snps[i]
  cat(sprintf("Processing SNP %d/%d: %s\n", i, length(key_snps), s))
  
  modules_for_snp <- dfl2 %>%
    filter(snp == s) %>%
    pull(module) %>%
    unique()
  
  for (m in modules_for_snp) {
    genes_in_module <- filtered_list_50[[m]]
    if (is.null(genes_in_module) || length(genes_in_module) == 0) next
    
    p_subset <- p_long %>%
      filter(snp == s, gene_id %in% genes_in_module) %>%
      filter(!is.na(p_value))
    if (nrow(p_subset) == 0) next
    
    key <- paste(s, m, sep = "_")
    
    # ── 原有：计算 pi1 ──────────────────────────────────────
    pi1_val <- tryCatch({
      qobj <- qvalue(p = p_subset$p_value)
      1 - qobj$pi0
    }, error = function(e) NA_real_)
    
    results_list[[key]] <- data.frame(
      snp               = s,
      module            = m,
      n_genes_in_module = nrow(p_subset),
      pi1               = round(pi1_val, 4)
    )
    
    # ── 新增：保存 p value 向量 ──────────────────────────────
    pval_list[[key]] <- p_subset %>%
      dplyr::select(gene_id, p_value)          # 保留基因名，方便后续对照
  }
}

results_df  <- bind_rows(results_list)

# pval_df：长格式，包含 snp、module、gene_id、p_value 四列
pval_df <- bind_rows(
  lapply(names(pval_list), function(k) {
    parts <- strsplit(k, "_")[[1]]
    # key 格式是 "snp_module"，snp 本身可能含下划线，所以用 results_df 对应
    pval_list[[k]] %>%
      mutate(snp_module_key = k)
  })
) %>%
  left_join(
    results_df %>% dplyr::select(snp, module) %>%
      mutate(snp_module_key = paste(snp, module, sep = "_")),
    by = "snp_module_key"
  ) %>%
  dplyr::select(snp, module, gene_id, p_value)


```



```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

plot_guide_heatmap <- function(guide_id, pval_df) {
  
  df <- pval_df %>% filter(snp == guide_id)
  
  all_genes   <- unique(df$gene_id)
  all_modules <- unique(df$module)
  
  # 构建完整 module x gene 矩阵
  full_df <- expand.grid(module = all_modules, gene_id = all_genes, stringsAsFactors = FALSE) %>%
    left_join(df %>% dplyr::select(module, gene_id, p_value), by = c("module", "gene_id")) %>%
    mutate(has_value = !is.na(p_value))
  
  # ── 基因排序：按 module membership 聚集 ──────────────────
  # 每个 gene 的"primary module"= 它出现在哪些 module 里
  # 策略：先按 module 数量分组，再按最小 p_value 的 module 排序
gene_order <- df %>%
    filter(!is.na(p_value)) %>%           # 先去掉 NA
    mutate(p_value = as.numeric(p_value)) %>%  # 确保是 numeric，不是 list
    group_by(gene_id) %>%
    slice_min(p_value, n = 1, with_ties = FALSE) %>%  # 替换 arrange + slice(1)
    ungroup() %>%
    arrange(module, p_value) %>%
    pull(gene_id)
  
  # 有些 gene 可能只在 full_df 里（expand.grid 补全的），补到末尾
  remaining_genes <- setdiff(all_genes, gene_order)
  gene_order_full <- c(gene_order, remaining_genes)
  
  # ── Module 排序：按各自基因数量降序 ──────────────────────
  module_order <- df %>%
    group_by(module) %>%
    summarise(n_genes = n_distinct(gene_id), .groups = "drop") %>%
    arrange(desc(n_genes)) %>%
    pull(module)
  
  # 应用排序
  full_df <- full_df %>%
    mutate(
      gene_id = factor(gene_id, levels = gene_order_full),
      module  = factor(module,  levels = module_order)
    )
  
  # ── 添加 module 分隔线位置 ────────────────────────────────
  # 统计每个 module 在 x 轴上占据的基因位置，画竖线分隔
  module_boundaries <- df %>%
    mutate(gene_id = factor(gene_id, levels = gene_order_full)) %>%
    group_by(module) %>%
    summarise(
      max_pos = max(as.integer(gene_id), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(max_pos) %>%
    # 边界线画在每个 module 最后一个基因之后 0.5 处
    mutate(xintercept = max_pos + 0.5) %>%
    filter(xintercept < length(gene_order_full) + 0.5)  # 去掉最右边界
  
  ggplot(full_df, aes(x = gene_id, y = module)) +
    geom_tile(data = filter(full_df, !has_value), fill = "grey92", color = NA) +
    geom_tile(data = filter(full_df, has_value),  aes(fill = p_value), color = NA) +
    # module 分隔竖线
    geom_vline(
      data        = module_boundaries,
      aes(xintercept = xintercept),
      color       = "grey40",
      linewidth   = 0.4,
      linetype    = "dashed",
      inherit.aes = FALSE
    ) +
    scale_fill_gradient(
      low      = "red",
      high     = "white",
      limits   = c(0, 1),
      name     = "p-value",
      na.value = "grey92"
    ) +
    labs(
      title = paste("Guide:", guide_id),
      x     = "Gene (ordered by module membership)",
      y     = "Module"
    ) +
    theme_minimal() +
    theme(
      axis.text.x  = element_blank(),
      axis.ticks.x = element_blank(),
      panel.grid   = element_blank(),
      plot.title   = element_text(hjust = 0.5)
    )
}

library(ggplot2)
library(dplyr)
library(tidyr)

plot_module_qqplot <- function(guide_id, pval_df) {
  
  df <- pval_df %>% filter(snp == guide_id)
  
  all_modules <- unique(df$module)
  
  # 为每个 module 生成 QQ plot 数据
  qq_df <- lapply(all_modules, function(mod) {
    p_vals <- df %>%
      filter(module == mod) %>%
      pull(p_value) %>%
      na.omit() %>%
      sort()
    
    n <- length(p_vals)
    if (n == 0) return(NULL)
    
    expected <- -log10((seq_len(n)) / (n + 1))
    observed <- -log10(p_vals)
    
    data.frame(
      module   = mod,
      expected = expected,
      observed = observed
    )
  }) %>%
    bind_rows()
  
  # 计算 lambda 和 pi1 per module
  stats_df <- df %>%
    group_by(module) %>%
    summarise(
      p_vals = list(na.omit(p_value)),
      .groups = "drop"
    ) %>%
    rowwise() %>%
    mutate(
      n = length(p_vals),
      
      # lambda：基因组膨胀因子，用 p-value 对应的卡方统计量中位数估算
      lambda = ifelse(
        n > 1,
        round(
          median(qchisq(p_vals, df = 1, lower.tail = FALSE), na.rm = TRUE) /
            qchisq(0.5, df = 1, lower.tail = FALSE),
          3
        ),
        NA_real_
      ),
      
      # pi1：真实信号比例 = 1 - pi0
      # 用 Storey 方法估算 pi0：在 p > lambda_thresh 区域，p-value 近似均匀分布
      # pi0 ≈ #{p > 0.5} / (n * 0.5)，lambda_thresh 取 0.5
      pi1 = ifelse(
        n > 1,
        round(
          1 - sum(p_vals > 0.5) / (n * 0.5),
          3
        ),
        NA_real_
      ),
      pi1 = pmax(0, pmin(1, pi1)),   # 截断到 [0, 1]
      
      label = paste0("λ = ", lambda, "\nπ₁ = ", pi1)
    ) %>%
    ungroup()
  
  # 把 label 加入 qq_df 用于 facet 内标注
  label_df <- qq_df %>%
    group_by(module) %>%
    summarise(
      x_pos = 0,
      y_pos = max(observed, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(stats_df %>% dplyr::select(module, label), by = "module")
  
  ggplot(qq_df, aes(x = expected, y = observed)) +
    geom_abline(intercept = 0, slope = 1,
                color = "red", linetype = "dashed", linewidth = 0.6) +
    geom_point(alpha = 0.6, size = 0.8, color = "steelblue") +
    geom_text(
      data    = label_df,
      aes(x = x_pos, y = y_pos, label = label),
      hjust   = 0, vjust = 1,
      size    = 2.8, color = "black",
      inherit.aes = FALSE
    ) +
    facet_wrap(~ module, scales = "free") +
    labs(
      title = paste("QQ Plot by Module — Guide:", guide_id),
      x     = expression(Expected ~ -log[10](p)),
      y     = expression(Observed ~ -log[10](p))
    ) +
    theme_minimal() +
    theme(
      strip.text  = element_text(face = "bold", size = 9),
      panel.grid  = element_blank(),
      axis.line   = element_line(color = "grey50"),
      plot.title  = element_text(hjust = 0.5)
    )
}

```

```{r}
plot_guide_heatmap("gRNA:ENSG00000184634", pval_df)
plot_module_qqplot("gRNA:ENSG00000184634", pval_df)
plot_guide_heatmap("gRNA:ENSG00000196367", pval_df)
plot_module_qqplot("gRNA:ENSG00000196367", pval_df)
plot_guide_heatmap("gRNA:ENSG00000088930", pval_df)
plot_module_qqplot("gRNA:ENSG00000088930", pval_df)
plot_guide_heatmap("gRNA:ENSG00000196396", pval_df)
plot_module_qqplot("gRNA:ENSG00000196396", pval_df)
plot_guide_heatmap("gRNA:ENSG00000145191", pval_df)
plot_module_qqplot("gRNA:ENSG00000145191", pval_df)
```


```{r,include=F}

results_key <- results_df %>%
  filter(snp %in% key_snps, !is.na(pi1))

pi1_all <- results_df$pi1[!is.na(results_df$pi1)]
pi1_key <- results_key$pi1

# 先画 all SNPs（freq=FALSE 直接输出密度，不要手动改 $density）
hist(pi1_all,
     breaks = 200,
     freq   = FALSE,
     col    = rgb(0.2, 0.4, 0.8, 0.5),
     border = NA,
     xlim   = c(0, 1),
     xlab   = "Pi1",
     ylab   = "Density",
     main   = "Pi1 distribution: all SNPs vs key SNPs")

# 叠加 key SNPs
hist(pi1_key,
     breaks = 200,
     freq   = FALSE,
     col    = rgb(0.9, 0.3, 0.2, 0.5),
     border = NA,
     add    = TRUE)

legend("topright",
       legend = c("All peripheral genes", "Key peripheral genes (>5 modules)"),
       fill   = c(rgb(0.2, 0.4, 0.8, 0.5),
                  rgb(0.9, 0.3, 0.2, 0.5)),
       border = NA,
       bty    = "n")
```






## Negative control: non-targeting perturbations

To assess calibration of the mediation framework, we repeat the full pipeline using **514 non-targeting perturbations** from the Replogle dataset. Non-targeting perturbations do not disrupt any gene, so they should not exhibit systematic associations with MCH. The resulting 24,158 negative-control (perturbation, gene-set, MCH) triples serve as an empirical null.

Under a 5% FDR threshold applied to the AdaFilter decisions from this negative-control run, only 18 pairs (0.07%) are called significant — demonstrating excellent calibration and strong control of false positives across the full pipeline.

```{r}
library(data.table)
library(dplyr)
library(stringr)
library(purrr)
library(heavytailcombtest)

# -----------------------------
# 0) Read inputs
# Non-targeting perturbation IDs are indexed from grna_names.rds
# using the logical vector in non_targeting_guide.rds
# -----------------------------
df <- fread("/project/xuanyao/jiaming/Getting_started/data/geneset_to_trait/MCH/MCH_summary.tsv")
df2 <- df[grepl("M1.01", df$Name), ]

Repologle_associated_perturbation <- readRDS("/project/xuanyao/jiaming/Getting_started/data/repologle/grna_names.rds")[readRDS("/project/xuanyao/jiaming/Getting_started/data/repologle/non_targeting_guide.rds")]



discovery_pairs_cis <- readRDS("/project/xuanyao/jiaming/Getting_started/data/repologle/discovery_pairs_cis.rds")


gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/repologle/topic_geneset/flames+200kb/de_50_0.01_2_0.5.rds")




# -----------------------------
# 1) Build pv_map from df2
# -----------------------------
gene_id <- str_extract(df2$Name, "(?<=\\()[^()]+(?=\\))")

lookup <- df2 %>%
  mutate(ID = gene_id) %>%
  filter(!is.na(ID), !is.na(p_value)) %>%
  group_by(ID) %>%
  summarise(p_value = min(p_value), .groups = "drop")

pv_map <- setNames(lookup$p_value, lookup$ID)


# -----------------------------
# 2) Helpers
# -----------------------------
clean_p <- function(p) {
  p <- as.numeric(p)
  p[is.na(p) | p <= 0] <- NA
  p
}

safe_comb <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  combination.test(
    p,
    weights = NA,
    method = "Cauchy",
    tail.idx = 1
  )
}

safe_comb_fisher <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  stat <- -2 * sum(log(p))
  stats::pchisq(stat, df = 2 * length(p), lower.tail = FALSE)
}


# -----------------------------
# 3) Make filtered modules per perturbation (and record n_removed)
# -----------------------------
make_filtered_modules <- function(gene_symbol, gene_modules_id, pairs_df) {
  target_name <- sub("^gRNA:", "", gene_symbol)
  # print(target_name)
  # print(pairs_df$grna_target)

  ids_to_remove <- unique(pairs_df$response_id[pairs_df$grna_target == target_name])

  filtered_modules <- lapply(gene_modules_id, function(mod_ids) {
    setdiff(mod_ids, ids_to_remove)
  })

  list(
    modules   = filtered_modules,
    n_removed = length(ids_to_remove)
  )
}

filtered_by_perturbation <- setNames(
  lapply(
    Repologle_associated_perturbation,
    make_filtered_modules,
    gene_modules_id = gene_modules_id,
    pairs_df = discovery_pairs_cis
  ),
  Repologle_associated_perturbation
)


# -----------------------------
# 4) Compute module-level combined p-values per perturbation
#    (and add n_removed into module_results)
# -----------------------------
compute_for_one_pert <- function(pert) {
  obj <- filtered_by_perturbation[[pert]]
  gene_modules_id_pert <- obj$modules
  n_removed <- obj$n_removed

  # module -> p-values vector under this perturbation
  module_pvals <- lapply(gene_modules_id_pert, function(genes) {
    p <- pv_map[genes]
    names(p) <- genes
    p[!is.na(p)]
  })

  # combine within each module
  p_acat <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb(p))
  })

  p_fisher <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb_fisher(p))
  })

  tibble(
    snp       = pert,
    n_removed = n_removed, 
    module    = names(module_pvals),
    n_genes   = lengths(gene_modules_id_pert),
    n_matched = sapply(module_pvals, function(v) sum(!is.na(v))),
    p_acat    = p_acat,
    p_fisher  = p_fisher
  )
}

module_results <- map_dfr(Repologle_associated_perturbation, compute_for_one_pert)


```

```{r}
library(tidyr)
library(dplyr)
library(tibble)

# Load the same OPERA p-value matrix used for the primary MCH analysis.
# For the negative control we use only the non-targeting rows (snp_geneset_p_value_control).
snp_geneset_p_value<- readRDS("/project/xuanyao/jiaming/paper/output/snake/repologle_MCH/flames+200kb/K50/3000_t_ACAT_resample/p_values/geneset_p_value.rds")
snp_geneset_p_value_test <- snp_geneset_p_value[which(grepl("ENSG",rownames(snp_geneset_p_value))),]
snp_geneset_p_value_control <- snp_geneset_p_value[which(grepl("non",rownames(snp_geneset_p_value))),]
df1_long <- snp_geneset_p_value_control %>%
  as.data.frame() %>%
  rownames_to_column("snp") %>%      # 这里真正把 rownames 变成一列 snp
  pivot_longer(
    -snp,
    names_to  = "module",
    values_to = "p_snp_geneset"
  )

## 2. 模块层面的表，只保留需要的列
df2 <- module_results %>%
  dplyr::select(module, p_fisher, everything())

## 3. 按 module 合并，再调整列顺序
merged <- df1_long %>%
  left_join(df2, by = c("snp", "module")) %>%  # 关键改动
  relocate(snp, module, p_snp_geneset, p_fisher)


# library(dplyr)
# 
# df1 <- as.data.frame(snp_geneset_p_value)
# df1$snp <- rownames(df1)
# 
# df1_long <- df1 %>%
#   pivot_longer(-snp, names_to = "module", values_to = "p_snp_geneset")
# 
# ## 图二：geneset_acat_pvals_M1.0001（列：module, p_acat）
# df2 <- geneset_acat_pvals_M1.0001 %>%
#   dplyr::select(module, p_fisher)
# 
# merged <- df1_long %>%
#   left_join(df2, by = "module") %>%
#   relocate(snp, module, p_snp_geneset, p_fisher)
# 
# 

X <- merged[,c(3,4)]

library(adaFilter)

result <- adaFilter(X, r = 2,alpha=0.05)

merged$decision <- result$decision
colnames(merged)[c(3,4)] <- c("snp_geneset","geneset_trait")

head(merged[merged$decision==1,])
```

# Asthma analysis (Nadig dataset)

The same three-stage mediation framework is applied to **asthma** (GWAS: GCST010043) using the **Nadig** Perturb-seq dataset (T cells). GWAS loci were mapped to candidate effector genes using both a 200 kb window and FLAMES, yielding **75 asthma-relevant perturbations**.

Topic modeling was applied to perturbation-receiving cells (split-sample design as for MCH), and the resulting gene sets were used as candidate mediators. Because the Nadig dataset is smaller and from a different cell type (immune rather than erythroid), a different optimal number of topics (K = 20 or K = 10/15 for different locus-mapping strategies) was used. The analysis demonstrates that the framework generalizes across traits with distinct genetic architectures, perturbation coverage scales, and cellular contexts.

## Asthma, FLAMES / 200 kb locus mapping, K = 20

```{r}
library(gprofiler2)
 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/topic_geneset/flames_200kb/de_20_0.01_2_0.5.rds")

 de_10_0.01_2_0.5 <-gene_modules_id

gene_sets <- lapply(de_10_0.01_2_0.5, function(x) unique(as.character(x)))

# gene_sets <- lapply(de_10_0.01_2_0.5 , function(x) unique(as.character(x)))

# 建议给每个 set 一个名字（没有就补上）
if (is.null(names(gene_sets))) names(gene_sets) <- paste0("set_", seq_along(gene_sets))

res <- gost(
  query = gene_sets,                 # list of 100 gene vectors
  organism = "hsapiens",
  multi_query = TRUE,
  correction_method = "fdr",
  user_threshold = 0.05,
  sources = c("GO:BP","GO:MF","KEGG")
)



# geneset 名字：如果你有 gene_sets 的 names 就用它；否则自动生成
gs_names <- names(gene_sets)
if (is.null(gs_names)) gs_names <- paste0("set_", seq_along(res$result$significant[[1]]))

# GO term 名字：用 term_id 或 term_name（建议 term_id 更稳定）
go_names <- paste0(res$result$source, ":", res$result$term_id)
# 如果你更想用可读名字：
# go_names <- paste0(res$result$source, ":", res$result$term_name)

# 1) list -> 矩阵：每列一个 term，每行一个 geneset
sig_term_by_set <- do.call(cbind, res$result$significant)
storage.mode(sig_term_by_set) <- "logical"

# 2) 转成“geneset(行) × term(列)”的 index matrix
idx_mat <- sig_term_by_set  # 现在行=geneset, 列=term（通常就是这个方向）
rownames(idx_mat) <- gs_names
colnames(idx_mat) <- go_names


# 如果你发现方向反了（极少数情况），就用这一句修正：
# idx_mat <- t(idx_mat)

library(dplyr)
library(tidyr)

go_tbl <- as_tibble(res$result)

# geneset 名称（如果你有 names(gene_sets) 就用它；否则自动生成）
G <- length(go_tbl$p_values[[1]])
gs_names <- names(gene_sets)
if (is.null(gs_names) || length(gs_names) != G) gs_names <- paste0("set_", seq_len(G))

# 展开：term × geneset 的长表
go_long <- go_tbl %>%
  mutate(
    query   = list(gs_names),
    p_value = p_values,
    sig     = significant       # significant 列也是 list，记录每个 geneset 是否显著
  ) %>%
  dplyr::select(source, term_id, term_name, query, p_value, sig) %>%
  unnest(c(query, p_value, sig)) %>%
  filter(sig == TRUE)  

# 每个 geneset 取 top10（p_value 最小）
top10_tbl <- go_long %>%
  group_by(query) %>%
  arrange(p_value, .by_group = TRUE) %>%
  ungroup()

# 你想要的：每个 geneset 的 top10 term_name（以及可选的 p_value）
top10_names <- top10_tbl %>%
  group_by(query) %>%
  summarise(top10_term_name = list(term_name), .groups = "drop")



# 注意！！！！！！！！！！！！！！！是按 1， 10 ，11 等排序
```

```{r}
top10_names

```




### Geneset–trait associations and AdaFilter mediation test (Asthma, K20, flames_200kb)

Same pipeline as MCH: Fisher aggregation of gene-level asthma burden-test p-values within topic modules (with cis-gene exclusion), joined to OPERA gene→gene-set p-values, and tested jointly with AdaFilter. Note: `alpha=0.1` is used here given the smaller number of perturbations (75 vs. 1,297 for MCH), which reduces statistical power.

```{r}
library(data.table)
library(dplyr)
library(stringr)
library(purrr)
library(heavytailcombtest)

# -----------------------------
# 0) Read inputs
# -----------------------------
df <- fread("/project/xuanyao/jiaming/Getting_started/data/geneset_to_trait/asthma/asthma.tsv")
df2 <- df[grepl("M1.01", df$Name), ]

Nadig_associated_perturbation <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/asthma/GCST010043/flames_200kb/Nadig_associated_perturbation_flames_200kb.rds")



discovery_pairs_cis <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/discovery_pairs_cis.rds")


 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/topic_geneset/flames_200kb/de_20_0.01_2_0.5.rds")

all_measured_genes <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/all_measured_genes.rds")

# 只保留测到的 genes
gene_modules_id <- lapply(gene_modules_id, function(vec) intersect(vec, all_measured_genes))


# -----------------------------
# 1) Build pv_map from df2
# -----------------------------
gene_id <- str_extract(df2$Name, "(?<=\\()[^()]+(?=\\))")

lookup <- df2 %>%
  mutate(ID = gene_id) %>%
  filter(!is.na(ID), !is.na(p_value)) %>%
  group_by(ID) %>%
  summarise(p_value = min(p_value), .groups = "drop")

pv_map <- setNames(lookup$p_value, lookup$ID)


# -----------------------------
# 2) Helpers
# -----------------------------
clean_p <- function(p) {
  p <- as.numeric(p)
  p[is.na(p) | p <= 0] <- NA
  p
}

safe_comb <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  combination.test(
    p,
    weights = NA,
    method = "Cauchy",
    tail.idx = 1
  )
}

safe_comb_fisher <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  stat <- -2 * sum(log(p))
  stats::pchisq(stat, df = 2 * length(p), lower.tail = FALSE)
}


# -----------------------------
# 3) Make filtered modules per perturbation (and record n_removed)
# -----------------------------
make_filtered_modules <- function(gene_symbol, gene_modules_id, pairs_df) {
  target_name <- paste0("tar-",gene_symbol)

  ids_to_remove <- unique(pairs_df$response_id[pairs_df$grna_target == target_name])

  filtered_modules <- lapply(gene_modules_id, function(mod_ids) {
    setdiff(mod_ids, ids_to_remove)
  })

  list(
    modules   = filtered_modules,
    n_removed = length(ids_to_remove)
  )
}

filtered_by_perturbation <- setNames(
  lapply(
    Nadig_associated_perturbation,
    make_filtered_modules,
    gene_modules_id = gene_modules_id,
    pairs_df = discovery_pairs_cis
  ),
  Nadig_associated_perturbation
)


# -----------------------------
# 4) Compute module-level combined p-values per perturbation
#    (and add n_removed into module_results)
# -----------------------------
compute_for_one_pert <- function(pert) {
  obj <- filtered_by_perturbation[[pert]]
  gene_modules_id_pert <- obj$modules
  n_removed <- obj$n_removed

  # module -> p-values vector under this perturbation
  module_pvals <- lapply(gene_modules_id_pert, function(genes) {
    p <- pv_map[genes]
    names(p) <- genes
    p[!is.na(p)]
  })

  # combine within each module
  p_acat <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb(p))
  })

  p_fisher <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb_fisher(p))
  })

  tibble(
    snp       = pert,
    n_removed = n_removed,  # ✅ 每个 perturbation 的 ids_to_remove 数量（复制到每一行 module）
    module    = names(module_pvals),
    n_genes   = lengths(gene_modules_id_pert),
    n_matched = sapply(module_pvals, function(v) sum(!is.na(v))),
    p_acat    = p_acat,
    p_fisher  = p_fisher
  )
}

module_results <- map_dfr(Nadig_associated_perturbation, compute_for_one_pert)

# 查看结果
# module_results %>%
#   arrange(p_fisher) %>%
#   print(n = 30)

```


```{r}
library(tidyr)
library(dplyr)
library(tibble)

snp_geneset_p_value<-  readRDS("/project/xuanyao/jiaming/paper/output/snake/Nadig_asthma/flames_200kb/K20/3000_t_ACAT_resample/p_values/geneset_p_value.rds")
snp_geneset_p_value_test <- snp_geneset_p_value[which(!grepl("non-",rownames(snp_geneset_p_value))),]
snp_geneset_p_value_control <- snp_geneset_p_value[which(grepl("non-",rownames(snp_geneset_p_value))),]

df1_long <- snp_geneset_p_value_test %>%
  as.data.frame() %>%
  rownames_to_column("snp") %>%      # 这里真正把 rownames 变成一列 snp
  pivot_longer(
    -snp,
    names_to  = "module",
    values_to = "p_snp_geneset"
  )

## 2. 模块层面的表，只保留需要的列
df2 <- module_results %>%
  dplyr::select(module, p_fisher,snp)

## 3. 按 module 合并，再调整列顺序
merged <- df1_long %>%
  left_join(df2, by = c("snp", "module")) %>%  # 关键改动
  relocate(snp, module, p_snp_geneset, p_fisher)%>% tidyr::drop_na()


# library(dplyr)
# 
# df1 <- as.data.frame(snp_geneset_p_value)
# df1$snp <- rownames(df1)
# 
# df1_long <- df1 %>%
#   pivot_longer(-snp, names_to = "module", values_to = "p_snp_geneset")
# 
# ## 图二：geneset_acat_pvals_M1.0001（列：module, p_acat）
# df2 <- geneset_acat_pvals_M1.0001 %>%
#   dplyr::select(module, p_fisher)
# 
# merged <- df1_long %>%
#   left_join(df2, by = "module") %>%
#   relocate(snp, module, p_snp_geneset, p_fisher)
# 
# 

X <- merged[,c(3,4)]

library(adaFilter)

result <- adaFilter(X, r = 2,alpha=0.1)

merged$decision <- result$decision
colnames(merged)[c(3,4)] <- c("snp_geneset","geneset_trait")

head(merged[merged$decision==1,])
```



```{r,fig.width=15,fig.height=15}

dfl2 <- merged[merged$decision==1,]
dfl2$gene_name <- dfl2$snp

```


```{r}
result <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/asthma/GCST010043/flames_200kb/result.rds")
symbol_to_rsid <- result %>%         # 你的 1251×7 表
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol, rsid) %>%
  distinct()

# ── 2. 把 rsid 加入 tbl_df ───────────────────────────────
tbl_with_rsid <- dfl2 %>%
  left_join(symbol_to_rsid, by = c("snp" = "symbol"))

# ── 3. 按 rsid 分组合并 ──────────────────────────────────
# 没有 rsid 匹配的行（rsid = NA）保持原样，snp 不变
merged_df <- tbl_with_rsid %>%
  group_by(rsid, module) %>%
  mutate(
    merged_snp = ifelse(
      all(is.na(rsid)),
      snp,
      paste0(
        paste(sort(unique(snp)), collapse = "-"),
        " (", rsid[!is.na(rsid)][1], ")"
      )
    )
  ) %>%
  ungroup() %>%
  group_by(merged_snp, module) %>%
  summarise(
    snp_geneset   = min(snp_geneset,   na.rm = TRUE),
    geneset_trait = min(geneset_trait, na.rm = TRUE),
    decision      = max(decision,      na.rm = TRUE),
    gene_name     = paste(sort(unique(gene_name)), collapse = "-"),
    .groups       = "drop"
  ) 

merged_df$gene_name <- merged_df$merged_snp
```


```{r}
head(merged_df)
```


```{r}
flames_symbols <- result %>%
  filter(source == "FLAMES") %>%
  pull(symbol) %>%
  unique() %>%
  na.omit()

# 检查 merged_df 的 snp 列中是否包含这些 symbol
# merged_snp 格式如 "RPS18-SLC39A7 (rs411919)"，需要拆解出 symbol 部分
highlight_genes <- merged_df %>%
  filter(
    sapply(merged_snp, function(s) {
      # 提取括号前的 symbol 部分，按 "-" 拆分
      sym_part <- sub("\\s*\\(.*\\)$", "", s)       # 去掉 " (rsXXX)"
      syms     <- strsplit(sym_part, "-")[[1]]       # 拆分多个 symbol
      any(syms %in% flames_symbols)                  # 任一 symbol 在 FLAMES 中
    })
  )

cat("FLAMES_gene 行数:", nrow(highlight_genes), "\n")
print(highlight_genes)
```




```{r,fig.width=14,fig.height=14}

plot_gene_module_network2(
  merged_df,
  top_n_gene = 200,
  size_scale = 1,
  highlight_genes = character(0),
    highlight_mod = character(0),

  layout_fun = igraph::layout_with_fr   # 关键：显式用 igraph 的
)
```

# Asthma, FLAMES + 200 kb combined locus mapping

In this section, GWAS loci are mapped using the union of FLAMES fine-mapping and a 200 kb genomic window. 3 values of K (10, 15 ,20 topics) are tested to assess robustness of the results to the number of latent topics. The downstream pipeline (GO enrichment → Fisher aggregation → AdaFilter → network visualization) is identical to the MCH analysis.

## K = 10 topics

### GO enrichment

```{r}
library(gprofiler2)
 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/topic_geneset/flames+200kb/de_10_0.01_2_0.5.rds")

 de_10_0.01_2_0.5 <-gene_modules_id

gene_sets <- lapply(de_10_0.01_2_0.5, function(x) unique(as.character(x)))

# gene_sets <- lapply(de_10_0.01_2_0.5 , function(x) unique(as.character(x)))

# 建议给每个 set 一个名字（没有就补上）
if (is.null(names(gene_sets))) names(gene_sets) <- paste0("set_", seq_along(gene_sets))

res <- gost(
  query = gene_sets,                 # list of 100 gene vectors
  organism = "hsapiens",
  multi_query = TRUE,
  correction_method = "fdr",
  user_threshold = 0.05,
  sources = c("GO:BP","GO:MF","KEGG")
)



# geneset 名字：如果你有 gene_sets 的 names 就用它；否则自动生成
gs_names <- names(gene_sets)
if (is.null(gs_names)) gs_names <- paste0("set_", seq_along(res$result$significant[[1]]))

# GO term 名字：用 term_id 或 term_name（建议 term_id 更稳定）
go_names <- paste0(res$result$source, ":", res$result$term_id)
# 如果你更想用可读名字：
# go_names <- paste0(res$result$source, ":", res$result$term_name)

# 1) list -> 矩阵：每列一个 term，每行一个 geneset
sig_term_by_set <- do.call(cbind, res$result$significant)
storage.mode(sig_term_by_set) <- "logical"

# 2) 转成“geneset(行) × term(列)”的 index matrix
idx_mat <- sig_term_by_set  # 现在行=geneset, 列=term（通常就是这个方向）
rownames(idx_mat) <- gs_names
colnames(idx_mat) <- go_names


# 如果你发现方向反了（极少数情况），就用这一句修正：
# idx_mat <- t(idx_mat)

library(dplyr)
library(tidyr)

go_tbl <- as_tibble(res$result)

# geneset 名称（如果你有 names(gene_sets) 就用它；否则自动生成）
G <- length(go_tbl$p_values[[1]])
gs_names <- names(gene_sets)
if (is.null(gs_names) || length(gs_names) != G) gs_names <- paste0("set_", seq_len(G))

# 展开：term × geneset 的长表
go_long <- go_tbl %>%
  mutate(
    query   = list(gs_names),
    p_value = p_values,
    sig     = significant       # significant 列也是 list，记录每个 geneset 是否显著
  ) %>%
  dplyr::select(source, term_id, term_name, query, p_value, sig) %>%
  unnest(c(query, p_value, sig)) %>%
  filter(sig == TRUE)  

# 每个 geneset 取 top10（p_value 最小）
top10_tbl <- go_long %>%
  group_by(query) %>%
  arrange(p_value, .by_group = TRUE) %>%
  ungroup()

# 你想要的：每个 geneset 的 top10 term_name（以及可选的 p_value）
top10_names <- top10_tbl %>%
  group_by(query) %>%
  summarise(top10_term_name = list(term_name), .groups = "drop")



# 注意！！！！！！！！！！！！！！！是按 1， 10 ，11 等排序
```

```{r}
top10_names
```

### Geneset–trait associations and AdaFilter mediation test (Asthma, K10, flames+200kb)

```{r}
library(data.table)
library(dplyr)
library(stringr)
library(purrr)
library(heavytailcombtest)

# -----------------------------
# 0) Read inputs
# -----------------------------
df <- fread("/project/xuanyao/jiaming/Getting_started/data/geneset_to_trait/asthma/asthma.tsv")
df2 <- df[grepl("M1.01", df$Name), ]

Nadig_associated_perturbation <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/asthma/GCST010043/flames+200kb/Nadig_associated_perturbation_flames_200kb.rds")



discovery_pairs_cis <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/discovery_pairs_cis.rds")


 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/topic_geneset/flames+200kb/de_10_0.01_2_0.5.rds")

all_measured_genes <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/all_measured_genes.rds")

# 只保留测到的 genes
gene_modules_id <- lapply(gene_modules_id, function(vec) intersect(vec, all_measured_genes))


# -----------------------------
# 1) Build pv_map from df2
# -----------------------------
gene_id <- str_extract(df2$Name, "(?<=\\()[^()]+(?=\\))")

lookup <- df2 %>%
  mutate(ID = gene_id) %>%
  filter(!is.na(ID), !is.na(p_value)) %>%
  group_by(ID) %>%
  summarise(p_value = min(p_value), .groups = "drop")

pv_map <- setNames(lookup$p_value, lookup$ID)


# -----------------------------
# 2) Helpers
# -----------------------------
clean_p <- function(p) {
  p <- as.numeric(p)
  p[is.na(p) | p <= 0] <- NA
  p
}

safe_comb <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  combination.test(
    p,
    weights = NA,
    method = "Cauchy",
    tail.idx = 1
  )
}

safe_comb_fisher <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  stat <- -2 * sum(log(p))
  stats::pchisq(stat, df = 2 * length(p), lower.tail = FALSE)
}


# -----------------------------
# 3) Make filtered modules per perturbation (and record n_removed)
# -----------------------------
make_filtered_modules <- function(gene_symbol, gene_modules_id, pairs_df) {
  target_name <- paste0("tar-",gene_symbol)

  ids_to_remove <- unique(pairs_df$response_id[pairs_df$grna_target == target_name])

  filtered_modules <- lapply(gene_modules_id, function(mod_ids) {
    setdiff(mod_ids, ids_to_remove)
  })

  list(
    modules   = filtered_modules,
    n_removed = length(ids_to_remove)
  )
}

filtered_by_perturbation <- setNames(
  lapply(
    Nadig_associated_perturbation,
    make_filtered_modules,
    gene_modules_id = gene_modules_id,
    pairs_df = discovery_pairs_cis
  ),
  Nadig_associated_perturbation
)


# -----------------------------
# 4) Compute module-level combined p-values per perturbation
#    (and add n_removed into module_results)
# -----------------------------
compute_for_one_pert <- function(pert) {
  obj <- filtered_by_perturbation[[pert]]
  gene_modules_id_pert <- obj$modules
  n_removed <- obj$n_removed

  # module -> p-values vector under this perturbation
  module_pvals <- lapply(gene_modules_id_pert, function(genes) {
    p <- pv_map[genes]
    names(p) <- genes
    p[!is.na(p)]
  })

  # combine within each module
  p_acat <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb(p))
  })

  p_fisher <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb_fisher(p))
  })

  tibble(
    snp       = pert,
    n_removed = n_removed,  # ✅ 每个 perturbation 的 ids_to_remove 数量（复制到每一行 module）
    module    = names(module_pvals),
    n_genes   = lengths(gene_modules_id_pert),
    n_matched = sapply(module_pvals, function(v) sum(!is.na(v))),
    p_acat    = p_acat,
    p_fisher  = p_fisher
  )
}

module_results <- map_dfr(Nadig_associated_perturbation, compute_for_one_pert)

# 查看结果
# module_results %>%
#   arrange(p_fisher) %>%
#   print(n = 30)

```


```{r}
library(tidyr)
library(dplyr)
library(tibble)

snp_geneset_p_value<-  readRDS("/project/xuanyao/jiaming/paper/output/snake/Nadig_asthma/flames+200kb/K10/3000_t_ACAT_resample/p_values/geneset_p_value.rds")
snp_geneset_p_value_test <- snp_geneset_p_value[which(!grepl("non-",rownames(snp_geneset_p_value))),]
snp_geneset_p_value_control <- snp_geneset_p_value[which(grepl("non-",rownames(snp_geneset_p_value))),]

df1_long <- snp_geneset_p_value_test %>%
  as.data.frame() %>%
  rownames_to_column("snp") %>%      # 这里真正把 rownames 变成一列 snp
  pivot_longer(
    -snp,
    names_to  = "module",
    values_to = "p_snp_geneset"
  )

## 2. 模块层面的表，只保留需要的列
df2 <- module_results %>%
  dplyr::select(module, p_fisher,snp)

## 3. 按 module 合并，再调整列顺序
merged <- df1_long %>%
  left_join(df2, by = c("snp", "module")) %>%  # 关键改动
  relocate(snp, module, p_snp_geneset, p_fisher)%>% tidyr::drop_na()


# library(dplyr)
# 
# df1 <- as.data.frame(snp_geneset_p_value)
# df1$snp <- rownames(df1)
# 
# df1_long <- df1 %>%
#   pivot_longer(-snp, names_to = "module", values_to = "p_snp_geneset")
# 
# ## 图二：geneset_acat_pvals_M1.0001（列：module, p_acat）
# df2 <- geneset_acat_pvals_M1.0001 %>%
#   dplyr::select(module, p_fisher)
# 
# merged <- df1_long %>%
#   left_join(df2, by = "module") %>%
#   relocate(snp, module, p_snp_geneset, p_fisher)
# 
# 

X <- merged[,c(3,4)]

library(adaFilter)

result <- adaFilter(X, r = 2,alpha=0.1)

merged$decision <- result$decision
colnames(merged)[c(3,4)] <- c("snp_geneset","geneset_trait")

head(merged[merged$decision==1,])
```



```{r,fig.width=15,fig.height=15}

dfl2 <- merged[merged$decision==1,]
dfl2$gene_name <- dfl2$snp
# plot_gene_module_network2(
#   dfl2,
#   top_n_gene = 300,
#   size_scale = 1,
#   highlight_genes = character(0),
#     highlight_mod = character(0),
#   gene_modules_id = gene_modules_id,   # <- 传入
# 
#   layout_fun = igraph::layout_with_fr   # 关键：显式用 igraph 的
# )


```


```{r}
result <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/asthma/GCST010043/flames+200kb/result.rds")
symbol_to_rsid <- result %>%         # 你的 1251×7 表
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol, rsid) %>%
  distinct()

# ── 2. 把 rsid 加入 tbl_df ───────────────────────────────
tbl_with_rsid <- dfl2 %>%
  left_join(symbol_to_rsid, by = c("snp" = "symbol"))

# ── 3. 按 rsid 分组合并 ──────────────────────────────────
# 没有 rsid 匹配的行（rsid = NA）保持原样，snp 不变
merged_df <- tbl_with_rsid %>%
  group_by(rsid, module) %>%
  mutate(
    merged_snp = ifelse(
      all(is.na(rsid)),
      snp,
      paste0(
        paste(sort(unique(snp)), collapse = "-"),
        " (", rsid[!is.na(rsid)][1], ")"
      )
    )
  ) %>%
  ungroup() %>%
  group_by(merged_snp, module) %>%
  summarise(
    snp_geneset   = min(snp_geneset,   na.rm = TRUE),
    geneset_trait = min(geneset_trait, na.rm = TRUE),
    decision      = max(decision,      na.rm = TRUE),
    gene_name     = paste(sort(unique(gene_name)), collapse = "-"),
    .groups       = "drop"
  ) 

merged_df$gene_name <- merged_df$merged_snp
```






```{r}
library(dplyr)
library(tidyr)

# 1. 从result提取rsid的位置信息
result_pos <- result %>%
  dplyr::select(rsid, locations) %>%
  separate(locations, into = c("chr", "pos"), sep = ":", convert = TRUE) %>%
  distinct(rsid, .keep_all = TRUE)

# 2. 从merged_snp提取rsid（括号内的rs号）
merged_df <- merged_df %>%
  mutate(snp_rsid = str_extract(merged_snp, "rs\\d+"))

# 3. 给merged_df加上位置信息
merged_with_pos <- merged_df %>%
  left_join(result_pos, by = c("snp_rsid" = "rsid"))

# 4. 计算每对merged_snp与result中所有rsid的距离，找最小值
# 先获取result所有位置
result_all_pos <- result_pos %>%
  rename(ref_chr = chr, ref_pos = pos)

distance_result <- merged_with_pos %>%
  rowwise() %>%
  mutate(
    min_distance = {
      current_chr <- chr
      current_pos <- pos
      current_rsid <- snp_rsid
      
      others <- merged_with_pos %>% 
        filter(chr == current_chr,
               snp_rsid != current_rsid)
      
      if (nrow(others) == 0) NA_real_
      else min(abs(others$pos - current_pos))
    },
    closest_snp = {
      current_chr <- chr
      current_pos <- pos
      current_rsid <- snp_rsid
      
      others <- merged_with_pos %>% 
        filter(chr == current_chr,
               snp_rsid != current_rsid)
      
      if (nrow(others) == 0) NA_character_
      else others$merged_snp[which.min(abs(others$pos - current_pos))]
    }
  ) %>%
  ungroup()

distance_result %>%
  dplyr::select(merged_snp, chr, pos, closest_snp, min_distance) %>%
  arrange(min_distance)
```



```{r}
flames_symbols <- result %>%
  filter(source == "FLAMES") %>%
  pull(symbol) %>%
  unique() %>%
  na.omit()

# 检查 merged_df 的 snp 列中是否包含这些 symbol
# merged_snp 格式如 "RPS18-SLC39A7 (rs411919)"，需要拆解出 symbol 部分
highlight_genes <- merged_df %>%
  filter(
    sapply(merged_snp, function(s) {
      # 提取括号前的 symbol 部分，按 "-" 拆分
      sym_part <- sub("\\s*\\(.*\\)$", "", s)       # 去掉 " (rsXXX)"
      syms     <- strsplit(sym_part, "-")[[1]]       # 拆分多个 symbol
      any(syms %in% flames_symbols)                  # 任一 symbol 在 FLAMES 中
    })
  )

cat("FLAMES_genes 行数:", nrow(highlight_genes), "\n")
print(highlight_genes)
```


### dandelion plot


```{r,fig.width=14,fig.height=14}

plot_gene_module_network2(
  merged_df,
  top_n_gene = 200,
  size_scale = 1,
  highlight_genes = character(0),
    highlight_mod = character(0),
    gene_modules_id = gene_modules_id,   # <- 传入


  layout_fun = igraph::layout_with_fr   # 关键：显式用 igraph 的
)
```



## K = 15 topics

```{r}
library(gprofiler2)
 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/topic_geneset/flames+200kb/de_15_0.01_2_0.5.rds")

 de_10_0.01_2_0.5 <-gene_modules_id

gene_sets <- lapply(de_10_0.01_2_0.5, function(x) unique(as.character(x)))

# gene_sets <- lapply(de_10_0.01_2_0.5 , function(x) unique(as.character(x)))

# 建议给每个 set 一个名字（没有就补上）
if (is.null(names(gene_sets))) names(gene_sets) <- paste0("set_", seq_along(gene_sets))

res <- gost(
  query = gene_sets,                 # list of 100 gene vectors
  organism = "hsapiens",
  multi_query = TRUE,
  correction_method = "fdr",
  user_threshold = 0.05,
  sources = c("GO:BP","GO:MF","KEGG")
)



# geneset 名字：如果你有 gene_sets 的 names 就用它；否则自动生成
gs_names <- names(gene_sets)
if (is.null(gs_names)) gs_names <- paste0("set_", seq_along(res$result$significant[[1]]))

# GO term 名字：用 term_id 或 term_name（建议 term_id 更稳定）
go_names <- paste0(res$result$source, ":", res$result$term_id)
# 如果你更想用可读名字：
# go_names <- paste0(res$result$source, ":", res$result$term_name)

# 1) list -> 矩阵：每列一个 term，每行一个 geneset
sig_term_by_set <- do.call(cbind, res$result$significant)
storage.mode(sig_term_by_set) <- "logical"

# 2) 转成“geneset(行) × term(列)”的 index matrix
idx_mat <- sig_term_by_set  # 现在行=geneset, 列=term（通常就是这个方向）
rownames(idx_mat) <- gs_names
colnames(idx_mat) <- go_names


# 如果你发现方向反了（极少数情况），就用这一句修正：
# idx_mat <- t(idx_mat)

library(dplyr)
library(tidyr)

go_tbl <- as_tibble(res$result)

# geneset 名称（如果你有 names(gene_sets) 就用它；否则自动生成）
G <- length(go_tbl$p_values[[1]])
gs_names <- names(gene_sets)
if (is.null(gs_names) || length(gs_names) != G) gs_names <- paste0("set_", seq_len(G))

# 展开：term × geneset 的长表
go_long <- go_tbl %>%
  mutate(
    query   = list(gs_names),
    p_value = p_values,
    sig     = significant       # significant 列也是 list，记录每个 geneset 是否显著
  ) %>%
  dplyr::select(source, term_id, term_name, query, p_value, sig) %>%
  unnest(c(query, p_value, sig)) %>%
  filter(sig == TRUE)  

# 每个 geneset 取 top10（p_value 最小）
top10_tbl <- go_long %>%
  group_by(query) %>%
  arrange(p_value, .by_group = TRUE) %>%
  ungroup()

# 你想要的：每个 geneset 的 top10 term_name（以及可选的 p_value）
top10_names <- top10_tbl %>%
  group_by(query) %>%
  summarise(top10_term_name = list(term_name), .groups = "drop")



# 注意！！！！！！！！！！！！！！！是按 1， 10 ，11 等排序
```



### Geneset–trait associations and AdaFilter mediation test (Asthma, K15, flames+200kb)




```{r}
library(data.table)
library(dplyr)
library(stringr)
library(purrr)
library(heavytailcombtest)

# -----------------------------
# 0) Read inputs
# -----------------------------
df <- fread("/project/xuanyao/jiaming/Getting_started/data/geneset_to_trait/asthma/asthma.tsv")
df2 <- df[grepl("M1.01", df$Name), ]

Nadig_associated_perturbation <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/asthma/GCST010043/flames+200kb/Nadig_associated_perturbation_flames_200kb.rds")



discovery_pairs_cis <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/discovery_pairs_cis.rds")


 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/topic_geneset/flames+200kb/de_15_0.01_2_0.5.rds")

all_measured_genes <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/all_measured_genes.rds")

# 只保留测到的 genes
gene_modules_id <- lapply(gene_modules_id, function(vec) intersect(vec, all_measured_genes))


# -----------------------------
# 1) Build pv_map from df2
# -----------------------------
gene_id <- str_extract(df2$Name, "(?<=\\()[^()]+(?=\\))")

lookup <- df2 %>%
  mutate(ID = gene_id) %>%
  filter(!is.na(ID), !is.na(p_value)) %>%
  group_by(ID) %>%
  summarise(p_value = min(p_value), .groups = "drop")

pv_map <- setNames(lookup$p_value, lookup$ID)


# -----------------------------
# 2) Helpers
# -----------------------------
clean_p <- function(p) {
  p <- as.numeric(p)
  p[is.na(p) | p <= 0] <- NA
  p
}

safe_comb <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  combination.test(
    p,
    weights = NA,
    method = "Cauchy",
    tail.idx = 1
  )
}

safe_comb_fisher <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  stat <- -2 * sum(log(p))
  stats::pchisq(stat, df = 2 * length(p), lower.tail = FALSE)
}


# -----------------------------
# 3) Make filtered modules per perturbation (and record n_removed)
# -----------------------------
make_filtered_modules <- function(gene_symbol, gene_modules_id, pairs_df) {
  target_name <- paste0("tar-",gene_symbol)

  ids_to_remove <- unique(pairs_df$response_id[pairs_df$grna_target == target_name])

  filtered_modules <- lapply(gene_modules_id, function(mod_ids) {
    setdiff(mod_ids, ids_to_remove)
  })

  list(
    modules   = filtered_modules,
    n_removed = length(ids_to_remove)
  )
}

filtered_by_perturbation <- setNames(
  lapply(
    Nadig_associated_perturbation,
    make_filtered_modules,
    gene_modules_id = gene_modules_id,
    pairs_df = discovery_pairs_cis
  ),
  Nadig_associated_perturbation
)


# -----------------------------
# 4) Compute module-level combined p-values per perturbation
#    (and add n_removed into module_results)
# -----------------------------
compute_for_one_pert <- function(pert) {
  obj <- filtered_by_perturbation[[pert]]
  gene_modules_id_pert <- obj$modules
  n_removed <- obj$n_removed

  # module -> p-values vector under this perturbation
  module_pvals <- lapply(gene_modules_id_pert, function(genes) {
    p <- pv_map[genes]
    names(p) <- genes
    p[!is.na(p)]
  })

  # combine within each module
  p_acat <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb(p))
  })

  p_fisher <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb_fisher(p))
  })

  tibble(
    snp       = pert,
    n_removed = n_removed,  # ✅ 每个 perturbation 的 ids_to_remove 数量（复制到每一行 module）
    module    = names(module_pvals),
    n_genes   = lengths(gene_modules_id_pert),
    n_matched = sapply(module_pvals, function(v) sum(!is.na(v))),
    p_acat    = p_acat,
    p_fisher  = p_fisher
  )
}

module_results <- map_dfr(Nadig_associated_perturbation, compute_for_one_pert)

# 查看结果
module_results %>%
  arrange(p_fisher) %>%
  print(n = 30)

```


```{r}
library(tidyr)
library(dplyr)
library(tibble)

snp_geneset_p_value<-  readRDS("/project/xuanyao/jiaming/paper/output/snake/Nadig_asthma/flames+200kb/K15/3000_t_ACAT_resample/p_values/geneset_p_value.rds")
snp_geneset_p_value_test <- snp_geneset_p_value[which(!grepl("non-",rownames(snp_geneset_p_value))),]
snp_geneset_p_value_control <- snp_geneset_p_value[which(grepl("non-",rownames(snp_geneset_p_value))),]

df1_long <- snp_geneset_p_value_test %>%
  as.data.frame() %>%
  rownames_to_column("snp") %>%      # 这里真正把 rownames 变成一列 snp
  pivot_longer(
    -snp,
    names_to  = "module",
    values_to = "p_snp_geneset"
  )

## 2. 模块层面的表，只保留需要的列
df2 <- module_results %>%
  dplyr::select(module, p_fisher,snp)

## 3. 按 module 合并，再调整列顺序
merged <- df1_long %>%
  left_join(df2, by = c("snp", "module")) %>%  # 关键改动
  relocate(snp, module, p_snp_geneset, p_fisher)%>% tidyr::drop_na()


# library(dplyr)
# 
# df1 <- as.data.frame(snp_geneset_p_value)
# df1$snp <- rownames(df1)
# 
# df1_long <- df1 %>%
#   pivot_longer(-snp, names_to = "module", values_to = "p_snp_geneset")
# 
# ## 图二：geneset_acat_pvals_M1.0001（列：module, p_acat）
# df2 <- geneset_acat_pvals_M1.0001 %>%
#   dplyr::select(module, p_fisher)
# 
# merged <- df1_long %>%
#   left_join(df2, by = "module") %>%
#   relocate(snp, module, p_snp_geneset, p_fisher)
# 
# 

X <- merged[,c(3,4)]

library(adaFilter)

result <- adaFilter(X, r = 2,alpha=0.1)

merged$decision <- result$decision
colnames(merged)[c(3,4)] <- c("snp_geneset","geneset_trait")

merged[merged$decision==1,]
```



```{r,fig.width=15,fig.height=15}

dfl2 <- merged[merged$decision==1,]
dfl2$gene_name <- dfl2$snp
plot_gene_module_network2(
  dfl2,
  top_n_gene = 200,
  size_scale = 1,
  highlight_genes = character(0),
    highlight_mod = character(0),
  gene_modules_id = gene_modules_id,   # <- 传入

  layout_fun = igraph::layout_with_fr   # 关键：显式用 igraph 的
)
```
```{r}
result <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/asthma/GCST010043/flames+200kb/result.rds")
symbol_to_rsid <- result %>%         # 你的 1251×7 表
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol, rsid) %>%
  distinct()

# ── 2. 把 rsid 加入 tbl_df ───────────────────────────────
tbl_with_rsid <- dfl2 %>%
  left_join(symbol_to_rsid, by = c("snp" = "symbol"))

# ── 3. 按 rsid 分组合并 ──────────────────────────────────
# 没有 rsid 匹配的行（rsid = NA）保持原样，snp 不变
merged_df <- tbl_with_rsid %>%
  group_by(rsid, module) %>%
  mutate(
    merged_snp = ifelse(
      all(is.na(rsid)),
      snp,
      paste0(
        paste(sort(unique(snp)), collapse = "-"),
        " (", rsid[!is.na(rsid)][1], ")"
      )
    )
  ) %>%
  ungroup() %>%
  group_by(merged_snp, module) %>%
  summarise(
    snp_geneset   = min(snp_geneset,   na.rm = TRUE),
    geneset_trait = min(geneset_trait, na.rm = TRUE),
    decision      = max(decision,      na.rm = TRUE),
    gene_name     = paste(sort(unique(gene_name)), collapse = "-"),
    .groups       = "drop"
  ) 

merged_df$gene_name <- merged_df$merged_snp
```


```{r}
merged_df
```
```{r}
library(dplyr)
library(tidyr)

# 1. 从result提取rsid的位置信息
result_pos <- result %>%
  dplyr::select(rsid, locations) %>%
  separate(locations, into = c("chr", "pos"), sep = ":", convert = TRUE) %>%
  distinct(rsid, .keep_all = TRUE)

# 2. 从merged_snp提取rsid（括号内的rs号）
merged_df <- merged_df %>%
  mutate(snp_rsid = str_extract(merged_snp, "rs\\d+"))

# 3. 给merged_df加上位置信息
merged_with_pos <- merged_df %>%
  left_join(result_pos, by = c("snp_rsid" = "rsid"))

# 4. 计算每对merged_snp与result中所有rsid的距离，找最小值
# 先获取result所有位置
result_all_pos <- result_pos %>%
  rename(ref_chr = chr, ref_pos = pos)

distance_result <- merged_with_pos %>%
  rowwise() %>%
  mutate(
    min_distance = {
      current_chr <- chr
      current_pos <- pos
      current_rsid <- snp_rsid
      
      others <- merged_with_pos %>% 
        filter(chr == current_chr,
               snp_rsid != current_rsid)
      
      if (nrow(others) == 0) NA_real_
      else min(abs(others$pos - current_pos))
    },
    closest_snp = {
      current_chr <- chr
      current_pos <- pos
      current_rsid <- snp_rsid
      
      others <- merged_with_pos %>% 
        filter(chr == current_chr,
               snp_rsid != current_rsid)
      
      if (nrow(others) == 0) NA_character_
      else others$merged_snp[which.min(abs(others$pos - current_pos))]
    }
  ) %>%
  ungroup()

distance_result %>%
  dplyr::select(merged_snp, chr, pos, closest_snp, min_distance) %>%
  arrange(min_distance)
```


```{r}
flames_symbols <- result %>%
  filter(source == "FLAMES") %>%
  pull(symbol) %>%
  unique() %>%
  na.omit()

# 检查 merged_df 的 snp 列中是否包含这些 symbol
# merged_snp 格式如 "RPS18-SLC39A7 (rs411919)"，需要拆解出 symbol 部分
highlight_genes <- merged_df %>%
  filter(
    sapply(merged_snp, function(s) {
      # 提取括号前的 symbol 部分，按 "-" 拆分
      sym_part <- sub("\\s*\\(.*\\)$", "", s)       # 去掉 " (rsXXX)"
      syms     <- strsplit(sym_part, "-")[[1]]       # 拆分多个 symbol
      any(syms %in% flames_symbols)                  # 任一 symbol 在 FLAMES 中
    })
  )

cat("FLAMES_genes 行数:", nrow(highlight_genes), "\n")
print(highlight_genes)
```

### dandelion plot



```{r,fig.width=14,fig.height=14}

plot_gene_module_network2(
  merged_df,
  top_n_gene = 200,
  size_scale = 1,
  highlight_genes = character(0),
    highlight_mod = character(0),
  gene_modules_id = gene_modules_id,   # <- 传入

  layout_fun = igraph::layout_with_fr   # 关键：显式用 igraph 的
)
```


## K=20

### GO enrichment

```{r}
library(gprofiler2)
 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/topic_geneset/flames+200kb/de_20_0.01_2_0.5.rds")

 de_10_0.01_2_0.5 <-gene_modules_id

gene_sets <- lapply(de_10_0.01_2_0.5, function(x) unique(as.character(x)))

# gene_sets <- lapply(de_10_0.01_2_0.5 , function(x) unique(as.character(x)))

# 建议给每个 set 一个名字（没有就补上）
if (is.null(names(gene_sets))) names(gene_sets) <- paste0("set_", seq_along(gene_sets))

res <- gost(
  query = gene_sets,                 # list of 100 gene vectors
  organism = "hsapiens",
  multi_query = TRUE,
  correction_method = "fdr",
  user_threshold = 0.05,
  sources = c("GO:BP","GO:MF","KEGG")
)



# geneset 名字：如果你有 gene_sets 的 names 就用它；否则自动生成
gs_names <- names(gene_sets)
if (is.null(gs_names)) gs_names <- paste0("set_", seq_along(res$result$significant[[1]]))

# GO term 名字：用 term_id 或 term_name（建议 term_id 更稳定）
go_names <- paste0(res$result$source, ":", res$result$term_id)
# 如果你更想用可读名字：
# go_names <- paste0(res$result$source, ":", res$result$term_name)

# 1) list -> 矩阵：每列一个 term，每行一个 geneset
sig_term_by_set <- do.call(cbind, res$result$significant)
storage.mode(sig_term_by_set) <- "logical"

# 2) 转成“geneset(行) × term(列)”的 index matrix
idx_mat <- sig_term_by_set  # 现在行=geneset, 列=term（通常就是这个方向）
rownames(idx_mat) <- gs_names
colnames(idx_mat) <- go_names


# 如果你发现方向反了（极少数情况），就用这一句修正：
# idx_mat <- t(idx_mat)

library(dplyr)
library(tidyr)

go_tbl <- as_tibble(res$result)

# geneset 名称（如果你有 names(gene_sets) 就用它；否则自动生成）
G <- length(go_tbl$p_values[[1]])
gs_names <- names(gene_sets)
if (is.null(gs_names) || length(gs_names) != G) gs_names <- paste0("set_", seq_len(G))

# 展开：term × geneset 的长表
go_long <- go_tbl %>%
  mutate(
    query   = list(gs_names),
    p_value = p_values,
    sig     = significant       # significant 列也是 list，记录每个 geneset 是否显著
  ) %>%
  dplyr::select(source, term_id, term_name, query, p_value, sig) %>%
  unnest(c(query, p_value, sig)) %>%
  filter(sig == TRUE)  

# 每个 geneset 取 top10（p_value 最小）
top10_tbl <- go_long %>%
  group_by(query) %>%
  arrange(p_value, .by_group = TRUE) %>%
  ungroup()

# 你想要的：每个 geneset 的 top10 term_name（以及可选的 p_value）
top10_names <- top10_tbl %>%
  group_by(query) %>%
  summarise(top10_term_name = list(term_name), .groups = "drop")



# 注意！！！！！！！！！！！！！！！是按 1， 10 ，11 等排序
```

```{r}
top10_names
```



### Geneset–trait associations and AdaFilter mediation test (Asthma, K20, flames+200kb)


```{r}
library(data.table)
library(dplyr)
library(stringr)
library(purrr)
library(heavytailcombtest)

# -----------------------------
# 0) Read inputs
# -----------------------------
df <- fread("/project/xuanyao/jiaming/Getting_started/data/geneset_to_trait/asthma/asthma.tsv")
df2 <- df[grepl("M1.01", df$Name), ]

Nadig_associated_perturbation <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/asthma/GCST010043/flames+200kb/Nadig_associated_perturbation_flames_200kb.rds")



discovery_pairs_cis <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/discovery_pairs_cis.rds")


 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/topic_geneset/flames+200kb/de_20_0.01_2_0.5.rds")

all_measured_genes <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Nadig/all_measured_genes.rds")

# 只保留测到的 genes
gene_modules_id <- lapply(gene_modules_id, function(vec) intersect(vec, all_measured_genes))


# -----------------------------
# 1) Build pv_map from df2
# -----------------------------
gene_id <- str_extract(df2$Name, "(?<=\\()[^()]+(?=\\))")

lookup <- df2 %>%
  mutate(ID = gene_id) %>%
  filter(!is.na(ID), !is.na(p_value)) %>%
  group_by(ID) %>%
  summarise(p_value = min(p_value), .groups = "drop")

pv_map <- setNames(lookup$p_value, lookup$ID)


# -----------------------------
# 2) Helpers
# -----------------------------
clean_p <- function(p) {
  p <- as.numeric(p)
  p[is.na(p) | p <= 0] <- NA
  p
}

safe_comb <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  combination.test(
    p,
    weights = NA,
    method = "Cauchy",
    tail.idx = 1
  )
}

safe_comb_fisher <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  stat <- -2 * sum(log(p))
  stats::pchisq(stat, df = 2 * length(p), lower.tail = FALSE)
}


# -----------------------------
# 3) Make filtered modules per perturbation (and record n_removed)
# -----------------------------
make_filtered_modules <- function(gene_symbol, gene_modules_id, pairs_df) {
  target_name <- paste0("tar-",gene_symbol)

  ids_to_remove <- unique(pairs_df$response_id[pairs_df$grna_target == target_name])

  filtered_modules <- lapply(gene_modules_id, function(mod_ids) {
    setdiff(mod_ids, ids_to_remove)
  })

  list(
    modules   = filtered_modules,
    n_removed = length(ids_to_remove)
  )
}

filtered_by_perturbation <- setNames(
  lapply(
    Nadig_associated_perturbation,
    make_filtered_modules,
    gene_modules_id = gene_modules_id,
    pairs_df = discovery_pairs_cis
  ),
  Nadig_associated_perturbation
)


# -----------------------------
# 4) Compute module-level combined p-values per perturbation
#    (and add n_removed into module_results)
# -----------------------------
compute_for_one_pert <- function(pert) {
  obj <- filtered_by_perturbation[[pert]]
  gene_modules_id_pert <- obj$modules
  n_removed <- obj$n_removed

  # module -> p-values vector under this perturbation
  module_pvals <- lapply(gene_modules_id_pert, function(genes) {
    p <- pv_map[genes]
    names(p) <- genes
    p[!is.na(p)]
  })

  # combine within each module
  p_acat <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb(p))
  })

  p_fisher <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb_fisher(p))
  })

  tibble(
    snp       = pert,
    n_removed = n_removed,  # ✅ 每个 perturbation 的 ids_to_remove 数量（复制到每一行 module）
    module    = names(module_pvals),
    n_genes   = lengths(gene_modules_id_pert),
    n_matched = sapply(module_pvals, function(v) sum(!is.na(v))),
    p_acat    = p_acat,
    p_fisher  = p_fisher
  )
}

module_results <- map_dfr(Nadig_associated_perturbation, compute_for_one_pert)

# 查看结果
# module_results %>%
#   arrange(p_fisher) %>%
#   print(n = 30)

```


```{r}
library(tidyr)
library(dplyr)
library(tibble)

snp_geneset_p_value<-  readRDS("/project/xuanyao/jiaming/paper/output/snake/Nadig_asthma/flames+200kb/K20/3000_t_ACAT_resample/p_values/geneset_p_value.rds")
snp_geneset_p_value_test <- snp_geneset_p_value[which(!grepl("non-",rownames(snp_geneset_p_value))),]
snp_geneset_p_value_control <- snp_geneset_p_value[which(grepl("non-",rownames(snp_geneset_p_value))),]

df1_long <- snp_geneset_p_value_test %>%
  as.data.frame() %>%
  rownames_to_column("snp") %>%      # 这里真正把 rownames 变成一列 snp
  pivot_longer(
    -snp,
    names_to  = "module",
    values_to = "p_snp_geneset"
  )

## 2. 模块层面的表，只保留需要的列
df2 <- module_results %>%
  dplyr::select(module, p_fisher,snp)

## 3. 按 module 合并，再调整列顺序
merged <- df1_long %>%
  left_join(df2, by = c("snp", "module")) %>%  # 关键改动
  relocate(snp, module, p_snp_geneset, p_fisher)%>% tidyr::drop_na()


# library(dplyr)
# 
# df1 <- as.data.frame(snp_geneset_p_value)
# df1$snp <- rownames(df1)
# 
# df1_long <- df1 %>%
#   pivot_longer(-snp, names_to = "module", values_to = "p_snp_geneset")
# 
# ## 图二：geneset_acat_pvals_M1.0001（列：module, p_acat）
# df2 <- geneset_acat_pvals_M1.0001 %>%
#   dplyr::select(module, p_fisher)
# 
# merged <- df1_long %>%
#   left_join(df2, by = "module") %>%
#   relocate(snp, module, p_snp_geneset, p_fisher)
# 
# 

X <- merged[,c(3,4)]

library(adaFilter)

result <- adaFilter(X, r = 2,alpha=0.1)

merged$decision <- result$decision
colnames(merged)[c(3,4)] <- c("snp_geneset","geneset_trait")

head(merged[merged$decision==1,])
```


```{r,fig.width=15,fig.height=15}

dfl2 <- merged[merged$decision==1,]
dfl2$gene_name <- dfl2$snp

```
```{r}
result <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/asthma/GCST010043/flames+200kb/result.rds")
symbol_to_rsid <- result %>%         # 你的 1251×7 表
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol, rsid) %>%
  distinct()

# ── 2. 把 rsid 加入 tbl_df ───────────────────────────────
tbl_with_rsid <- dfl2 %>%
  left_join(symbol_to_rsid, by = c("snp" = "symbol"))

# ── 3. 按 rsid 分组合并 ──────────────────────────────────
# 没有 rsid 匹配的行（rsid = NA）保持原样，snp 不变
merged_df <- tbl_with_rsid %>%
  group_by(rsid, module) %>%
  mutate(
    merged_snp = ifelse(
      all(is.na(rsid)),
      snp,
      paste0(
        paste(sort(unique(snp)), collapse = "-"),
        " (", rsid[!is.na(rsid)][1], ")"
      )
    )
  ) %>%
  ungroup() %>%
  group_by(merged_snp, module) %>%
  summarise(
    snp_geneset   = min(snp_geneset,   na.rm = TRUE),
    geneset_trait = min(geneset_trait, na.rm = TRUE),
    decision      = max(decision,      na.rm = TRUE),
    gene_name     = paste(sort(unique(gene_name)), collapse = "-"),
    .groups       = "drop"
  ) 

merged_df$gene_name <- merged_df$merged_snp
```


```{r}
head(merged_df)
```


```{r}
flames_symbols <- result %>%
  filter(source == "FLAMES") %>%
  pull(symbol) %>%
  unique() %>%
  na.omit()

# 检查 merged_df 的 snp 列中是否包含这些 symbol
# merged_snp 格式如 "RPS18-SLC39A7 (rs411919)"，需要拆解出 symbol 部分
highlight_genes <- merged_df %>%
  filter(
    sapply(merged_snp, function(s) {
      # 提取括号前的 symbol 部分，按 "-" 拆分
      sym_part <- sub("\\s*\\(.*\\)$", "", s)       # 去掉 " (rsXXX)"
      syms     <- strsplit(sym_part, "-")[[1]]       # 拆分多个 symbol
      any(syms %in% flames_symbols)                  # 任一 symbol 在 FLAMES 中
    })
  )

cat("FLAMES_genes 行数:", nrow(highlight_genes), "\n")
print(highlight_genes)
```




```{r,fig.width=14,fig.height=14}

plot_gene_module_network2(
  merged_df,
  top_n_gene = 200,
  size_scale = 1,
  highlight_genes = character(0),
    highlight_mod = character(0),
  gene_modules_id = gene_modules_id,   # <- 传入

  layout_fun = igraph::layout_with_fr   # 关键：显式用 igraph 的
)
```

# Asthma_CDT4

## GO enrichment

```{r}
library(gprofiler2)
 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/CDT4/topic_geneset/flames+200kb/de_50_0.01_2_0.5.rds")

 de_10_0.01_2_0.5 <-gene_modules_id

gene_sets <- lapply(de_10_0.01_2_0.5, function(x) unique(as.character(x)))

# gene_sets <- lapply(de_10_0.01_2_0.5 , function(x) unique(as.character(x)))

# 建议给每个 set 一个名字（没有就补上）
if (is.null(names(gene_sets))) names(gene_sets) <- paste0("set_", seq_along(gene_sets))

res <- gost(
  query = gene_sets,                 # list of 100 gene vectors
  organism = "hsapiens",
  multi_query = TRUE,
  correction_method = "fdr",
  user_threshold = 0.05,
  sources = c("GO:BP","GO:MF","KEGG")
)



# geneset 名字：如果你有 gene_sets 的 names 就用它；否则自动生成
gs_names <- names(gene_sets)
if (is.null(gs_names)) gs_names <- paste0("set_", seq_along(res$result$significant[[1]]))

# GO term 名字：用 term_id 或 term_name（建议 term_id 更稳定）
go_names <- paste0(res$result$source, ":", res$result$term_id)
# 如果你更想用可读名字：
# go_names <- paste0(res$result$source, ":", res$result$term_name)

# 1) list -> 矩阵：每列一个 term，每行一个 geneset
sig_term_by_set <- do.call(cbind, res$result$significant)
storage.mode(sig_term_by_set) <- "logical"

# 2) 转成“geneset(行) × term(列)”的 index matrix
idx_mat <- sig_term_by_set  # 现在行=geneset, 列=term（通常就是这个方向）
rownames(idx_mat) <- gs_names
colnames(idx_mat) <- go_names


# 如果你发现方向反了（极少数情况），就用这一句修正：
# idx_mat <- t(idx_mat)

library(dplyr)
library(tidyr)

go_tbl <- as_tibble(res$result)

# geneset 名称（如果你有 names(gene_sets) 就用它；否则自动生成）
G <- length(go_tbl$p_values[[1]])
gs_names <- names(gene_sets)
if (is.null(gs_names) || length(gs_names) != G) gs_names <- paste0("set_", seq_len(G))

# 展开：term × geneset 的长表
go_long <- go_tbl %>%
  mutate(
    query   = list(gs_names),
    p_value = p_values,
    sig     = significant       # significant 列也是 list，记录每个 geneset 是否显著
  ) %>%
  dplyr::select(source, term_id, term_name, query, p_value, sig) %>%
  unnest(c(query, p_value, sig)) %>%
  filter(sig == TRUE)  

# 每个 geneset 取 top10（p_value 最小）
top10_tbl <- go_long %>%
  group_by(query) %>%
  arrange(p_value, .by_group = TRUE) %>%
  ungroup()

# 你想要的：每个 geneset 的 top10 term_name（以及可选的 p_value）
top10_names <- top10_tbl %>%
  group_by(query) %>%
  summarise(top10_term_name = list(term_name), .groups = "drop")



# 注意！！！！！！！！！！！！！！！是按 1， 10 ，11 等排序
```

```{r,eval=F}
library(openxlsx)
wb <- createWorkbook()
for (q in sort(unique(top10_tbl$query))) {
  df_sheet <- top10_tbl %>%
    filter(query == q) %>%
    dplyr::select(term_name, p_value) %>%
    as.data.frame()
  addWorksheet(wb, q)
  writeData(wb, q, df_sheet)
}
saveWorkbook(wb, "/project/xuanyao/jiaming/paper/GO_terms_pvalue_athma_CD4_k50.xlsx", overwrite = TRUE)
```

```{r,eval=F}
## K20
top10_names[11,2][[1]]
```
```{r,eval=F}
top10_names[37,2][[1]]
```

## Geneset–trait associations and AdaFilter mediation test (Asthma, K50, flames+200kb)

```{r}
library(data.table)
library(dplyr)
library(stringr)
library(purrr)
library(heavytailcombtest)

# -----------------------------
# 0) Read inputs
# -----------------------------
df <- fread("/project/xuanyao/jiaming/Getting_started/data/geneset_to_trait/asthma/asthma.tsv")
df2 <- df[grepl("M1.01", df$Name), ]

Nadig_associated_perturbation <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/asthma/GCST010043/flames+200kb/CDT4_associated_perturbation.rds")



discovery_pairs_cis <- readRDS("/project/xuanyao/jiaming/Getting_started/data/CDT4/standard_input_format_asthma/flames+200kb/discovery_pairs_cis.rds")


 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/CDT4/topic_geneset/flames+200kb/de_50_0.01_2_0.5.rds")

# gene_modules_id<-filtered_list_2


# -----------------------------
# 1) Build pv_map from df2
# -----------------------------
gene_id <- str_extract(df2$Name, "^[^(]+")

lookup <- df2 %>%
  mutate(ID = gene_id) %>%
  filter(!is.na(ID), !is.na(p_value)) %>%
  group_by(ID) %>%
  summarise(p_value = min(p_value), .groups = "drop")

pv_map <- setNames(lookup$p_value, lookup$ID)


# -----------------------------
# 2) Helpers
# -----------------------------
clean_p <- function(p) {
  p <- as.numeric(p)
  p[is.na(p) | p <= 0] <- NA
  p
}

safe_comb <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  combination.test(
    p,
    weights = NA,
    method = "Cauchy",
    tail.idx = 1
  )
}

safe_comb_fisher <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  stat <- -2 * sum(log(p))
  stats::pchisq(stat, df = 2 * length(p), lower.tail = FALSE)
}


# -----------------------------
# 3) Make filtered modules per perturbation (and record n_removed)
# -----------------------------
make_filtered_modules <- function(gene_symbol, gene_modules_id, pairs_df) {
  target_name <- gene_symbol

  ids_to_remove <- unique(pairs_df$response_id[pairs_df$grna_target == target_name])

  filtered_modules <- lapply(gene_modules_id, function(mod_ids) {
    setdiff(mod_ids, ids_to_remove)
  })

  list(
    modules   = filtered_modules,
    n_removed = length(ids_to_remove)
  )
}

filtered_by_perturbation <- setNames(
  lapply(
    Nadig_associated_perturbation,
    make_filtered_modules,
    gene_modules_id = gene_modules_id,
    pairs_df = discovery_pairs_cis
  ),
  Nadig_associated_perturbation
)


# -----------------------------
# 4) Compute module-level combined p-values per perturbation
#    (and add n_removed into module_results)
# -----------------------------
compute_for_one_pert <- function(pert) {
  obj <- filtered_by_perturbation[[pert]]
  gene_modules_id_pert <- obj$modules
  n_removed <- obj$n_removed

  # module -> p-values vector under this perturbation
  module_pvals <- lapply(gene_modules_id_pert, function(genes) {
    p <- pv_map[genes]
    names(p) <- genes
    p[!is.na(p)]
  })

  # combine within each module
  p_acat <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb(p))
  })

  p_fisher <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb_fisher(p))
  })

  tibble(
    snp       = pert,
    n_removed = n_removed,  # ✅ 每个 perturbation 的 ids_to_remove 数量（复制到每一行 module）
    module    = names(module_pvals),
    n_genes   = lengths(gene_modules_id_pert),
    n_matched = sapply(module_pvals, function(v) sum(!is.na(v))),
    p_acat    = p_acat,
    p_fisher  = p_fisher
  )
}

module_results <- map_dfr(Nadig_associated_perturbation, compute_for_one_pert)

# 查看结果
module_results %>%
  arrange(p_fisher) %>%
  print(n = 10)

```


```{r}
library(tidyr)
library(dplyr)
library(tibble)

snp_geneset_p_value<-  readRDS("/project/xuanyao/jiaming/paper/output/snake/CDT4_asthma/flames+200kb/K50_optimized/3000_t_ACAT_resample/p_values/geneset_p_value.rds")
snp_geneset_p_value_test <- snp_geneset_p_value[which(!grepl("NTC-",rownames(snp_geneset_p_value))),]
snp_geneset_p_value_control <- snp_geneset_p_value[which(grepl("NTC-",rownames(snp_geneset_p_value))),]

df1_long <- snp_geneset_p_value_test %>%
  as.data.frame() %>%
  rownames_to_column("snp") %>%      # 这里真正把 rownames 变成一列 snp
  pivot_longer(
    -snp,
    names_to  = "module",
    values_to = "p_snp_geneset"
  )

## 2. 模块层面的表，只保留需要的列
df2 <- module_results %>%
  dplyr::select(module, p_fisher,snp)

## 3. 按 module 合并，再调整列顺序
merged <- df1_long %>%
  left_join(df2, by = c("snp", "module")) %>%  # 关键改动
  relocate(snp, module, p_snp_geneset, p_fisher)%>% tidyr::drop_na()


# library(dplyr)
# 
# df1 <- as.data.frame(snp_geneset_p_value)
# df1$snp <- rownames(df1)
# 
# df1_long <- df1 %>%
#   pivot_longer(-snp, names_to = "module", values_to = "p_snp_geneset")
# 
# ## 图二：geneset_acat_pvals_M1.0001（列：module, p_acat）
# df2 <- geneset_acat_pvals_M1.0001 %>%
#   dplyr::select(module, p_fisher)
# 
# merged <- df1_long %>%
#   left_join(df2, by = "module") %>%
#   relocate(snp, module, p_snp_geneset, p_fisher)
# 
# 

library(adaFilter)

module_sizes <- lengths(gene_modules_id)
merged <- merged %>% filter(module_sizes[module] > 1)

X <- merged[,c(3,4)]
result <- adaFilter(X, r = 2,alpha=0.15)

merged$decision <- result$decision
colnames(merged)[c(3,4)] <- c("snp_geneset","geneset_trait")

dfl2<-merged[merged$decision==1,]
```


The chunk below maps significant peripheral gene IDs (ENSG format from OPERA row names) back to gene symbols, and clusters them by genomic locus (200 kb window) so that co-located genes at the same GWAS locus are grouped together in downstream visualizations.

```{r}
# 0. ENSG -> symbol mapping from GTF
gtf <- rtracklayer::import("/project/xuanyao/jiaming/Getting_started/data/protein_coding_genes/Homo_sapiens.GRCh38.112.gtf.gz",
                            feature.type = "gene")
gtf_df <- as.data.frame(gtf)[, c("gene_id", "gene_name")]
colnames(gtf_df) <- c("ensg", "symbol")
ensg_to_symbol <- setNames(gtf_df$symbol, gtf_df$ensg)

# 1. Build rsid -> cluster_id: merge GWAS loci within 200kb on the same chromosome
gwas_result <- readRDS("/project/xuanyao/jiaming/Getting_started/data/GWAS/asthma/GCST010043/flames+200kb/result.rds")

rsid_locations <- gwas_result %>%
  dplyr::select(rsid, locations) %>%
  distinct() %>%
  filter(!is.na(locations)) %>%
  mutate(
    chr = sub(":.*", "", locations),
    # suppressWarnings: non-autosomal (X/Y/MT) or malformed entries produce NA,
    # which are removed by the chr filter below — this is expected
    pos = suppressWarnings(as.numeric(sub(".*:", "", locations)))
  ) %>%
  filter(chr %in% as.character(1:22)) %>%
  arrange(chr, pos) %>%
  group_by(chr) %>%
  mutate(cluster_num = cumsum(c(1, diff(pos) > 200000))) %>%
  ungroup() %>%
  mutate(cluster_id = paste0(chr, "_", cluster_num)) %>%
  dplyr::select(rsid, chr, pos, cluster_id)

# 2. symbol -> rsid mapping from gwas_result (bridge: ENSG -> symbol -> rsid)
symbol_to_rsid <- gwas_result %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol, rsid) %>%
  distinct()

tbl <- dfl2 %>%
  mutate(symbol = snp) %>%
  # many-to-many is expected: one gene can be near multiple GWAS loci (multiple rsids),
  # and the same gene can appear in multiple modules. Rows are collapsed in step 4
  # by group_by(module, cluster_id).
  left_join(symbol_to_rsid, by = "symbol", relationship = "many-to-many") %>%
  left_join(rsid_locations %>% dplyr::select(rsid, cluster_id), by = "rsid") %>%
  mutate(cluster_id = ifelse(is.na(cluster_id), paste0("solo_", ensg), cluster_id))

# 4. Within each module, find connected components of a gene–cluster bipartite graph.
#    Two genes (or two clusters) end up in the same component if they are bridged by a
#    shared gene or a shared cluster anywhere in the chain — enabling transitive merging.
#    e.g. gene1↔cluster_A, gene1↔cluster_B  →  gene1 + all genes in cluster_B form one row.
library(igraph)

merged_df <- tbl %>%
  group_by(module) %>%
  group_modify(function(df, key) {
    # Build bipartite edges: gene node  <->  cluster node
    # Use ENSG as fallback node label when symbol is missing
    edges <- df %>%
      transmute(
        gene_node    = paste0("g:", ifelse(!is.na(symbol), symbol, ensg)),
        cluster_node = paste0("c:", cluster_id)
      ) %>%
      distinct()

    g    <- igraph::graph_from_data_frame(edges, directed = FALSE)
    comp <- igraph::components(g)$membership          # named vector: node -> comp_id

    # Map each row back to its component via the cluster node
    node_comp <- tibble(node = names(comp), comp_id = as.integer(comp))

    df %>%
      left_join(
        node_comp %>%
          filter(startsWith(node, "c:")) %>%
          transmute(cluster_id = sub("^c:", "", node), comp_id),
        by = "cluster_id"
      ) %>%
      group_by(comp_id) %>%
      summarise(
        snp_geneset      = min(snp_geneset,   na.rm = TRUE),
        geneset_trait    = min(geneset_trait, na.rm = TRUE),
        merged_gene_name = paste(sort(unique(symbol[!is.na(symbol)])), collapse = "-"),
        merged_loci      = paste(sort(unique(rsid[!is.na(rsid)])),     collapse = "-"),
        .groups = "drop"
      ) %>%
      dplyr::select(-comp_id)
  }) %>%
  ungroup() %>%
  mutate(
    summary = paste0(merged_gene_name, "(", merged_loci, ")")
  ) %>%
  dplyr::select(module, merged_gene_name, merged_loci, summary)

head(merged_df)
```


## Gene–module network visualization

```{r plot-asthma-network-fn, include=FALSE}
# ── Simplified network plot (no biological colour coding) ────────────────────
plot_gene_module_network_asthma <- function(df12,
                                            top_n_per_module = NULL,
                                            top_n_gene       = 600,
                                            pcap             = 1e-300,
                                            layout_fun       = NULL,
                                            out_pdf          = NULL,
                                            width            = 12, height = 9,
                                            label_genes      = TRUE,
                                            label_modules    = TRUE,
                                            size_scale       = 0.7,
                                            mod_size_range   = c(15, 40),
                                            gene_size_range  = c(4,  20)) {

  if (is.null(layout_fun)) layout_fun <- igraph::layout_with_fr

  node_gene_col   <- "#AED6F1"   # light blue  – peripheral gene/locus nodes
  node_module_col <- "#D5D8DC"   # light grey  – module nodes

  # ── data ─────────────────────────────────────────────────────────────────────
  d <- df12 %>%
    dplyr::transmute(
      module = as.character(module),
      gene   = as.character(merged_gene_name),
      p_snp  = pcap,
      p_mod  = pcap
    ) %>%
    dplyr::filter(!is.na(module), !is.na(gene)) %>%
    dplyr::group_by(module, gene) %>%
    dplyr::summarise(p_snp = min(p_snp, na.rm = TRUE),
                     p_mod = min(p_mod, na.rm = TRUE),
                     .groups = "drop")

  if (!is.null(top_n_per_module)) {
    d <- d %>%
      dplyr::group_by(module) %>%
      dplyr::arrange(p_snp, .by_group = TRUE) %>%
      dplyr::slice_head(n = top_n_per_module) %>%
      dplyr::ungroup()
  }

  if (!is.null(top_n_gene)) {
    top_genes <- d %>%
      dplyr::group_by(gene) %>%
      dplyr::summarise(best_p = min(p_snp, na.rm = TRUE), .groups = "drop") %>%
      dplyr::arrange(best_p) %>%
      dplyr::slice_head(n = top_n_gene) %>%
      dplyr::pull(gene)
    d <- d %>% dplyr::filter(gene %in% top_genes)
  }

  mod_tbl <- d %>%
    dplyr::group_by(module) %>%
    dplyr::summarise(p_mod = min(p_mod, na.rm = TRUE),
                     n_assoc = dplyr::n_distinct(gene), .groups = "drop")

  gene_tbl <- d %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(p_snp = min(p_snp, na.rm = TRUE),
                     n_assoc = dplyr::n_distinct(module), .groups = "drop")

  vertices <- dplyr::bind_rows(
    mod_tbl  %>% dplyr::transmute(name = module, type = "module", weight = n_assoc),
    gene_tbl %>% dplyr::transmute(name = gene,   type = "gene",   weight = n_assoc)
  )

  edges <- d %>% dplyr::transmute(from = module, to = gene, p_snp = p_snp)

  g <- igraph::graph_from_data_frame(edges, directed = FALSE, vertices = vertices)

  igraph::V(g)$shape       <- ifelse(igraph::V(g)$type == "module", "square", "circle")
  igraph::V(g)$frame.color <- NA
  igraph::V(g)$label.color <- "black"
  igraph::V(g)$color       <- ifelse(igraph::V(g)$type == "module",
                                     node_module_col, node_gene_col)

  mod_ids  <- which(igraph::V(g)$type == "module")
  gene_ids <- which(igraph::V(g)$type == "gene")

  igraph::V(g)$size <- rep(NA_real_, igraph::vcount(g))

  mod_w  <- igraph::V(g)$weight[mod_ids]
  igraph::V(g)$size[mod_ids] <- if (length(unique(mod_w)) == 1) {
    rep(mean(mod_size_range) * size_scale, length(mod_ids))
  } else {
    scales::rescale(mod_w, to = mod_size_range * size_scale)
  }

  gene_w <- igraph::V(g)$weight[gene_ids]
  igraph::V(g)$size[gene_ids] <- if (length(unique(gene_w)) == 1) {
    rep(mean(gene_size_range) * size_scale, length(gene_ids))
  } else {
    scales::rescale(gene_w, to = gene_size_range * size_scale)
  }

  igraph::E(g)$color <- "grey70"
  igraph::E(g)$width <- 1

  igraph::V(g)$label     <- NA_character_
  igraph::V(g)$label.cex <- ifelse(igraph::V(g)$type == "module", 0.9, 0.7)
  if (label_modules) igraph::V(g)$label[mod_ids]  <- igraph::V(g)$name[mod_ids]
  if (label_genes)   igraph::V(g)$label[gene_ids] <- igraph::V(g)$name[gene_ids]

  lay <- layout_fun(g)
  if (!is.null(out_pdf)) grDevices::pdf(out_pdf, width = width, height = height)
  plot(g, layout = lay, vertex.label.dist = 0, edge.curved = 0)

  usr    <- graphics::par("usr")
  x_left <- usr[1] + 0.02 * (usr[2] - usr[1])
  y_bot  <- usr[3] + 0.02 * (usr[4] - usr[3])
  graphics::legend(
    x = x_left, y = y_bot, xjust = 0, yjust = 0,
    legend = c("Module", "Gene/locus"),
    pt.bg  = c(node_module_col, node_gene_col),
    col    = c(node_module_col, node_gene_col),
    pch    = c(15, 19),
    pt.cex = 2,
    cex    = 1.1,
    bty    = "o", box.col = "grey30", bg = "white"
  )

  if (!is.null(out_pdf)) grDevices::dev.off()
  invisible(g)
}

# ── Simplified ego plot (no biological colour coding) ────────────────────────
plot_module_ego_asthma <- function(focal_module,
                                   df12,
                                   geneset_categories = NULL,
                                   gene_size      = 10,
                                   focal_size     = 30,
                                   other_mod_size = 22,
                                   label_genes    = TRUE,
                                   label_cex      = 0.75) {

  node_gene_col   <- "#AED6F1"   # light blue  – peripheral gene/locus nodes
  node_module_col <- "#D5D8DC"   # light grey  – module nodes

  focal_rows <- df12 %>% dplyr::filter(module == focal_module)
  if (nrow(focal_rows) == 0) {
    message("No genes for module: ", focal_module)
    return(invisible(NULL))
  }

  other_mod_rows <- df12 %>%
    dplyr::filter(merged_gene_name %in% focal_rows$merged_gene_name,
                  module != focal_module) %>%
    dplyr::select(module, merged_gene_name) %>%
    dplyr::distinct()

  focal_v <- data.frame(name = focal_module, vtype = "focal_module",
                         gene_raw = NA_character_, gene_loci = NA_character_,
                         stringsAsFactors = FALSE)

  gene_v <- focal_rows %>%
    dplyr::transmute(name = summary, vtype = "gene",
                     gene_raw = merged_gene_name, gene_loci = merged_loci)

  other_v <- other_mod_rows %>%
    dplyr::distinct(module) %>%
    dplyr::transmute(name = module, vtype = "other_module",
                     gene_raw = NA_character_, gene_loci = NA_character_)

  verts <- dplyr::bind_rows(focal_v, gene_v, other_v)

  edges_focal_gene <- focal_rows %>%
    dplyr::transmute(from = focal_module, to = summary)

  edges_gene_other <- other_mod_rows %>%
    dplyr::left_join(
      focal_rows %>% dplyr::select(merged_gene_name, summary),
      by = "merged_gene_name", relationship = "many-to-many"
    ) %>%
    dplyr::transmute(from = summary, to = module) %>%
    dplyr::distinct()

  all_edges <- dplyr::bind_rows(edges_focal_gene, edges_gene_other)

  g <- igraph::graph_from_data_frame(all_edges, directed = FALSE, vertices = verts)

  is_module_node <- igraph::V(g)$vtype %in% c("focal_module", "other_module")
  igraph::V(g)$shape       <- ifelse(is_module_node, "square", "circle")
  igraph::V(g)$frame.color <- NA
  igraph::V(g)$label.color <- "black"
  igraph::V(g)$color       <- ifelse(is_module_node, node_module_col, node_gene_col)
  igraph::V(g)$size        <- ifelse(igraph::V(g)$vtype == "focal_module", focal_size,
                                      ifelse(igraph::V(g)$vtype == "other_module",
                                             other_mod_size, gene_size))

  igraph::V(g)$label <- vapply(seq_len(igraph::vcount(g)), function(i) {
    vt <- igraph::V(g)$vtype[i]
    nm <- igraph::V(g)$name[i]
    if (vt == "focal_module") nm
    else if (vt == "other_module") NA_character_
    else if (label_genes) igraph::V(g)$gene_raw[i]
    else NA_character_
  }, character(1))

  igraph::V(g)$label.cex <- ifelse(igraph::V(g)$vtype == "focal_module", 0.85,
                                    ifelse(igraph::V(g)$vtype == "other_module", 0.7,
                                           label_cex))

  igraph::E(g)$color <- "grey70"
  igraph::E(g)$width <- 1

  # ── concentric layout ─────────────────────────────────────────────────────
  focal_idx <- which(igraph::V(g)$vtype == "focal_module")
  gene_idx  <- which(igraph::V(g)$vtype == "gene")
  other_idx <- which(igraph::V(g)$vtype == "other_module")
  n_genes   <- length(gene_idx)
  n_others  <- length(other_idx)

  lay <- matrix(0, nrow = igraph::vcount(g), ncol = 2)
  lay[focal_idx, ] <- c(0, 0)

  if (n_genes > 0) {
    th <- seq(0, 2 * pi, length.out = n_genes + 1)[seq_len(n_genes)]
    lay[gene_idx, 1] <- cos(th)
    lay[gene_idx, 2] <- sin(th)
  }

  if (n_others > 0) {
    gene_angles <- setNames(atan2(lay[gene_idx, 2], lay[gene_idx, 1]),
                             igraph::V(g)$name[gene_idx])
    other_names <- igraph::V(g)$name[other_idx]
    for (j in seq_along(other_idx)) {
      connected_summaries <- other_mod_rows %>%
        dplyr::filter(module == other_names[j]) %>%
        dplyr::left_join(focal_rows %>% dplyr::select(merged_gene_name, summary),
                         by = "merged_gene_name") %>%
        dplyr::pull(summary)
      valid_angles <- gene_angles[intersect(connected_summaries, names(gene_angles))]
      mean_angle <- if (length(valid_angles) > 0)
        atan2(mean(sin(valid_angles)), mean(cos(valid_angles)))
      else
        2 * pi * (j - 1) / n_others
      lay[other_idx[j], 1] <- 2 * cos(mean_angle)
      lay[other_idx[j], 2] <- 2 * sin(mean_angle)
    }
  }

  # per-node label placement: gene labels point outward
  n_v        <- igraph::vcount(g)
  lbl_dist   <- rep(0.5, n_v)
  lbl_degree <- rep(-pi / 2, n_v)
  for (i in seq_len(n_v)) {
    if (igraph::V(g)$vtype[i] == "gene") {
      lbl_degree[i] <- atan2(lay[i, 2], lay[i, 1])
      lbl_dist[i]   <- 1.5
    }
  }
  igraph::V(g)$label.dist   <- lbl_dist
  igraph::V(g)$label.degree <- lbl_degree

  cat_label <- if (!is.null(geneset_categories)) {
    cat_row <- geneset_categories[geneset_categories[[1]] == focal_module, , drop = FALSE]
    if (nrow(cat_row) > 0 && !is.na(cat_row[[2]][1]) && nzchar(cat_row[[2]][1]))
      cat_row[[2]][1]
    else
      NA_character_
  } else NA_character_

  main_title <- if (!is.na(cat_label))
    paste0("Module ", focal_module, "\n", cat_label)
  else
    paste0("Module ", focal_module)

  plot(g, layout = lay, main = main_title, edge.curved = 0.2)

  legend(
    "bottomright",
    legend = c("Module", "Gene/locus"),
    pt.bg  = c(node_module_col, node_gene_col),
    col    = "black",
    pch    = c(22, 21),
    pt.cex = 1.6,
    bty    = "n", cex = 0.8
  )

  invisible(g)
}
```

```{r plot-asthma-network, fig.width=20, fig.height=20}
plot_gene_module_network_asthma(
  df12      = merged_df,
  top_n_gene = 800
)
```

```{r}
library(readxl)
geneset_categories <- read_excel("/project/xuanyao/jiaming/paper/output/mediation_analysis/asthma/K50/geneset_categories.xlsx")
```


```{r plot-asthma-ego, fig.width=12, fig.height=12}
setwd("/project/xuanyao/jiaming/paper/output/mediation_analysis/asthma/K50")
for (mod in sort(unique(merged_df$module))) {
  plot_module_ego_asthma(
    focal_module       = mod,
    df12               = merged_df,
    geneset_categories = geneset_categories
  )
  pdf(paste0("asthma_ego_", mod, ".pdf"), width = 12, height = 12)
  plot_module_ego_asthma(
    focal_module       = mod,
    df12               = merged_df,
    geneset_categories = geneset_categories
  )
  dev.off()
}
```





### modules most related to asthma

```{r}
target_modules <- c("k9", "k20", "k24", "k25", "k28", "k32", "k35")

merged_sub <- merged %>% filter(module %in% target_modules)

X_sub <- merged_sub[, c("snp_geneset", "geneset_trait")]
result_sub <- adaFilter(X_sub, r = 2, alpha = 0.1)

merged_sub$decision <- result_sub$decision

merged_sub[merged_sub$decision == 1, ]
```

### modules related to immune

```{r}
target_modules <- c(  "k5", "k7", "k8", "k9", "k10", "k12", "k13",
  "k20", "k21", "k24", "k25", "k26", "k28",
  "k32", "k33", "k34", "k35", "k40", "k42",
  "k46", "k47")

merged_sub <- merged %>% filter(module %in% target_modules)

X_sub <- merged_sub[, c("snp_geneset", "geneset_trait")]
result_sub <- adaFilter(X_sub, r = 2, alpha = 0.1)

merged_sub$decision <- result_sub$decision

merged_sub[merged_sub$decision == 1, ]
```
```{r}
x<-merged[merged$decision==1,]
length(unique(x$module))
table(x$module)
```

### NTC


```{r}
library(data.table)
library(dplyr)
library(stringr)
library(purrr)
library(heavytailcombtest)

# -----------------------------
# 0) Read inputs
# -----------------------------
df <- fread("/project/xuanyao/jiaming/Getting_started/data/geneset_to_trait/asthma/asthma.tsv")
df2 <- df[grepl("M1.01", df$Name), ]

Nadig_associated_perturbation <- rownames(snp_geneset_p_value_control)



discovery_pairs_cis <- readRDS("/project/xuanyao/jiaming/Getting_started/data/CDT4/standard_input_format_asthma/flames+200kb/discovery_pairs_cis.rds")


 gene_modules_id <- readRDS("/project/xuanyao/jiaming/Getting_started/data/CDT4/topic_geneset/flames+200kb/de_50_0.01_2_0.5.rds")




# -----------------------------
# 1) Build pv_map from df2
# -----------------------------
gene_id <- str_extract(df2$Name, "^[^(]+")

lookup <- df2 %>%
  mutate(ID = gene_id) %>%
  filter(!is.na(ID), !is.na(p_value)) %>%
  group_by(ID) %>%
  summarise(p_value = min(p_value), .groups = "drop")

pv_map <- setNames(lookup$p_value, lookup$ID)


# -----------------------------
# 2) Helpers
# -----------------------------
clean_p <- function(p) {
  p <- as.numeric(p)
  p[is.na(p) | p <= 0] <- NA
  p
}

safe_comb <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  combination.test(
    p,
    weights = NA,
    method = "Cauchy",
    tail.idx = 1
  )
}

safe_comb_fisher <- function(p) {
  p <- as.numeric(p)
  p <- p[!is.na(p) & p > 0 & p <= 1]
  if (length(p) == 0) return(NA_real_)
  if (length(p) == 1) return(p)

  stat <- -2 * sum(log(p))
  stats::pchisq(stat, df = 2 * length(p), lower.tail = FALSE)
}


# -----------------------------
# 3) Make filtered modules per perturbation (and record n_removed)
# -----------------------------
make_filtered_modules <- function(gene_symbol, gene_modules_id, pairs_df) {
  target_name <- gene_symbol

  ids_to_remove <- unique(pairs_df$response_id[pairs_df$grna_target == target_name])

  filtered_modules <- lapply(gene_modules_id, function(mod_ids) {
    setdiff(mod_ids, ids_to_remove)
  })

  list(
    modules   = filtered_modules,
    n_removed = length(ids_to_remove)
  )
}

filtered_by_perturbation <- setNames(
  lapply(
    Nadig_associated_perturbation,
    make_filtered_modules,
    gene_modules_id = gene_modules_id,
    pairs_df = discovery_pairs_cis
  ),
  Nadig_associated_perturbation
)


# -----------------------------
# 4) Compute module-level combined p-values per perturbation
#    (and add n_removed into module_results)
# -----------------------------
compute_for_one_pert <- function(pert) {
  obj <- filtered_by_perturbation[[pert]]
  gene_modules_id_pert <- obj$modules
  n_removed <- obj$n_removed

  # module -> p-values vector under this perturbation
  module_pvals <- lapply(gene_modules_id_pert, function(genes) {
    p <- pv_map[genes]
    names(p) <- genes
    p[!is.na(p)]
  })

  # combine within each module
  p_acat <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb(p))
  })

  p_fisher <- map_dbl(module_pvals, ~ {
    p <- clean_p(.x)
    if (!length(p)) return(NA_real_)
    as.numeric(safe_comb_fisher(p))
  })

  tibble(
    snp       = pert,
    n_removed = n_removed,  # ✅ 每个 perturbation 的 ids_to_remove 数量（复制到每一行 module）
    module    = names(module_pvals),
    n_genes   = lengths(gene_modules_id_pert),
    n_matched = sapply(module_pvals, function(v) sum(!is.na(v))),
    p_acat    = p_acat,
    p_fisher  = p_fisher
  )
}

module_results <- map_dfr(Nadig_associated_perturbation, compute_for_one_pert)

# 查看结果
module_results %>%
  arrange(p_fisher) %>%
  print(n = 10)


library(tidyr)
library(dplyr)
library(tibble)

snp_geneset_p_value<-  readRDS("/project/xuanyao/jiaming/paper/output/snake/CDT4_asthma/flames+200kb/K50/3000_t_ACAT_resample/p_values/geneset_p_value.rds")
snp_geneset_p_value_test <- snp_geneset_p_value[which(!grepl("NTC-",rownames(snp_geneset_p_value))),]
snp_geneset_p_value_control <- snp_geneset_p_value[which(grepl("NTC-",rownames(snp_geneset_p_value))),]

df1_long <- snp_geneset_p_value_control %>%
  as.data.frame() %>%
  rownames_to_column("snp") %>%      # 这里真正把 rownames 变成一列 snp
  pivot_longer(
    -snp,
    names_to  = "module",
    values_to = "p_snp_geneset"
  )

## 2. 模块层面的表，只保留需要的列
df2 <- module_results %>%
  dplyr::select(module, p_fisher,snp)

## 3. 按 module 合并，再调整列顺序
merged <- df1_long %>%
  left_join(df2, by = c("snp", "module")) %>%  # 关键改动
  relocate(snp, module, p_snp_geneset, p_fisher)%>% tidyr::drop_na()


# library(dplyr)
# 
# df1 <- as.data.frame(snp_geneset_p_value)
# df1$snp <- rownames(df1)
# 
# df1_long <- df1 %>%
#   pivot_longer(-snp, names_to = "module", values_to = "p_snp_geneset")
# 
# ## 图二：geneset_acat_pvals_M1.0001（列：module, p_acat）
# df2 <- geneset_acat_pvals_M1.0001 %>%
#   dplyr::select(module, p_fisher)
# 
# merged <- df1_long %>%
#   left_join(df2, by = "module") %>%
#   relocate(snp, module, p_snp_geneset, p_fisher)
# 
# 

library(adaFilter)

module_sizes <- lengths(gene_modules_id)
merged <- merged %>% filter(module_sizes[module] > 1)

X <- merged[,c(3,4)]
result <- adaFilter(X, r = 2,alpha=0.15)

merged$decision <- result$decision
colnames(merged)[c(3,4)] <- c("snp_geneset","geneset_trait")

merged[merged$decision==1,]
```



