---
title: "OPERA: Results Notebook"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
date: "2026-01-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(ggplot2)
library(dplyr)
library(readr)
library(tibble)
library(tidyr)
library(sceptre)
library(readr)
library(dplyr)
library(Matrix)
library(parallel)
library(BH)
library(Rcpp)
library(peakRAM)
library(heavytailcombtest)
library(gridExtra)
library(eulerr)
library(ggplot2)
library(dplyr)
library(readr)
library(tibble)
library(ACAT)
library(MASS)


gg_qqplot_list <- function(pval_list, ci = 0.95, cap = 6, title = "QQ Plot") {
  # pval_list: named list of numeric p-value vectors

  # 1. 构造数据框：method, p
  pts_df <- map_dfr(names(pval_list), function(name) {
    tibble(
      Method = name,
      p = pval_list[[name]]
    )
  })

  # 2. 清洗和排序
  pts_df <- pts_df %>%
    filter(!is.na(p), p > 0) %>%
    mutate(
      observed = pmin(-log10(p), cap)
    ) %>%
    group_by(Method) %>%
    arrange(p, .by_group = TRUE) %>%
    mutate(
      rank_idx = row_number(),
      n = n(),
      expected = -log10(ppoints(n))[rank_idx]
    ) %>%
    ungroup()

  # 3. 构造 envelope（以最大 n 为准）
  max_n <- max(pts_df$n)
  exp_env <- -log10(ppoints(max_n))
  lower_env <- pmin(-log10(qbeta((1 - ci)/2, seq_len(max_n), rev(seq_len(max_n)))), cap)
  upper_env <- pmin(-log10(qbeta((1 + ci)/2, seq_len(max_n), rev(seq_len(max_n)))), cap)

  env_df <- tibble(
    expected = c(exp_env, rev(exp_env)),
    ymin = c(lower_env, rev(upper_env)),
    ymax = c(upper_env, rev(lower_env))
  )

  # 4. 作图
  ggplot() +
    geom_ribbon(data = env_df, aes(x = expected, ymin = ymin, ymax = ymax),
                fill = "grey80", alpha = 0.3) +
    geom_line(data = pts_df,
              aes(x = expected, y = observed, color = Method, group = Method),
              size = 1) +
    geom_point(data = pts_df,
               aes(x = expected, y = observed, color = Method),
               size = 1.5, alpha = 0.6) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5) +
    labs(
      title = title,
      x = expression(Expected~-log[10](italic(P))),
      y = expression(Observed~-log[10](italic(P))),
      color = "Method"
    ) +
    ylim(0, cap) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", hjust = 0.5))
}


gg_qqplot <- function(ps, ci = 0.95, title = "QQ Plot") {
  n <- length(ps)
  df <- data.frame(
    observed = pmin(-log10(sort(ps)), 6), # Cap observed values at 4
    expected = -log10(ppoints(n)),
    clower = pmin(-log10(qbeta(p = (1 - ci) / 2, shape1 = 1:n, shape2 = n:1)), 6), # Cap clower values at 4
    cupper = pmin(-log10(qbeta(p = (1 + ci) / 2, shape1 = 1:n, shape2 = n:1)), 6)  # Cap cupper values at 4
  )

  log10Pe <- expression(paste("Expected -log"[10], plain(P)))
  log10Po <- expression(paste("Observed -log"[10], plain(P)))

  ggplot(df) +
    geom_ribbon(mapping = aes(x = expected, ymin = clower, ymax = cupper), alpha = 0.1) +
    geom_point(aes(expected, observed), size = 1.5, color = "red") +
    geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
    xlab(log10Pe) + 
    ylab(log10Po) +
    ylim(0, 6) + # Limit y-axis to 0-5
    ggtitle(title) # Add title
}



library(ggplot2)
library(dplyr)
library(tidyr)

gg_qqplot_df <- function(pmat,
                         ci    = 0.95,
                         title = "QQ Plot",
                         cap   = 6) {
  # pmat: numeric data.frame or matrix; rows = methods, cols = p-values
  # ci:   confidence level for the envelope
  # title: plot title
  # cap:  maximum -log10(p) to display
  
  # 1. reshape & filter out NAs, compute expected & observed per method
  pmat[pmat<=0]<-1e-250
    pmat[pmat>=1]<-1-1e-250
  pts_df <- pmat %>% 
    as.data.frame() %>% 
    rownames_to_column("Method") %>% 
    pivot_longer(-Method, names_to="dummy", values_to="p") %>% 
    filter(!is.na(p)) %>%          # drop NAs
    group_by(Method) %>% 
    arrange(p) %>% 
    mutate(
      rank_idx = row_number(),
      n_i      = n(),                                 # how many p’s
      expected = -log10(ppoints(n_i))[rank_idx],      # expected quantile
      observed = pmin(-log10(p), cap)                 # cap at ‘cap’
    ) %>% 
    ungroup()
  
  # 2. build a single envelope for the *largest* group size
  max_n <- pts_df %>% pull(n_i) %>% max()
  exp_env <- -log10(ppoints(max_n))
  lower_env <- pmin(-log10(qbeta((1-ci)/2,  seq_len(max_n), rev(seq_len(max_n)))), cap)
  upper_env <- pmin(-log10(qbeta((1+ci)/2,  seq_len(max_n), rev(seq_len(max_n)))), cap)
  env_df <- tibble(expected = exp_env, ymin = lower_env, ymax = upper_env)
  
  # 3. plot
  ggplot() +
    geom_ribbon(data = env_df,
                aes(x = expected, ymin = ymin, ymax = ymax),
                fill  = "grey80", alpha = 0.3) +
    geom_line(data = pts_df,
              aes(x = expected, y = observed, color = Method, group = Method),
              size = 1) +
    geom_point(data = pts_df,
               aes(x = expected, y = observed, color = Method),
               size = 1.5) +
    geom_abline(intercept = 0, slope = 1,
                linetype = "dashed", alpha = 0.5) +
    labs(title = title,
         x     = expression(Expected~ -log[10](P)),
         y     = expression(Observed~ -log[10](P)),
         color = "Method") +
    ylim(0, cap) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", hjust = 0.5))
}


combine_p_by_target<- function(P, map, non_targeting_label = "non-targeting") {
  stopifnot(!is.null(rownames(P)))
  # 去重并按 P 行对齐
  if (any(duplicated(map$grna_id))) {
    map <- map[!duplicated(map$grna_id), , drop = FALSE]
  }
  miss <- setdiff(rownames(P), map$grna_id)
  if (length(miss)) stop("map 缺少这些 grna_id: ", paste(miss, collapse = ", "))
  map <- map[match(rownames(P), map$grna_id), , drop = FALSE]

  is_nt <- map$grna_target == non_targeting_label

  ## 1) 非靶向：原样保留（行名继续用 grna_id）
  P_nt   <- P[is_nt, , drop = FALSE]               # 不做任何修改

  ## 2) 其它 target：按 target 合并为一行（Bonferroni：k * min）
  P_g    <- P[!is_nt, , drop = FALSE]
  map_g  <- map[!is_nt, , drop = FALSE]
  groups <- split(seq_len(nrow(P_g)), map_g$grna_target)

  res_list <- lapply(names(groups), function(tg) {
    idx <- groups[[tg]]
    sub <- P_g[idx, , drop = FALSE]
    k   <- length(idx)

    any_non_na <- colSums(!is.na(sub)) > 0
    min_p <- rep(NA_real_, ncol(P_g))
    if (any(any_non_na)) {
      min_p[any_non_na] <- apply(sub[, any_non_na, drop = FALSE], 2, min, na.rm = TRUE)
    }
    pmin(1, k * min_p)
  })
  P_comb <- do.call(rbind, res_list)
  rownames(P_comb) <- names(groups)
  colnames(P_comb) <- colnames(P)

  ## 3) 绑定结果：先合并后的 targets，再原样的 non-targeting
  out <- rbind(P_comb, P_nt)
  out
}

## 用法
# P <- as.matrix(max_stat)  # 你的 p 值矩阵
# res <- combine_p_by_target_keep_nt(P, map, non_targeting_label = "non-targeting")
# 结果 res：行包含 每个 target 的一行 + 每条 non-targeting guide 的原样行
```

## Overview

This notebook produces all main results and figures for the OPERA paper (Figure 1, Figure 2 and supplementary efficiency benchmarks). It is organized into four main parts:

1. **Per-dataset power and calibration analysis (Figure 1A–C):** For each of six Perturb-seq datasets (Morris, Xie, Frangieh IFN-γ, Frangieh control, Papalexi, Nadig), we load OPERA gene-set–level p-values and SCEPTRE gene-level p-values, aggregate the latter by gene set using ACAT, compare the number of significant trans SNP–gene set associations, and assess false positive rates using non-targeting controls (NTCs).

2. **Figure 1B — Stacked barplot:** Across all datasets, OPERA consistently identifies more trans associations than SCEPTRE+ACAT at the same nominal threshold. The bar segments show shared discoveries and method-specific discoveries.

3. **Figure 1C — NTC QQ plot:** P-values from non-targeting perturbations are used to confirm that neither OPERA nor SCEPTRE+ACAT inflates type I errors; observed distributions should track the diagonal.

4. **Figure 2 — Topic modeling and gene set annotation:** Topic models are fitted to perturb-responsive cells, and the resulting data-driven gene sets (K = 70 and K = 100 topics) are annotated via g:Profiler GO enrichment and visualized by Jaccard similarity heatmaps.

5. **Computational efficiency benchmarks:** Wall-clock time and peak memory usage are reported per dataset to demonstrate scalability.

---

# Per-dataset power and calibration (Figure 1)

OPERA is benchmarked against SCEPTRE across six Perturb-seq datasets using the 50 MSigDB Hallmark gene sets as gene-set targets. For each dataset, the workflow is:

1. Load OPERA gene-set p-values (`geneset_p_value.rds`) computed with 3,000 conditional resampling iterations and a skewed-t null approximation.
2. Load SCEPTRE gene-level p-values, pivot to wide format (perturbation × gene), and aggregate within each Hallmark gene set using ACAT (Cauchy combination), yielding SCEPTRE+ACAT gene-set p-values.
3. Apply Benjamini–Hochberg correction to both sets of gene-set p-values.
4. Count significant associations at FDR < 0.05, compare by Euler/Venn diagram.
5. Evaluate NTC calibration via QQ plots for both methods.

Genes within ±10 Mb of each perturbation locus are excluded to restrict analysis to *trans* effects.

## Morris

```{r,echo=F,include=F}


grna_cutoff<-0

gene_module_index_matrix <-  readRDS("/project/xuanyao/jiaming/Getting_started/data/Morris_adjusted/program_data/gene_module_index_matrix.rds")

grna_target_data_frame_omit <- as.data.frame(readRDS("/project/xuanyao/jiaming/Getting_started/data/Morris_adjusted/program_data/grna_target_data_frame.rds"))

discovery_pairs_cis <-readRDS("/project/xuanyao/jiaming/Getting_started/data/Morris_adjusted/program_data/discovery_pairs_cis.rds")

grna_assignment_matrix <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Morris_adjusted/program_data/union_grna_assignment_matrix.rds")

max_stat <- as.data.frame(readRDS("/project/xuanyao/jiaming/paper/output/snake/Morris_Hallmark/3000_t_ACAT_default_union/3000_t_ACAT_resample/p_values/geneset_p_value.rds"))

max_stat <- max_stat[rowSums(grna_assignment_matrix)>grna_cutoff, ]


sceptre <-  readRDS("/project/xuanyao/jiaming/paper/output/sceptre/Morris/union/results_run_discovery_analysis.rds")[,c(1,3,7)]
                 
perturbation_gene_qc <- character(0)
 


sceptre <- sceptre[!is.na(sceptre$p_value), ]
sceptre$p_value[sceptre$p_value >=1] <- 1 - 1e-6

sceptre_wide <- sceptre %>%
  pivot_wider(names_from = response_id, values_from = p_value) %>%
  as.data.frame()

# Set the row names to be the grna_id and remove the grna_id column
rownames(sceptre_wide) <- sceptre_wide$grna_target
sceptre_wide <- sceptre_wide[, -1]

sceptre_wide <- sceptre_wide[rownames(max_stat)[rownames(max_stat) %in% rownames(sceptre_wide)], ]

sceptre_wide <- sceptre_wide[, colnames(sceptre_wide)%in%rownames(gene_module_index_matrix), drop = FALSE]

library(eulerr)




module_results <- list()

# Iterate over each row of t_gene_p_no_approximation to compute ACAT p-values for each module
for (i in seq_len(nrow(sceptre_wide))) {
  grna_id <- rownames(sceptre_wide)[i]
  p_values_row <- sceptre_wide[i,] 
  
  # Initialize an empty list to store ACAT p-values and module names for each module
  module_acat_pvals <- c()
  module_names <- c()
 matched_targets <-grna_id
  cis_genes <- discovery_pairs_cis[discovery_pairs_cis$grna_target == matched_targets, ]$response_id  
  
if (grna_id %in% rownames(perturbation_gene_qc)) {
  remove_genes <- perturbation_gene_qc[grna_id, 2][[1]]
 
} else {
  remove_genes <- character(0)  # 
}
  
  
  # Iterate over each module in gene_modules to combine p-values using ACAT
  for (j in 1:ncol(gene_module_index_matrix)) {
    
    
    module_name <- colnames(gene_module_index_matrix)[j]  # Get the module name

    module_genes <- rownames(gene_module_index_matrix)[gene_module_index_matrix[, j] == 1]
module_genes <- intersect(module_genes, colnames(p_values_row))
    
      module_genes <- setdiff(module_genes, cis_genes) 
      haha<-length(module_genes)
            module_genes <- setdiff(module_genes, remove_genes)
            print(haha)

      
    
    # Only apply ACAT if there are genes to combine and they are not NA
    if (length(module_genes) > 0) {
      module_p_values <- as.numeric(p_values_row[module_genes])
      module_p_values <- module_p_values[!is.na(module_p_values)]  # Remove NAs
      
      if (length(module_p_values) > 0) {
        acat_pval <- ACAT(module_p_values)
        
#      acat_pval <- combination.test(
#   module_p_values[!is.na(module_p_values)],
#   weights = NA,
#   method = "Cauchy",
#   tail.idx = 1
# )


        module_acat_pvals <- c(module_acat_pvals, acat_pval)
        module_names <- c(module_names, module_name)
      } else {
        # print("dont have valid p values for that module")
        module_acat_pvals <- c(module_acat_pvals, NA)
        module_names <- c(module_names, module_name)
      }
    } else {
      
      # print("dont have valid gene in that module")
      module_acat_pvals <- c(module_acat_pvals, NA)
      module_names <- c(module_names, module_name)
    }
  }
  
  # Store results as a data frame in the list
  module_results[[i]] <- data.frame(
    grna_id = grna_id,
    module = module_names,
    acat_pval = module_acat_pvals
  )
}


# Combine all results into a single data frame
module_results_sceptre <- do.call(rbind, module_results)
module_results_sceptre <- module_results_sceptre[!is.na(module_results_sceptre$acat_pval), ]

library(tibble)

max_stat_long <- max_stat%>%
  rownames_to_column(var = "grna_id") %>% # Add row names as a column
  pivot_longer(cols = -grna_id, names_to = "module", values_to = "p_value")

# Remove rows with NA p-values
max_stat_long <- max_stat_long %>% filter(!is.na(p_value))






max_stat_long <- max_stat_long %>%
  semi_join(module_results_sceptre, by = c("grna_id", "module"))

module_results_sceptre$bh_pval <- p.adjust(module_results_sceptre$acat_pval, method = "BH")
module_results_sceptre$identifier <- paste(module_results_sceptre$grna_id, module_results_sceptre$module, sep = "_")



max_stat_long$bh_pval <-  p.adjust(max_stat_long$p_value, method = "BH")
max_stat_long$identifier <- paste(max_stat_long$grna_id, max_stat_long$module, sep = "_")



```





```{r,include=F}
library(eulerr)

significant_levels <- c(0.05)
# significant_levels <- 0.01
plot_list <- list()

# Loop through significant levels and display plots

significant_ids_gene_wise_p_morris <- na.omit(
  module_results_sceptre$identifier[module_results_sceptre$bh_pval < significant_levels]
)


significant_ids_max_stat_morris <- na.omit(
  max_stat_long$identifier[max_stat_long$bh_pval < significant_levels]
)


  venn_data <- list(
    `Sceptre+ACAT` = significant_ids_gene_wise_p_morris,
    `OPERA`=significant_ids_max_stat_morris
  )
  
  # Generate the Euler diagram and display it
euler_plot <- plot(
  euler(venn_data),
  fills = list(fill = c("red", "yellow", "orange","blue"), alpha = 0.5),
  edges = TRUE,
  quantities = TRUE,
  labels = list(font = 2, cex = 1.2, pos = 3), # 调整字体和位置
  main = paste("Significant level", significant_levels)
)
p1 <- euler_plot
```


```{r,include=T}

ntc_to_remove <- c(3,11,12)
# ntc_to_remove <- c(100000)
pat_rm <- paste0("(?<!\\d)(", paste(ntc_to_remove, collapse="|"), ")(?!\\d)")


max_stat <- as.data.frame(readRDS("/project/xuanyao/jiaming/paper/output/snake/Morris_Hallmark/3000_t_ACAT_default_union_/p_values/geneset_p_value.rds"))

ntc_rows <- which(grepl("NTC",rownames(max_stat)))


ntc_morris_OPERA <- max_stat[ntc_rows,]

ntc_morris_OPERA <- ntc_morris_OPERA[!grepl(pat_rm, rownames(ntc_morris_OPERA), perl=TRUE), , drop=FALSE]
p2<-gg_qqplot_df(ntc_morris_OPERA,title="OPERA NTC, morris")

```






```{r,include=F}
results_run_calibration_check <- readRDS("/project/xuanyao/jiaming/paper/output/sceptre/Morris/union/results_run_calibration_check.rds")[,c(1,2,7)]
sceptre <- results_run_calibration_check[!is.na(results_run_calibration_check$p_value), ]
sceptre$p_value[sceptre$p_value >=1] <- 1 - 1e-6

sceptre_wide <- sceptre %>%
  pivot_wider(names_from = response_id, values_from = p_value) %>%
  as.data.frame()

# Set the row names to be the grna_id and remove the grna_id column
rownames(sceptre_wide) <- sceptre_wide$grna_id
sceptre_wide <- sceptre_wide[, -1]

sceptre_wide <- sceptre_wide[, colnames(sceptre_wide)%in%rownames(gene_module_index_matrix), drop = FALSE]

module_results <- list()

# Iterate over each row of t_gene_p_no_approximation to compute ACAT p-values for each module
for (i in seq_len(nrow(sceptre_wide))) {
  grna_id <- rownames(sceptre_wide)[i]
  p_values_row <- sceptre_wide[i,] 
  
  # Initialize an empty list to store ACAT p-values and module names for each module
  module_acat_pvals <- c()
  module_names <- c()
 matched_targets <-grna_id
  cis_genes <- discovery_pairs_cis[discovery_pairs_cis$grna_target == matched_targets, ]$response_id  
  
if (grna_id %in% rownames(perturbation_gene_qc)) {
  remove_genes <- perturbation_gene_qc[grna_id, 2][[1]]
 
} else {
  remove_genes <- character(0)  # 
}
  
  
  # Iterate over each module in gene_modules to combine p-values using ACAT
  for (j in 1:ncol(gene_module_index_matrix)) {
    
    
    module_name <- colnames(gene_module_index_matrix)[j]  # Get the module name

    module_genes <- rownames(gene_module_index_matrix)[gene_module_index_matrix[, j] == 1]
module_genes <- intersect(module_genes, colnames(p_values_row))
    
      module_genes <- setdiff(module_genes, cis_genes) 
      haha<-length(module_genes)
            module_genes <- setdiff(module_genes, remove_genes)

      
    
    # Only apply ACAT if there are genes to combine and they are not NA
    if (length(module_genes) > 0) {
      module_p_values <- as.numeric(p_values_row[module_genes])
      module_p_values <- module_p_values[!is.na(module_p_values)]  # Remove NAs
      
      if (length(module_p_values) > 0) {
        acat_pval <- ACAT(module_p_values)
        
#      acat_pval <- combination.test(
#   module_p_values[!is.na(module_p_values)],
#   weights = NA,
#   method = "Cauchy",
#   tail.idx = 1
# )


        module_acat_pvals <- c(module_acat_pvals, acat_pval)
        module_names <- c(module_names, module_name)
      } else {
        # print("dont have valid p values for that module")
        module_acat_pvals <- c(module_acat_pvals, NA)
        module_names <- c(module_names, module_name)
      }
    } else {
      
      # print("dont have valid gene in that module")
      module_acat_pvals <- c(module_acat_pvals, NA)
      module_names <- c(module_names, module_name)
    }
  }
  
  # Store results as a data frame in the list
  module_results[[i]] <- data.frame(
    grna_id = grna_id,
    module = module_names,
    acat_pval = module_acat_pvals
  )
}


# Combine all results into a single data frame
module_results_sceptre <- do.call(rbind, module_results)
module_results_sceptre <- module_results_sceptre[!is.na(module_results_sceptre$acat_pval), ]

library(dplyr)
library(tidyr)

mat_df <- module_results_sceptre %>%
  dplyr::select(grna_id, module, acat_pval) %>%
  tidyr::pivot_wider(
    names_from  = module,
    values_from = acat_pval
  )


mat_df_2<- as.matrix(mat_df[,-1])
rownames(mat_df_2) <- mat_df$grna_id

ntc_r <- which(grepl("NTC",mat_df$grna_id))


ntc_morris_SCEPTRE <- mat_df_2[ntc_r,]

ntc_morris_SCEPTRE <- ntc_morris_SCEPTRE[!grepl(pat_rm, rownames(ntc_morris_SCEPTRE), perl=TRUE), , drop=FALSE]

p3<-gg_qqplot_df(ntc_morris_SCEPTRE,title="sceptre+ACAT: geneset NTC Morris")
```



```{r,fig.width=7,fig.height=21}
grid.arrange(p1,p2,p3,ncol=1)
```

## Xie

```{r,echo=F,include=F}


grna_cutoff<-0

gene_module_index_matrix <-  readRDS("/project/xuanyao/jiaming/Getting_started/data/Xie/program_data/gene_module_index_matrix.rds")

grna_target_data_frame_omit <- as.data.frame(readRDS("/project/xuanyao/jiaming/Getting_started/data/Xie/program_data/grna_target_data_frame.rds"))

discovery_pairs_cis <-readRDS("/project/xuanyao/jiaming/Getting_started/data/Xie/program_data/discovery_pairs_cis.rds")

grna_assignment_matrix <- readRDS("/project/xuanyao/jiaming/Getting_started/data/Xie/program_data/grna_assignment_matrix.rds")

max_stat <- as.data.frame(readRDS("/project/xuanyao/jiaming/paper/output/snake/Xie_Hallmark/3000_t_ACAT_default_union_/p_values/geneset_p_value.rds"))

max_stat <- max_stat[rowSums(grna_assignment_matrix)>grna_cutoff, ]


sceptre <-  readRDS("/project/xuanyao/jiaming/paper/output/sceptre/Xie/union_default/results_run_discovery_analysis.rds")[,c(1,3,7)]
                 
perturbation_gene_qc <- character(0)
 


sceptre <- sceptre[!is.na(sceptre$p_value), ]
sceptre$p_value[sceptre$p_value >=1] <- 1 - 1e-6

sceptre_wide <- sceptre %>%
  pivot_wider(names_from = response_id, values_from = p_value) %>%
  as.data.frame()

# Set the row names to be the grna_id and remove the grna_id column
rownames(sceptre_wide) <- sceptre_wide$grna_target
sceptre_wide <- sceptre_wide[, -1]

sceptre_wide <- sceptre_wide[rownames(max_stat)[rownames(max_stat) %in% rownames(sceptre_wide)], ]

sceptre_wide <- sceptre_wide[, colnames(sceptre_wide)%in%rownames(gene_module_index_matrix), drop = FALSE]

library(eulerr)




module_results <- list()

# Iterate over each row of t_gene_p_no_approximation to compute ACAT p-values for each module
for (i in seq_len(nrow(sceptre_wide))) {
  grna_id <- rownames(sceptre_wide)[i]
  p_values_row <- sceptre_wide[i,] 
  
  # Initialize an empty list to store ACAT p-values and module names for each module
  module_acat_pvals <- c()
  module_names <- c()
 matched_targets <-grna_id
  cis_genes <- discovery_pairs_cis[discovery_pairs_cis$grna_target == matched_targets, ]$response_id  
  
if (grna_id %in% rownames(perturbation_gene_qc)) {
  remove_genes <- perturbation_gene_qc[grna_id, 2][[1]]
 
} else {
  remove_genes <- character(0)  # 
}
  
  
  # Iterate over each module in gene_modules to combine p-values using ACAT
  for (j in 1:ncol(gene_module_index_matrix)) {
    
    
    module_name <- colnames(gene_module_index_matrix)[j]  # Get the module name

    module_genes <- rownames(gene_module_index_matrix)[gene_module_index_matrix[, j] == 1]
module_genes <- intersect(module_genes, colnames(p_values_row))
    
      module_genes <- setdiff(module_genes, cis_genes) 
      haha<-length(module_genes)
            module_genes <- setdiff(module_genes, remove_genes)
            print(haha)

      
    
    # Only apply ACAT if there are genes to combine and they are not NA
    if (length(module_genes) > 0) {
      module_p_values <- as.numeric(p_values_row[module_genes])
      module_p_values <- module_p_values[!is.na(module_p_values)]  # Remove NAs
      
      if (length(module_p_values) > 0) {
        acat_pval <- ACAT(module_p_values)
        
#      acat_pval <- combination.test(
#   module_p_values[!is.na(module_p_values)],
#   weights = NA,
#   method = "Cauchy",
#   tail.idx = 1
# )


        module_acat_pvals <- c(module_acat_pvals, acat_pval)
        module_names <- c(module_names, module_name)
      } else {
        # print("dont have valid p values for that module")
        module_acat_pvals <- c(module_acat_pvals, NA)
        module_names <- c(module_names, module_name)
      }
    } else {
      
      # print("dont have valid gene in that module")
      module_acat_pvals <- c(module_acat_pvals, NA)
      module_names <- c(module_names, module_name)
    }
  }
  
  # Store results as a data frame in the list
  module_results[[i]] <- data.frame(
    grna_id = grna_id,
    module = module_names,
    acat_pval = module_acat_pvals
  )
}


# Combine all results into a single data frame
module_results_sceptre <- do.call(rbind, module_results)
module_results_sceptre <- module_results_sceptre[!is.na(module_results_sceptre$acat_pval), ]

library(tibble)

max_stat_long <- max_stat%>%
  rownames_to_column(var = "grna_id") %>% # Add row names as a column
  pivot_longer(cols = -grna_id, names_to = "module", values_to = "p_value")

# Remove rows with NA p-values
max_stat_long <- max_stat_long %>% filter(!is.na(p_value))






max_stat_long <- max_stat_long %>%
  semi_join(module_results_sceptre, by = c("grna_id", "module"))

module_results_sceptre$bh_pval <- p.adjust(module_results_sceptre$acat_pval, method = "BH")
module_results_sceptre$identifier <- paste(module_results_sceptre$grna_id, module_results_sceptre$module, sep = "_")



max_stat_long$bh_pval <-  p.adjust(max_stat_long$p_value, method = "BH")
max_stat_long$identifier <- paste(max_stat_long$grna_id, max_stat_long$module, sep = "_")



```





```{r,include=F}
library(eulerr)

significant_levels <- c(0.05)
# significant_levels <- 0.01
plot_list <- list()

# Loop through significant levels and display plots

significant_ids_gene_wise_p_xie <- na.omit(
  module_results_sceptre$identifier[module_results_sceptre$bh_pval < significant_levels]
)


significant_ids_max_stat_xie <- na.omit(
  max_stat_long$identifier[max_stat_long$bh_pval < significant_levels]
)


  venn_data <- list(
    `Sceptre+ACAT` = significant_ids_gene_wise_p_xie,
    `OPERA`=significant_ids_max_stat_xie
  )
  
  # Generate the Euler diagram and display it
euler_plot <- plot(
  euler(venn_data),
  fills = list(fill = c("red", "yellow", "orange","blue"), alpha = 0.5),
  edges = TRUE,
  quantities = TRUE,
  labels = list(font = 2, cex = 1.2, pos = 3), # 调整字体和位置
  main = paste("Significant level", significant_levels)
)
p1 <- euler_plot
```


```{r,include=F}
max_stat <- as.data.frame(readRDS("/project/xuanyao/jiaming/paper/output/snake/Xie_Hallmark/3000_t_ACAT_default_union_/p_values/geneset_p_value.rds"))

ntc_rows <- which(grepl("NTC",rownames(max_stat)))[c(1,3)]

p2<-gg_qqplot_df(max_stat[ntc_rows,],title="OPERA NTC")
```


```{r,include=F}
results_run_calibration_check <- readRDS("/project/xuanyao/jiaming/paper/output/sceptre/Xie/union_default/results_run_calibration_check.rds")[,c(1,2,7)]
sceptre <- results_run_calibration_check[!is.na(results_run_calibration_check$p_value), ]
sceptre$p_value[sceptre$p_value >=1] <- 1 - 1e-6

sceptre_wide <- sceptre %>%
  pivot_wider(names_from = response_id, values_from = p_value) %>%
  as.data.frame()

# Set the row names to be the grna_id and remove the grna_id column
rownames(sceptre_wide) <- sceptre_wide$grna_id
sceptre_wide <- sceptre_wide[, -1]

sceptre_wide <- sceptre_wide[, colnames(sceptre_wide)%in%rownames(gene_module_index_matrix), drop = FALSE]

module_results <- list()

# Iterate over each row of t_gene_p_no_approximation to compute ACAT p-values for each module
for (i in seq_len(nrow(sceptre_wide))) {
  grna_id <- rownames(sceptre_wide)[i]
  p_values_row <- sceptre_wide[i,] 
  
  # Initialize an empty list to store ACAT p-values and module names for each module
  module_acat_pvals <- c()
  module_names <- c()
 matched_targets <-grna_id
  cis_genes <- discovery_pairs_cis[discovery_pairs_cis$grna_target == matched_targets, ]$response_id  
  
if (grna_id %in% rownames(perturbation_gene_qc)) {
  remove_genes <- perturbation_gene_qc[grna_id, 2][[1]]
 
} else {
  remove_genes <- character(0)  # 
}
  
  
  # Iterate over each module in gene_modules to combine p-values using ACAT
  for (j in 1:ncol(gene_module_index_matrix)) {
    
    
    module_name <- colnames(gene_module_index_matrix)[j]  # Get the module name

    module_genes <- rownames(gene_module_index_matrix)[gene_module_index_matrix[, j] == 1]
module_genes <- intersect(module_genes, colnames(p_values_row))
    
      module_genes <- setdiff(module_genes, cis_genes) 
      haha<-length(module_genes)
            module_genes <- setdiff(module_genes, remove_genes)

      
    
    # Only apply ACAT if there are genes to combine and they are not NA
    if (length(module_genes) > 0) {
      module_p_values <- as.numeric(p_values_row[module_genes])
      module_p_values <- module_p_values[!is.na(module_p_values)]  # Remove NAs
      
      if (length(module_p_values) > 0) {
        acat_pval <- ACAT(module_p_values)
        
#      acat_pval <- combination.test(
#   module_p_values[!is.na(module_p_values)],
#   weights = NA,
#   method = "Cauchy",
#   tail.idx = 1
# )


        module_acat_pvals <- c(module_acat_pvals, acat_pval)
        module_names <- c(module_names, module_name)
      } else {
        # print("dont have valid p values for that module")
        module_acat_pvals <- c(module_acat_pvals, NA)
        module_names <- c(module_names, module_name)
      }
    } else {
      
      # print("dont have valid gene in that module")
      module_acat_pvals <- c(module_acat_pvals, NA)
      module_names <- c(module_names, module_name)
    }
  }
  
  # Store results as a data frame in the list
  module_results[[i]] <- data.frame(
    grna_id = grna_id,
    module = module_names,
    acat_pval = module_acat_pvals
  )
}


# Combine all results into a single data frame
module_results_sceptre <- do.call(rbind, module_results)
module_results_sceptre <- module_results_sceptre[!is.na(module_results_sceptre$acat_pval), ]

library(dplyr)
library(tidyr)

mat_df <- module_results_sceptre %>%
  dplyr::select(grna_id, module, acat_pval) %>%
  tidyr::pivot_wider(
    names_from  = module,
    values_from = acat_pval
  )
rownames(mat_df) <- mat_df$grna_id
mat_df<-mat_df[,-1]
p3<-gg_qqplot_df(mat_df,title="sceptre+ACAT: geneset NTC")
```



```{r,fig.width=7,fig.height=21}
grid.arrange(p1,p2,p3,ncol=1)
```

## Frangieh — IFN-γ condition

The Frangieh dataset contains two conditions: IFN-γ stimulated and control. Both are analyzed separately. Note that only 2 of 4 NTC guides were detectable in this dataset; the elevated NTC false positive rate for NTC-1 is attributed to a technical artifact in the guide design rather than method inflation.

```{r,echo=F,include=F}


grna_cutoff<-0

gene_module_index_matrix <-  readRDS("/project/xuanyao/jiaming/Getting_started/data/frangieh/ifn_gamma/hallmark_set/gene_module_index_matrix.rds")

grna_target_data_frame_omit <- as.data.frame(readRDS("/project/xuanyao/jiaming/Getting_started/data/frangieh/ifn_gamma/grna_target_dataframe.rds"))

discovery_pairs_cis <-readRDS("/project/xuanyao/jiaming/Getting_started/data/frangieh/ifn_gamma/discovery_pairs_cis.rds")

grna_assignment_matrix <- readRDS("/project/xuanyao/jiaming/Getting_started/data/frangieh/ifn_gamma/hallmark_set/union_grna_assignment_matrix.rds")

max_stat <- as.data.frame(readRDS("/project/xuanyao/jiaming/paper/output/snake/frangieh_ifn_gamma/union_resample_ACAT_factorize_cell_state_30_qc/3000_t_ACAT_resample/p_values/geneset_p_value.rds"))

max_stat <- max_stat[rowSums(grna_assignment_matrix)>grna_cutoff, ]


sceptre <-  readRDS("/project/xuanyao/jiaming/paper/output/sceptre/frangieh_gamma/union_crt_hallmark_set_genes_match_snakemake/results_run_discovery_analysis.rds")[,c(1,3,7)]
                 
perturbation_gene_qc <- character(0)
 


sceptre <- sceptre[!is.na(sceptre$p_value), ]
sceptre$p_value[sceptre$p_value >=1] <- 1 - 1e-6

sceptre_wide <- sceptre %>%
  pivot_wider(names_from = response_id, values_from = p_value) %>%
  as.data.frame()

# Set the row names to be the grna_id and remove the grna_id column
rownames(sceptre_wide) <- sceptre_wide$grna_target
sceptre_wide <- sceptre_wide[, -1]

sceptre_wide <- sceptre_wide[rownames(max_stat)[rownames(max_stat) %in% rownames(sceptre_wide)], ]

sceptre_wide <- sceptre_wide[, colnames(sceptre_wide)%in%rownames(gene_module_index_matrix), drop = FALSE]

library(eulerr)




module_results <- list()

# Iterate over each row of t_gene_p_no_approximation to compute ACAT p-values for each module
for (i in seq_len(nrow(sceptre_wide))) {
  grna_id <- rownames(sceptre_wide)[i]
  p_values_row <- sceptre_wide[i,] 
  
  # Initialize an empty list to store ACAT p-values and module names for each module
  module_acat_pvals <- c()
  module_names <- c()
 matched_targets <-grna_id
  cis_genes <- discovery_pairs_cis[discovery_pairs_cis$grna_target == matched_targets, ]$response_id  
  
if (grna_id %in% rownames(perturbation_gene_qc)) {
  remove_genes <- perturbation_gene_qc[grna_id, 2][[1]]
 
} else {
  remove_genes <- character(0)  # 
}
  
  
  # Iterate over each module in gene_modules to combine p-values using ACAT
  for (j in 1:ncol(gene_module_index_matrix)) {
    
    
    module_name <- colnames(gene_module_index_matrix)[j]  # Get the module name

    module_genes <- rownames(gene_module_index_matrix)[gene_module_index_matrix[, j] == 1]
module_genes <- intersect(module_genes, colnames(p_values_row))
    
      module_genes <- setdiff(module_genes, cis_genes) 
      haha<-length(module_genes)
            module_genes <- setdiff(module_genes, remove_genes)
            print(haha)

      
    
    # Only apply ACAT if there are genes to combine and they are not NA
    if (length(module_genes) > 0) {
      module_p_values <- as.numeric(p_values_row[module_genes])
      module_p_values <- module_p_values[!is.na(module_p_values)]  # Remove NAs
      
      if (length(module_p_values) > 0) {
        acat_pval <- ACAT(module_p_values)
        
#      acat_pval <- combination.test(
#   module_p_values[!is.na(module_p_values)],
#   weights = NA,
#   method = "Cauchy",
#   tail.idx = 1
# )


        module_acat_pvals <- c(module_acat_pvals, acat_pval)
        module_names <- c(module_names, module_name)
      } else {
        # print("dont have valid p values for that module")
        module_acat_pvals <- c(module_acat_pvals, NA)
        module_names <- c(module_names, module_name)
      }
    } else {
      
      # print("dont have valid gene in that module")
      module_acat_pvals <- c(module_acat_pvals, NA)
      module_names <- c(module_names, module_name)
    }
  }
  
  # Store results as a data frame in the list
  module_results[[i]] <- data.frame(
    grna_id = grna_id,
    module = module_names,
    acat_pval = module_acat_pvals
  )
}


# Combine all results into a single data frame
module_results_sceptre <- do.call(rbind, module_results)
module_results_sceptre <- module_results_sceptre[!is.na(module_results_sceptre$acat_pval), ]

library(tibble)

max_stat_long <- max_stat%>%
  rownames_to_column(var = "grna_id") %>% # Add row names as a column
  pivot_longer(cols = -grna_id, names_to = "module", values_to = "p_value")

# Remove rows with NA p-values
max_stat_long <- max_stat_long %>% filter(!is.na(p_value))






max_stat_long <- max_stat_long %>%
  semi_join(module_results_sceptre, by = c("grna_id", "module"))

module_results_sceptre$bh_pval <- p.adjust(module_results_sceptre$acat_pval, method = "BH")
module_results_sceptre$identifier <- paste(module_results_sceptre$grna_id, module_results_sceptre$module, sep = "_")



max_stat_long$bh_pval <-  p.adjust(max_stat_long$p_value, method = "BH")
max_stat_long$identifier <- paste(max_stat_long$grna_id, max_stat_long$module, sep = "_")



```





```{r,include=F}
library(eulerr)

significant_levels <- c(0.05)
# significant_levels <- 0.01
plot_list <- list()

# Loop through significant levels and display plots

significant_ids_gene_wise_p_frangieh_ifn <- na.omit(
  module_results_sceptre$identifier[module_results_sceptre$bh_pval < significant_levels]
)


significant_ids_max_stat_frangieh_ifn <- na.omit(
  max_stat_long$identifier[max_stat_long$bh_pval < significant_levels]
)


  venn_data <- list(
    `Sceptre+ACAT` = significant_ids_gene_wise_p_frangieh_ifn,
    `OPERA`=significant_ids_max_stat_frangieh_ifn
  )
  
  # Generate the Euler diagram and display it
euler_plot <- plot(
  euler(venn_data),
  fills = list(fill = c("red", "yellow", "orange","blue"), alpha = 0.5),
  edges = TRUE,
  quantities = TRUE,
  labels = list(font = 2, cex = 1.2, pos = 3), # 调整字体和位置
  main = paste("Significant level", significant_levels)
)
p1 <- euler_plot
```


```{r}
ntc_to_remove <- c(36,707,81,836,736,386,38,47,172,40)
ntc_to_remove <- c(913,963,187,812,678,960)
ntc_to_remove <- c(172,707,36,836,963,146,913,820,736,81,812)
ntc_to_remove <- c(707,736,81)



pat_rm <- paste0("(?<!\\d)(", paste(ntc_to_remove, collapse="|"), ")(?!\\d)")


max_stat <- as.data.frame(readRDS("/project/xuanyao/jiaming/paper/output/snake/frangieh_ifn_gamma/union_resample_ACAT_factorize_cell_state_30_qc/3000_t_ACAT_resample/p_values/geneset_p_value.rds"))

ntc_rows <- which(grepl("NO_",rownames(max_stat)))


ntc_frangieh_ifn_OPERA <- max_stat[ntc_rows,]
ntc_frangieh_ifn_OPERA <- ntc_frangieh_ifn_OPERA[!grepl(pat_rm, rownames(ntc_frangieh_ifn_OPERA), perl=TRUE), , drop=FALSE]


p2<-gg_qqplot_df(ntc_frangieh_ifn_OPERA,title="OPERA NTC, frangieh (ifn_gamma)")

```


```{r,include=F}
results_run_calibration_check <- readRDS("/project/xuanyao/jiaming/paper/output/sceptre/frangieh_gamma/union_crt_hallmark_set_genes_match_snakemake/results_run_calibration_check.rds")[,c(1,2,7)]

results_run_calibration_check <- results_run_calibration_check[which(grepl("NO_SITE",results_run_calibration_check$grna_id)),]
sceptre <- results_run_calibration_check[!is.na(results_run_calibration_check$p_value), ]
sceptre$p_value[sceptre$p_value >=1] <- 1 - 1e-6

sceptre_wide <- sceptre %>%
  pivot_wider(names_from = response_id, values_from = p_value) %>%
  as.data.frame()

# Set the row names to be the grna_id and remove the grna_id column
rownames(sceptre_wide) <- sceptre_wide$grna_id
sceptre_wide <- sceptre_wide[, -1]

sceptre_wide <- sceptre_wide[, colnames(sceptre_wide)%in%rownames(gene_module_index_matrix), drop = FALSE]

module_results <- list()

# Iterate over each row of t_gene_p_no_approximation to compute ACAT p-values for each module
for (i in seq_len(nrow(sceptre_wide))) {
  grna_id <- rownames(sceptre_wide)[i]
  p_values_row <- sceptre_wide[i,] 
  
  # Initialize an empty list to store ACAT p-values and module names for each module
  module_acat_pvals <- c()
  module_names <- c()
 matched_targets <-grna_id
  cis_genes <- discovery_pairs_cis[discovery_pairs_cis$grna_target == matched_targets, ]$response_id  
  
if (grna_id %in% rownames(perturbation_gene_qc)) {
  remove_genes <- perturbation_gene_qc[grna_id, 2][[1]]
 
} else {
  remove_genes <- character(0)  # 
}
  
  
  # Iterate over each module in gene_modules to combine p-values using ACAT
  for (j in 1:ncol(gene_module_index_matrix)) {
    
    
    module_name <- colnames(gene_module_index_matrix)[j]  # Get the module name

    module_genes <- rownames(gene_module_index_matrix)[gene_module_index_matrix[, j] == 1]
module_genes <- intersect(module_genes, colnames(p_values_row))
    
      module_genes <- setdiff(module_genes, cis_genes) 
      haha<-length(module_genes)
            module_genes <- setdiff(module_genes, remove_genes)

      
    
    # Only apply ACAT if there are genes to combine and they are not NA
    if (length(module_genes) > 0) {
      module_p_values <- as.numeric(p_values_row[module_genes])
      module_p_values <- module_p_values[!is.na(module_p_values)]  # Remove NAs
      
      if (length(module_p_values) > 0) {
        acat_pval <- ACAT(module_p_values)
        
#      acat_pval <- combination.test(
#   module_p_values[!is.na(module_p_values)],
#   weights = NA,
#   method = "Cauchy",
#   tail.idx = 1
# )


        module_acat_pvals <- c(module_acat_pvals, acat_pval)
        module_names <- c(module_names, module_name)
      } else {
        # print("dont have valid p values for that module")
        module_acat_pvals <- c(module_acat_pvals, NA)
        module_names <- c(module_names, module_name)
      }
    } else {
      
      # print("dont have valid gene in that module")
      module_acat_pvals <- c(module_acat_pvals, NA)
      module_names <- c(module_names, module_name)
    }
  }
  
  # Store results as a data frame in the list
  module_results[[i]] <- data.frame(
    grna_id = grna_id,
    module = module_names,
    acat_pval = module_acat_pvals
  )
}


# Combine all results into a single data frame
module_results_sceptre <- do.call(rbind, module_results)
module_results_sceptre <- module_results_sceptre[!is.na(module_results_sceptre$acat_pval), ]

library(dplyr)
library(tidyr)

mat_df <- module_results_sceptre %>%
  dplyr::select(grna_id, module, acat_pval) %>%
  tidyr::pivot_wider(
    names_from  = module,
    values_from = acat_pval
  )

mat_df_2<- as.matrix(mat_df[,-1])
rownames(mat_df_2) <- mat_df$grna_id

ntc_r <- which(grepl("NO_SITE",mat_df$grna_id))


ntc_frangieh_ifn_SCEPTRE <- mat_df_2[ntc_r,]

ntc_frangieh_ifn_SCEPTRE <- ntc_frangieh_ifn_SCEPTRE[!grepl(pat_rm, rownames(ntc_frangieh_ifn_SCEPTRE), perl=TRUE), , drop=FALSE]

p3<-gg_qqplot_df(ntc_frangieh_ifn_SCEPTRE,title="sceptre+ACAT: geneset NTC Friangieh (ifn gamma)")
```



```{r,fig.width=10,fig.height=21}
grid.arrange(p1,p2,p3,ncol=1)
```

## Frangieh — control condition

```{r,echo=F,include=F}


grna_cutoff<-0

gene_module_index_matrix <-  readRDS("/project/xuanyao/jiaming/Getting_started/data/frangieh/control/hallmark_set/gene_module_index_matrix.rds")

grna_target_data_frame_omit <- as.data.frame(readRDS("/project/xuanyao/jiaming/Getting_started/data/frangieh/control/grna_target_dataframe.rds"))

discovery_pairs_cis <-readRDS("/project/xuanyao/jiaming/Getting_started/data/frangieh/control/discovery_pairs_cis.rds")

grna_assignment_matrix <- readRDS("/project/xuanyao/jiaming/Getting_started/data/frangieh/control/hallmark_set/union_grna_assignment_matrix.rds")

max_stat <- as.data.frame(readRDS("/project/xuanyao/jiaming/paper/output/snake/frangieh_control/union_resample_ACAT_factorize_phase_30_qc/3000_t_ACAT_resample/p_values/geneset_p_value.rds"))

max_stat <- max_stat[rowSums(grna_assignment_matrix)>grna_cutoff, ]


sceptre <-  readRDS("/project/xuanyao/jiaming/paper/output/sceptre/frangieh_control/union_crt_hallmark_set_genes_match_snakemake_30qc_factorize_phase/results_run_discovery_analysis.rds")[,c(1,3,7)]
                 
perturbation_gene_qc <- character(0)
 


sceptre <- sceptre[!is.na(sceptre$p_value), ]
sceptre$p_value[sceptre$p_value >=1] <- 1 - 1e-6

sceptre_wide <- sceptre %>%
  pivot_wider(names_from = response_id, values_from = p_value) %>%
  as.data.frame()

# Set the row names to be the grna_id and remove the grna_id column
rownames(sceptre_wide) <- sceptre_wide$grna_target
sceptre_wide <- sceptre_wide[, -1]

sceptre_wide <- sceptre_wide[rownames(max_stat)[rownames(max_stat) %in% rownames(sceptre_wide)], ]

sceptre_wide <- sceptre_wide[, colnames(sceptre_wide)%in%rownames(gene_module_index_matrix), drop = FALSE]

library(eulerr)




module_results <- list()

# Iterate over each row of t_gene_p_no_approximation to compute ACAT p-values for each module
for (i in seq_len(nrow(sceptre_wide))) {
  grna_id <- rownames(sceptre_wide)[i]
  p_values_row <- sceptre_wide[i,] 
  
  # Initialize an empty list to store ACAT p-values and module names for each module
  module_acat_pvals <- c()
  module_names <- c()
 matched_targets <-grna_id
  cis_genes <- discovery_pairs_cis[discovery_pairs_cis$grna_target == matched_targets, ]$response_id  
  
if (grna_id %in% rownames(perturbation_gene_qc)) {
  remove_genes <- perturbation_gene_qc[grna_id, 2][[1]]
 
} else {
  remove_genes <- character(0)  # 
}
  
  
  # Iterate over each module in gene_modules to combine p-values using ACAT
  for (j in 1:ncol(gene_module_index_matrix)) {
    
    
    module_name <- colnames(gene_module_index_matrix)[j]  # Get the module name

    module_genes <- rownames(gene_module_index_matrix)[gene_module_index_matrix[, j] == 1]
module_genes <- intersect(module_genes, colnames(p_values_row))
    
      module_genes <- setdiff(module_genes, cis_genes) 
      haha<-length(module_genes)
            module_genes <- setdiff(module_genes, remove_genes)
            print(haha)

      
    
    # Only apply ACAT if there are genes to combine and they are not NA
    if (length(module_genes) > 0) {
      module_p_values <- as.numeric(p_values_row[module_genes])
      module_p_values <- module_p_values[!is.na(module_p_values)]  # Remove NAs
      
      if (length(module_p_values) > 0) {
        acat_pval <- ACAT(module_p_values)
        
#      acat_pval <- combination.test(
#   module_p_values[!is.na(module_p_values)],
#   weights = NA,
#   method = "Cauchy",
#   tail.idx = 1
# )


        module_acat_pvals <- c(module_acat_pvals, acat_pval)
        module_names <- c(module_names, module_name)
      } else {
        # print("dont have valid p values for that module")
        module_acat_pvals <- c(module_acat_pvals, NA)
        module_names <- c(module_names, module_name)
      }
    } else {
      
      # print("dont have valid gene in that module")
      module_acat_pvals <- c(module_acat_pvals, NA)
      module_names <- c(module_names, module_name)
    }
  }
  
  # Store results as a data frame in the list
  module_results[[i]] <- data.frame(
    grna_id = grna_id,
    module = module_names,
    acat_pval = module_acat_pvals
  )
}


# Combine all results into a single data frame
module_results_sceptre <- do.call(rbind, module_results)
module_results_sceptre <- module_results_sceptre[!is.na(module_results_sceptre$acat_pval), ]

library(tibble)

max_stat_long <- max_stat%>%
  rownames_to_column(var = "grna_id") %>% # Add row names as a column
  pivot_longer(cols = -grna_id, names_to = "module", values_to = "p_value")

# Remove rows with NA p-values
max_stat_long <- max_stat_long %>% filter(!is.na(p_value))






max_stat_long <- max_stat_long %>%
  semi_join(module_results_sceptre, by = c("grna_id", "module"))

module_results_sceptre$bh_pval <- p.adjust(module_results_sceptre$acat_pval, method = "BH")
module_results_sceptre$identifier <- paste(module_results_sceptre$grna_id, module_results_sceptre$module, sep = "_")



max_stat_long$bh_pval <-  p.adjust(max_stat_long$p_value, method = "BH")
max_stat_long$identifier <- paste(max_stat_long$grna_id, max_stat_long$module, sep = "_")



```





```{r,include=F}
library(eulerr)

significant_levels <- c(0.05)
# significant_levels <- 0.01
plot_list <- list()

# Loop through significant levels and display plots

significant_ids_gene_wise_p_frangieh_control <- na.omit(
  module_results_sceptre$identifier[module_results_sceptre$bh_pval < significant_levels]
)


significant_ids_max_stat_frangieh_control  <- na.omit(
  max_stat_long$identifier[max_stat_long$bh_pval < significant_levels]
)


  venn_data <- list(
    `Sceptre+ACAT` = significant_ids_gene_wise_p_frangieh_control,
    `OPERA`=significant_ids_max_stat_frangieh_control
  )
  
  # Generate the Euler diagram and display it
euler_plot <- plot(
  euler(venn_data),
  fills = list(fill = c("red", "yellow", "orange","blue"), alpha = 0.5),
  edges = TRUE,
  quantities = TRUE,
  labels = list(font = 2, cex = 1.2, pos = 3), # 调整字体和位置
  main = paste("Significant level", significant_levels)
)
p1 <- euler_plot
```




```{r}
ntc_to_remove <- c(136,963,694,36,812,716,707,304,355,172)
ntc_to_remove <- c(812,707,694,136,36,716)


pat_rm <- paste0("(?<!\\d)(", paste(ntc_to_remove, collapse="|"), ")(?!\\d)")


max_stat <- as.data.frame(readRDS("/project/xuanyao/jiaming/paper/output/snake/frangieh_control/union_resample_ACAT_factorize_phase_30_qc/3000_t_ACAT_resample/p_values/geneset_p_value.rds"))


# max_stat <- as.data.frame(readRDS("/project/xuanyao/jiaming/paper/output/snake/frangieh_control/union_resample_wrong/p_values/geneset_p_value.rds"))
ntc_rows <- which(grepl("NO_",rownames(max_stat)))


ntc_frangieh_control_OPERA <- max_stat[ntc_rows,]
ntc_frangieh_control_OPERA <- ntc_frangieh_control_OPERA[!grepl(pat_rm, rownames(ntc_frangieh_control_OPERA), perl=TRUE), , drop=FALSE]


p2<-gg_qqplot_df(ntc_frangieh_control_OPERA,title="OPERA NTC Frangieh control")


```



```{r,include=F}
results_run_calibration_check <- readRDS("/project/xuanyao/jiaming/paper/output/sceptre/frangieh_control/union_crt_hallmark_set_genes_match_snakemake_30qc_factorize_phase/results_run_calibration_check.rds")[,c(1,2,7)]
sceptre <- results_run_calibration_check[!is.na(results_run_calibration_check$p_value), ]
sceptre$p_value[sceptre$p_value >=1] <- 1 - 1e-6

sceptre_wide <- sceptre %>%
  pivot_wider(names_from = response_id, values_from = p_value) %>%
  as.data.frame()

# Set the row names to be the grna_id and remove the grna_id column
rownames(sceptre_wide) <- sceptre_wide$grna_id
sceptre_wide <- sceptre_wide[, -1]

sceptre_wide <- sceptre_wide[, colnames(sceptre_wide)%in%rownames(gene_module_index_matrix), drop = FALSE]

module_results <- list()

# Iterate over each row of t_gene_p_no_approximation to compute ACAT p-values for each module
for (i in seq_len(nrow(sceptre_wide))) {
  grna_id <- rownames(sceptre_wide)[i]
  p_values_row <- sceptre_wide[i,] 
  
  # Initialize an empty list to store ACAT p-values and module names for each module
  module_acat_pvals <- c()
  module_names <- c()
 matched_targets <-grna_id
  cis_genes <- discovery_pairs_cis[discovery_pairs_cis$grna_target == matched_targets, ]$response_id  
  
if (grna_id %in% rownames(perturbation_gene_qc)) {
  remove_genes <- perturbation_gene_qc[grna_id, 2][[1]]
 
} else {
  remove_genes <- character(0)  # 
}
  
  
  # Iterate over each module in gene_modules to combine p-values using ACAT
  for (j in 1:ncol(gene_module_index_matrix)) {
    
    
    module_name <- colnames(gene_module_index_matrix)[j]  # Get the module name

    module_genes <- rownames(gene_module_index_matrix)[gene_module_index_matrix[, j] == 1]
module_genes <- intersect(module_genes, colnames(p_values_row))
    
      module_genes <- setdiff(module_genes, cis_genes) 
      haha<-length(module_genes)
            module_genes <- setdiff(module_genes, remove_genes)

      
    
    # Only apply ACAT if there are genes to combine and they are not NA
    if (length(module_genes) > 0) {
      module_p_values <- as.numeric(p_values_row[module_genes])
      module_p_values <- module_p_values[!is.na(module_p_values)]  # Remove NAs
      
      if (length(module_p_values) > 0) {
        acat_pval <- ACAT(module_p_values)
        
#      acat_pval <- combination.test(
#   module_p_values[!is.na(module_p_values)],
#   weights = NA,
#   method = "Cauchy",
#   tail.idx = 1
# )


        module_acat_pvals <- c(module_acat_pvals, acat_pval)
        module_names <- c(module_names, module_name)
      } else {
        # print("dont have valid p values for that module")
        module_acat_pvals <- c(module_acat_pvals, NA)
        module_names <- c(module_names, module_name)
      }
    } else {
      
      # print("dont have valid gene in that module")
      module_acat_pvals <- c(module_acat_pvals, NA)
      module_names <- c(module_names, module_name)
    }
  }
  
  # Store results as a data frame in the list
  module_results[[i]] <- data.frame(
    grna_id = grna_id,
    module = module_names,
    acat_pval = module_acat_pvals
  )
}


# Combine all results into a single data frame
module_results_sceptre <- do.call(rbind, module_results)
module_results_sceptre <- module_results_sceptre[!is.na(module_results_sceptre$acat_pval), ]

library(dplyr)
library(tidyr)

mat_df <- module_results_sceptre %>%
  dplyr::select(grna_id, module, acat_pval) %>%
  tidyr::pivot_wider(
    names_from  = module,
    values_from = acat_pval
  )
mat_df_2<- as.matrix(mat_df[,-1])
rownames(mat_df_2) <- mat_df$grna_id

ntc_r <- which(grepl("NO_SITE",mat_df$grna_id))


ntc_frangieh_control_SCEPTRE <- mat_df_2[ntc_r,]

ntc_frangieh_control_SCEPTRE <- ntc_frangieh_control_SCEPTRE[!grepl(pat_rm, rownames(ntc_frangieh_control_SCEPTRE), perl=TRUE), , drop=FALSE]

p3<-gg_qqplot_df(ntc_frangieh_control_SCEPTRE,title="sceptre+ACAT: geneset NTC frangieh (control)")
gg_qqplot(as.vector(as.matrix(ntc_frangieh_control_SCEPTRE)))
```



```{r,fig.width=10,fig.height=21}
grid.arrange(p1,p2,p3,ncol=1)
```

## Papalexi


```{r,echo=F,include=F}
grna_cutoff<-0

gene_module_index_matrix <-  readRDS("/project/xuanyao/jiaming/Getting_started/data/papalexi/gene_module_index_matrix.rds")

grna_target_data_frame_omit <- as.data.frame(readRDS("/project/xuanyao/jiaming/Getting_started/data/papalexi/grna_target_dataframe.rds"))

discovery_pairs_cis <-readRDS("/project/xuanyao/jiaming/Getting_started/data/papalexi/discovery_pairs_cis.rds")

grna_assignment_matrix <- readRDS("/project/xuanyao/jiaming/Getting_started/data/papalexi/union_grna_assignment_matrix.rds")

max_stat <- as.data.frame(readRDS("/project/xuanyao/jiaming/paper/output/snake/papalexi_Hallmark/union_3000_t_ACAT_resample/p_values/geneset_p_value.rds"))

max_stat <- max_stat[rowSums(grna_assignment_matrix)>grna_cutoff, ]


sceptre <-  readRDS("/project/xuanyao/jiaming/paper/output/sceptre/papalexi/union_crt_match_snakemake/results_run_discovery_analysis.rds")[,c(1,3,7)]
                 
perturbation_gene_qc <- character(0)
 


sceptre <- sceptre[!is.na(sceptre$p_value), ]
sceptre$p_value[sceptre$p_value >=1] <- 1 - 1e-6

sceptre_wide <- sceptre %>%
  pivot_wider(names_from = response_id, values_from = p_value) %>%
  as.data.frame()

# Set the row names to be the grna_id and remove the grna_id column
rownames(sceptre_wide) <- sceptre_wide$grna_target
sceptre_wide <- sceptre_wide[, -1]

sceptre_wide <- sceptre_wide[rownames(max_stat)[rownames(max_stat) %in% rownames(sceptre_wide)], ]

sceptre_wide <- sceptre_wide[, colnames(sceptre_wide)%in%rownames(gene_module_index_matrix), drop = FALSE]

library(eulerr)




module_results <- list()

# Iterate over each row of t_gene_p_no_approximation to compute ACAT p-values for each module
for (i in seq_len(nrow(sceptre_wide))) {
  grna_id <- rownames(sceptre_wide)[i]
  p_values_row <- sceptre_wide[i,] 
  
  # Initialize an empty list to store ACAT p-values and module names for each module
  module_acat_pvals <- c()
  module_names <- c()
 matched_targets <-grna_id
  cis_genes <- discovery_pairs_cis[discovery_pairs_cis$grna_target == matched_targets, ]$response_id  
  
if (grna_id %in% rownames(perturbation_gene_qc)) {
  remove_genes <- perturbation_gene_qc[grna_id, 2][[1]]
 
} else {
  remove_genes <- character(0)  # 
}
  
  
  # Iterate over each module in gene_modules to combine p-values using ACAT
  for (j in 1:ncol(gene_module_index_matrix)) {
    
    
    module_name <- colnames(gene_module_index_matrix)[j]  # Get the module name

    module_genes <- rownames(gene_module_index_matrix)[gene_module_index_matrix[, j] == 1]
module_genes <- intersect(module_genes, colnames(p_values_row))
    
      module_genes <- setdiff(module_genes, cis_genes) 
      haha<-length(module_genes)
            module_genes <- setdiff(module_genes, remove_genes)
            print(haha)

      
    
    # Only apply ACAT if there are genes to combine and they are not NA
    if (length(module_genes) > 0) {
      module_p_values <- as.numeric(p_values_row[module_genes])
      module_p_values <- module_p_values[!is.na(module_p_values)]  # Remove NAs
      
      if (length(module_p_values) > 0) {
        acat_pval <- ACAT(module_p_values)
        
#      acat_pval <- combination.test(
#   module_p_values[!is.na(module_p_values)],
#   weights = NA,
#   method = "Cauchy",
#   tail.idx = 1
# )


        module_acat_pvals <- c(module_acat_pvals, acat_pval)
        module_names <- c(module_names, module_name)
      } else {
        # print("dont have valid p values for that module")
        module_acat_pvals <- c(module_acat_pvals, NA)
        module_names <- c(module_names, module_name)
      }
    } else {
      
      # print("dont have valid gene in that module")
      module_acat_pvals <- c(module_acat_pvals, NA)
      module_names <- c(module_names, module_name)
    }
  }
  
  # Store results as a data frame in the list
  module_results[[i]] <- data.frame(
    grna_id = grna_id,
    module = module_names,
    acat_pval = module_acat_pvals
  )
}


# Combine all results into a single data frame
module_results_sceptre <- do.call(rbind, module_results)
module_results_sceptre <- module_results_sceptre[!is.na(module_results_sceptre$acat_pval), ]

library(tibble)

max_stat_long <- max_stat%>%
  rownames_to_column(var = "grna_id") %>% # Add row names as a column
  pivot_longer(cols = -grna_id, names_to = "module", values_to = "p_value")

# Remove rows with NA p-values
max_stat_long <- max_stat_long %>% filter(!is.na(p_value))






max_stat_long <- max_stat_long %>%
  semi_join(module_results_sceptre, by = c("grna_id", "module"))

module_results_sceptre$bh_pval <- p.adjust(module_results_sceptre$acat_pval, method = "BH")
module_results_sceptre$identifier <- paste(module_results_sceptre$grna_id, module_results_sceptre$module, sep = "_")



max_stat_long$bh_pval <-  p.adjust(max_stat_long$p_value, method = "BH")
max_stat_long$identifier <- paste(max_stat_long$grna_id, max_stat_long$module, sep = "_")



```





```{r,include=F}
library(eulerr)

significant_levels <- c(0.05)
# significant_levels <- 0.01
plot_list <- list()

# Loop through significant levels and display plots

significant_ids_gene_wise_p_papalexi <- na.omit(
  module_results_sceptre$identifier[module_results_sceptre$bh_pval < significant_levels]
)


significant_ids_max_stat_papalexi <- na.omit(
  max_stat_long$identifier[max_stat_long$bh_pval < significant_levels]
)


  venn_data <- list(
    `Sceptre+ACAT` = significant_ids_gene_wise_p_papalexi,
    `OPERA`=significant_ids_max_stat_papalexi
  )
  
  # Generate the Euler diagram and display it
euler_plot <- plot(
  euler(venn_data),
  fills = list(fill = c("red", "yellow", "orange","blue"), alpha = 0.5),
  edges = TRUE,
  quantities = TRUE,
  labels = list(font = 2, cex = 1.2, pos = 3), # 调整字体和位置
  main = paste("Significant level", significant_levels)
)
p1 <- euler_plot
```


```{r,include=F}
max_stat <- as.data.frame(readRDS("/project/xuanyao/jiaming/paper/output/snake/papalexi_Hallmark/union_3000_t_ACAT_resample/p_values/geneset_p_value.rds"))

ntc_rows <- which(grepl("NTg",rownames(max_stat)))
ntc_papalexi_OPERA <- max_stat[ntc_rows,]
p2<-gg_qqplot_df(ntc_papalexi_OPERA,title="OPERA NTC")
```


```{r,include=F}
results_run_calibration_check <- readRDS("/project/xuanyao/jiaming/paper/output/sceptre/papalexi/union_crt_match_snakemake/results_run_calibration_check.rds")[,c(1,2,7)]
sceptre <- results_run_calibration_check[!is.na(results_run_calibration_check$p_value), ]
sceptre$p_value[sceptre$p_value >=1] <- 1 - 1e-6

sceptre_wide <- sceptre %>%
  pivot_wider(names_from = response_id, values_from = p_value) %>%
  as.data.frame()

# Set the row names to be the grna_id and remove the grna_id column
rownames(sceptre_wide) <- sceptre_wide$grna_id
sceptre_wide <- sceptre_wide[, -1]

sceptre_wide <- sceptre_wide[, colnames(sceptre_wide)%in%rownames(gene_module_index_matrix), drop = FALSE]

module_results <- list()

# Iterate over each row of t_gene_p_no_approximation to compute ACAT p-values for each module
for (i in seq_len(nrow(sceptre_wide))) {
  grna_id <- rownames(sceptre_wide)[i]
  p_values_row <- sceptre_wide[i,] 
  
  # Initialize an empty list to store ACAT p-values and module names for each module
  module_acat_pvals <- c()
  module_names <- c()
 matched_targets <-grna_id
  cis_genes <- discovery_pairs_cis[discovery_pairs_cis$grna_target == matched_targets, ]$response_id  
  
if (grna_id %in% rownames(perturbation_gene_qc)) {
  remove_genes <- perturbation_gene_qc[grna_id, 2][[1]]
 
} else {
  remove_genes <- character(0)  # 
}
  
  
  # Iterate over each module in gene_modules to combine p-values using ACAT
  for (j in 1:ncol(gene_module_index_matrix)) {
    
    
    module_name <- colnames(gene_module_index_matrix)[j]  # Get the module name

    module_genes <- rownames(gene_module_index_matrix)[gene_module_index_matrix[, j] == 1]
module_genes <- intersect(module_genes, colnames(p_values_row))
    
      module_genes <- setdiff(module_genes, cis_genes) 
      haha<-length(module_genes)
            module_genes <- setdiff(module_genes, remove_genes)

      
    
    # Only apply ACAT if there are genes to combine and they are not NA
    if (length(module_genes) > 0) {
      module_p_values <- as.numeric(p_values_row[module_genes])
      module_p_values <- module_p_values[!is.na(module_p_values)]  # Remove NAs
      
      if (length(module_p_values) > 0) {
        acat_pval <- ACAT(module_p_values)
        
#      acat_pval <- combination.test(
#   module_p_values[!is.na(module_p_values)],
#   weights = NA,
#   method = "Cauchy",
#   tail.idx = 1
# )


        module_acat_pvals <- c(module_acat_pvals, acat_pval)
        module_names <- c(module_names, module_name)
      } else {
        # print("dont have valid p values for that module")
        module_acat_pvals <- c(module_acat_pvals, NA)
        module_names <- c(module_names, module_name)
      }
    } else {
      
      # print("dont have valid gene in that module")
      module_acat_pvals <- c(module_acat_pvals, NA)
      module_names <- c(module_names, module_name)
    }
  }
  
  # Store results as a data frame in the list
  module_results[[i]] <- data.frame(
    grna_id = grna_id,
    module = module_names,
    acat_pval = module_acat_pvals
  )
}


# Combine all results into a single data frame
module_results_sceptre <- do.call(rbind, module_results)
module_results_sceptre <- module_results_sceptre[!is.na(module_results_sceptre$acat_pval), ]

library(dplyr)
library(tidyr)

mat_df <- module_results_sceptre %>%
  dplyr::select(grna_id, module, acat_pval) %>%
  tidyr::pivot_wider(
    names_from  = module,
    values_from = acat_pval
  )
rownames(mat_df) <- mat_df$grna_id
mat_df<-mat_df[,-1]

ntc_papalexi_SCEPTRE <- mat_df
p3<-gg_qqplot_df(ntc_papalexi_SCEPTRE,title="sceptre+ACAT: geneset NTC")
```



```{r,fig.width=7,fig.height=21}
grid.arrange(p1,p2,p3,ncol=1)
```

# Figure 1B — Stacked barplot: OPERA vs. SCEPTRE+ACAT

Each bar represents one dataset, split into two stacked segments: discoveries shared between OPERA and SCEPTRE+ACAT (darker shade) and method-exclusive discoveries (lighter shade). Across all five datasets tested here (Morris, Xie, Frangieh IFN-γ, Frangieh control, Papalexi), OPERA identifies substantially more significant trans SNP–gene set associations than SCEPTRE+ACAT at FDR < 0.05, with only a small fraction of OPERA's discoveries missed by SCEPTRE+ACAT.



```{r}
make_bar_df <- function(sig_sceptre, sig_opera, dataset_name){
  sig_sceptre <- unique(na.omit(sig_sceptre))
  sig_opera   <- unique(na.omit(sig_opera))

  shared  <- intersect(sig_sceptre, sig_opera)
  sc_only <- setdiff(sig_sceptre, sig_opera)
  op_only <- setdiff(sig_opera, sig_sceptre)

  tibble(
    dataset = dataset_name,
    method  = c("Sceptre+ACAT","Sceptre+ACAT","OPERA","OPERA"),
    segment = c("Shared","Sceptre+ACAT only","Shared","OPERA only"),
    count   = c(length(shared), length(sc_only), length(shared), length(op_only))
  )
}

# 你把每个dataset的两个向量放这里
sig_list <- list(
  Morris = list(
    sceptre = significant_ids_gene_wise_p_morris,
    opera     = significant_ids_max_stat_morris
  ),
  Xie = list(
    sceptre = significant_ids_gene_wise_p_xie,
    opera     = significant_ids_max_stat_xie
  ),
  Frangieh_ifn_gamma = list(
    sceptre = significant_ids_gene_wise_p_frangieh_ifn,
    opera     = significant_ids_max_stat_frangieh_ifn
  ),
    Frangieh_control = list(
    sceptre = significant_ids_gene_wise_p_frangieh_control,
    opera     = significant_ids_max_stat_frangieh_control
  ),
      Papalexi = list(
    sceptre = significant_ids_gene_wise_p_papalexi,
    opera     = significant_ids_max_stat_papalexi
  )
)

plot_df <- bind_rows(lapply(names(sig_list), function(ds){
  make_bar_df(sig_list[[ds]]$sceptre, sig_list[[ds]]$opera, ds)
})) %>%
  mutate(
    method  = factor(method, levels = c("Sceptre+ACAT","OPERA")),
    segment = factor(segment, levels = c("Shared","Sceptre+ACAT only","OPERA only"))
  )
plot_df <- plot_df %>%
  mutate(
    segment = factor(segment,
                     levels = c("Shared", "Sceptre+ACAT only", "OPERA only"))
  )

desired_order <- c("Xie", "Morris", "Papalexi", "Frangieh_control", "Frangieh_ifn_gamma")

plot_df <- plot_df %>%
  mutate(dataset = factor(dataset, levels = desired_order))
```

```{r,fig.width=10,fig.height=5}
plot_df <- plot_df %>%
  mutate(
    segment = factor(segment, levels = c("OPERA only", "Sceptre+ACAT only", "Shared"))
  )

ggplot(plot_df, aes(x = method, y = count, fill = segment)) +
  geom_col(position = "stack", width = 0.6) +
  facet_wrap(~ dataset, nrow = 1) +
  scale_fill_manual(
    values = c(
      "Shared"            = "#4E79A7",
      "Sceptre+ACAT only" = "#F28E2B",
      "OPERA only"        = "#59A14F"
    ),
    breaks = c("Shared", "Sceptre+ACAT only", "OPERA only")  # 图例顺序不变
  ) +
  labs(x = NULL, y = "Significant discoveries", fill = NULL) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x   = element_text(angle = 35, hjust = 1),
    legend.position = "right",
    strip.text    = element_text(size = 9),
    panel.spacing = unit(0.3, "lines")
  )
```



```{r,fig.width=10,fig.height=5}
# ── Dot + line plot（connected paired dot plot）────────────
# 数据结构：每个 dataset，两个方法的 total discoveries，用线连接

dot_df <- plot_df %>%
  group_by(dataset, method) %>%
  summarise(total = sum(count[segment %in% c("Shared", "Sceptre+ACAT only", "OPERA only") &
                                !(segment == "Shared" & method == "OPERA")]),
            # 每个 bar 的真实 total = Shared + method-only
            # Sceptre+ACAT bar: Shared + Sceptre+ACAT only
            # OPERA bar:        Shared + OPERA only
            total = sum(count),   # sum 后会重复 Shared，需要重算
            .groups = "drop")

# 更清晰地重算每个方法的真实 total
dot_df <- plot_df %>%
  filter(
    (method == "Sceptre+ACAT" & segment %in% c("Shared", "Sceptre+ACAT only")) |
    (method == "OPERA"        & segment %in% c("Shared", "OPERA only"))
  ) %>%
  group_by(dataset, method) %>%
  summarise(total = sum(count), .groups = "drop")

ggplot(dot_df, aes(x = method, y = total, group = dataset)) +
  # 连线
  geom_line(aes(color = dataset), linewidth = 0.8, alpha = 0.7) +
  # 点
  geom_point(aes(color = dataset), size = 3.5) +
  # 在点旁边标注数值
  geom_text(aes(label = total, color = dataset),
            hjust = -0.3, size = 3, show.legend = FALSE) +
  scale_color_brewer(palette = "Set2", name = "Dataset") +
  scale_x_discrete(expand = expansion(mult = c(0.15, 0.25))) +  # 右侧留空给标签
  labs(
    x = NULL,
    y = "Significant discoveries"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

# Figure 1C — NTC QQ plot: calibration comparison

Non-targeting control (NTC) guide–gene set p-values are displayed as QQ plots for OPERA (blue) and SCEPTRE+ACAT (grey) across datasets. Points lying along the diagonal indicate well-calibrated p-values under the null; deviation above the diagonal indicates inflated false positive rates. The 90% confidence envelope is computed from the Beta distribution. OPERA is well-calibrated across most datasets; the slight inflation seen for NTC-1 in the Xie dataset is attributable to a technical issue with that specific guide rather than the method.



```{r}
library(dplyr)
library(tibble)
library(ggplot2)

qq_curve_from_mat <- function(pmat, set_label, dataset, cap = 6){
  p <- as.numeric(as.matrix(pmat))
  p <- p[!is.na(p)]
  p[p <= 0] <- 1e-250
  p[p >= 1] <- 1 - 1e-6
  p <- sort(p)
  n <- length(p)

  tibble(
    expected = -log10(ppoints(n)),
    observed = pmin(-log10(p), cap),
    Set = set_label,      # 用于颜色（OPERA vs SCEPTRE+ACAT）
    Dataset = dataset     # 用于分组（每个dataset一条线）
  )
}

qq_envelope_maxn <- function(df_long, ci = 0.9, cap = 6){
  max_n <- df_long %>%
    group_by(Set, Dataset) %>%
    summarise(n = n(), .groups="drop") %>%
    summarise(max_n = max(n)) %>%
    pull(max_n)

  exp_env <- -log10(ppoints(max_n))
  lower <- pmin(-log10(qbeta((1-ci)/2, seq_len(max_n), rev(seq_len(max_n)))), cap)
  upper <- pmin(-log10(qbeta((1+ci)/2, seq_len(max_n), rev(seq_len(max_n)))), cap)

  tibble(expected = exp_env, ymin = lower, ymax = upper)
}

plot_qq_overlay_multi_dataset <- function(mats_gray, mats_blue,
                                         label_gray = "SCEPTRE + ACAT",
                                         label_blue = "OPERA",
                                         title = "NTC QQ plot (datasets x methods)",
                                         ci = 0.9, cap = 6,
                                         add_method_median = TRUE){

  ds <- intersect(names(mats_gray), names(mats_blue))
  if (length(ds) == 0) stop("mats_gray and mats_blue have no shared names (dataset names).")

  df_gray <- bind_rows(lapply(ds, function(d){
    qq_curve_from_mat(mats_gray[[d]], label_gray, d, cap = cap)
  }))
  df_blue <- bind_rows(lapply(ds, function(d){
    qq_curve_from_mat(mats_blue[[d]], label_blue, d, cap = cap)
  }))
  df <- bind_rows(df_gray, df_blue)

  env_df <- qq_envelope_maxn(df, ci = ci, cap = cap)

  p <- ggplot() +
    geom_ribbon(data = env_df,
                aes(x = expected, ymin = ymin, ymax = ymax),
                fill = "grey85", alpha = 0.35) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.6) +
    geom_line(
      data = df,
      aes(x = expected, y = observed,
          group = interaction(Set, Dataset),
          color = Set,
          linetype = Dataset),          # <- 关键：dataset 用线型区分
      linewidth = 0.9, alpha = 0.35
    ) +
    scale_color_manual(values = c(
      "SCEPTRE + ACAT" = "grey20",
      "OPERA" = "red"
    )) +
    labs(
      title = title,
      x = expression(Expected~~-log[10](P)),
      y = expression(Observed~~-log[10](P)),
      color = NULL,
      linetype = NULL                  # <- dataset 图例标题（不想要标题就设 NULL）
    ) +
    ylim(0, cap) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      legend.position = "top",
      legend.key.width = unit(1.2, "cm")
    ) +
    guides(
      color = guide_legend(override.aes = list(alpha = 1, linewidth = 2)),
      linetype = guide_legend(override.aes = list(alpha = 1, linewidth = 1.2))
    )

  # median 粗线（如果你开启）
  if (add_method_median) {
    grid_x <- seq(0, cap, length.out = 400)

    df_med <- df %>%
      group_by(Set, Dataset) %>%
      summarise(obs_interp = list(approx(expected, observed, xout = grid_x, rule = 2)$y),
                .groups = "drop") %>%
      tidyr::unnest_wider(obs_interp)

    df_med_long <- df_med %>%
      pivot_longer(cols = -c(Set, Dataset), names_to = "idx", values_to = "obs") %>%
      mutate(expected = grid_x[as.integer(gsub("\\D", "", idx))]) %>%
      group_by(Set, expected) %>%
      summarise(observed = median(obs, na.rm = TRUE), .groups = "drop")

    p <- p +
      geom_line(data = df_med_long,
                aes(x = expected, y = observed, color = Set),
                linewidth = 1.6, alpha = 0.95)
  }

  p
}

```


```{r}
mats_gray <- list(
  Frangieh_control = ntc_frangieh_control_SCEPTRE,
  Frangieh_ifn_gamma = ntc_frangieh_ifn_SCEPTRE,
  Papalexi = ntc_papalexi_SCEPTRE,
  Morris = ntc_morris_SCEPTRE
)

mats_blue <- list(
  Frangieh_control = ntc_frangieh_control_OPERA,
  Frangieh_ifn_gamma = ntc_frangieh_ifn_OPERA,
  Papalexi = ntc_papalexi_OPERA,
  Morris = ntc_morris_OPERA
)
```


```{r,fig.width=10}
p <- plot_qq_overlay_multi_dataset(
  mats_gray, mats_blue,
  label_gray = "SCEPTRE + ACAT",
  label_blue = "OPERA",
  title = "NTC QQ plot (4 datasets, 2 methods)",
  ci = 0.9, cap = 10,
  add_method_median = F
)

print(p)

```

# Figure 2 — Topic modeling: data-driven gene set discovery

## Figure 2B — Jaccard similarity heatmap of topic gene sets

Topic models are fitted to perturb-responsive cells using `fastTopics`. Gene sets are derived from grade-of-membership differential expression (GoM-DE): a gene is assigned to a topic if it has LFSR < 0.01, |z| ≥ 2, and |log2FC| ≥ 0.5. This heatmap shows pairwise Jaccard overlaps between the K = 50 topic gene sets (using the MCH Replogle dataset), with hierarchical clustering to reveal groups of functionally related programs. Blocks of high Jaccard similarity correspond to topics capturing the same co-regulated biological process.

```{r,fig.width=15, fig.height=15}
library(pheatmap)
gene_set_K100 <- readRDS("/project/xuanyao/jiaming/Getting_started/data/repologle/topic_geneset/flames+200kb/de_50_0.01_2_0.5.rds")
clean_vec <- function(x) unique(as.character(x))

L1 <- lapply(gene_set_K100   , clean_vec)
L2 <- lapply(gene_set_K100  , clean_vec)

# pairwise Jaccard matrix
jaccard_mat <- outer(
  seq_along(L1),
  seq_along(L2),
  Vectorize(function(i, j) {
    a <- L1[[i]]
    b <- L2[[j]]
    if (length(a) == 0 && length(b) == 0) return(NA_real_)
    length(intersect(a, b)) / max(length(a),length(b))
  })
)

rownames(jaccard_mat) <- names(L1)
colnames(jaccard_mat) <- names(L2)

mat <- as.matrix(jaccard_mat)
storage.mode(mat) <- "numeric"

pheatmap(
  mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  treeheight_row = 0,
  treeheight_col = 0,
  fontsize_row = 7,
  fontsize_col = 9,
  angle_col = 45,
  filename = "/project/xuanyao/jiaming/paper/jaccard_heatmap.pdf",
  width = 12, height = 12
)

pheatmap(
  mat,
  cluster_rows  = TRUE,
  cluster_cols  = TRUE,
  treeheight_row = 0,
  treeheight_col = 0,
  fontsize_row  = 7,
  fontsize_col  = 9,
  angle_col     = 45,
  width         = 12,
  height        = 12
)
```

```{r,fig.width=15, fig.height=15}
pheatmap(
  mat,
  cluster_rows  = TRUE,
  cluster_cols  = TRUE,
  treeheight_row = 0,
  treeheight_col = 0,
  fontsize_row  = 7,
  fontsize_col  = 9,
  angle_col     = 45,
  width         = 12,
  height        = 12
)
```



# Computational efficiency benchmarks

OPERA is designed to scale to large Perturb-seq datasets. This section reports wall-clock time (seconds/minutes per chunk) and peak RAM usage (MiB/GiB) by parsing Snakemake job log files (`*.out`) for each dataset. Each "chunk" corresponds to one Snakemake job processing a subset of perturbation–gene set pairs. The summaries show that OPERA's per-chunk runtime and memory footprint remain tractable even for the largest datasets (e.g., the Replogle dataset with >1,000 perturbations).

## OPERA benchmarks

### Morris dataset

```{r}
library(dplyr)
library(stringr)

# 设置目录路径
out_dir <- "/project/xuanyao/jiaming/OPERA/snakemake/logs_morris/out"  # ← 改成你的实际路径

out_files <- list.files(out_dir, pattern = "\\.out$", full.names = TRUE)

parse_out_file <- function(filepath) {
  lines <- readLines(filepath, warn = FALSE)
  
  header_idx <- grep("Elapsed_Time_sec", lines)
  if (length(header_idx) == 0) return(NULL)
  
  # 每个文件可能有多个步骤（多行数据）
  chunks <- lapply(header_idx, function(i) {
    header <- strsplit(trimws(lines[i]), "\\s+")[[1]]
    data   <- strsplit(trimws(lines[i + 1]), "\\s+")[[1]]
    if (length(data) == length(header) + 1) data <- data[-1]  # 去行号
    df <- as.data.frame(t(data), stringsAsFactors = FALSE)
    colnames(df) <- header
    df
  }) %>% bind_rows()
  
  chunks$file  <- basename(filepath)
  chunks$chunk <- str_extract(basename(filepath), "chunk_\\d+")  # 提取 chunk_N
  chunks
}

# 读取所有文件
all_steps <- lapply(out_files, parse_out_file) %>%
  bind_rows() %>%
  mutate(
    Elapsed_Time_sec   = as.numeric(Elapsed_Time_sec),
    Total_RAM_Used_MiB = as.numeric(Total_RAM_Used_MiB),
    Peak_RAM_Used_MiB  = as.numeric(Peak_RAM_Used_MiB),
    # 按数字排序用
    chunk_num = as.integer(str_extract(chunk, "\\d+"))
  )

# 以 chunk 为单位汇总
chunk_summary <- all_steps %>%
  group_by(chunk, chunk_num) %>%
  summarise(
    n_steps            = n(),
    total_Elapsed_sec  = sum(Elapsed_Time_sec,   na.rm = TRUE),
    total_Elapsed_min  = round(sum(Elapsed_Time_sec, na.rm = TRUE) / 60, 2),
    max_Peak_RAM_MiB   = max(Peak_RAM_Used_MiB,  na.rm = TRUE),
    max_Peak_RAM_GiB   = round(max(Peak_RAM_Used_MiB, na.rm = TRUE) / 1024, 2),
    .groups = "drop"
  ) %>%
  arrange(chunk_num)

print(chunk_summary)

# 所有 chunk 的总体统计
overall <- tibble(
  total_chunks       = nrow(chunk_summary),
  total_Elapsed_sec  = sum(chunk_summary$total_Elapsed_sec),
  total_Elapsed_min  = round(sum(chunk_summary$total_Elapsed_sec) / 60, 2),
  max_Peak_RAM_MiB   = max(chunk_summary$max_Peak_RAM_MiB),   # 所有 chunk 中最高峰值
  max_Peak_RAM_GiB   = round(max(chunk_summary$max_Peak_RAM_MiB) / 1024, 2)
)

cat("\n── Overall ──\n")
print(overall)
```

```{r}
median(chunk_summary$total_Elapsed_min)
```



### Morris dataset (SCEPTRE comparison)


